# Making the data source

We will create a master dataset of watercourse segments of 100 m (`watercourse_100mseg`), in order to derive sampling frames of lotic types (3260 in particular).

It is a processed data source, derived from the raw data source `watercourse_segments` (version `watercourse_segments_20180601`).
The segments will have a 100 m length, except at the end of the linestrings of `watercourse_segments`.

Further we derive a second data source `watercourse_100msegpoints` which represent the endpoints of the 100 m segments.

```{r}
local_root <- find_root(has_file("generate_watercourse_100mseg.Rproj"))
datapath <- file.path(local_root, "data")
if (!dir.exists(datapath)) dir.create(datapath)
finalpath_segments <- find_root_file("n2khab_data/20_processed/watercourse_100mseg", 
                                     criterion = has_dir("n2khab_data"))
if (!dir.exists(finalpath_segments)) dir.create(finalpath_segments)
finalpath_segmentpoints <- find_root_file("n2khab_data/20_processed/watercourse_100msegpoints", 
                                          criterion = has_dir("n2khab_data"))
if (!dir.exists(finalpath_segmentpoints)) dir.create(finalpath_segmentpoints)
```

```{r}
grass_location <- file.path(datapath, "grass_watercourses")
if (!dir.exists(grass_location)) dir.create(file.path(grass_location, "PERMANENT"),
                                            recursive = TRUE)
grass_exists <- file.exists(file.path(grass_location, "PERMANENT/PROJ_INFO"))
if (.Platform$OS.type == "unix") {
  Sys.setenv(GRASS_PYTHON = system("which python3", intern = TRUE))
}
```


## Geoprocessing steps in GRASS

Read the `watercourse_segments` data source (this step still needs reproducibility).

```{r eval=!grass_exists, paged.print=FALSE}
watercourse_segments <- read_sf(file.path(datapath, "watercourse_segments"))
```

Initialize GRASS projectfolder if non-existant, or link to the already existing one:

```{r}
initGRASS(gisBase = link2GI::findGRASS()[1, 1],
          home = tempdir(),
          gisDbase = datapath, location = "grass_watercourses", 
          mapset = "PERMANENT", override = TRUE)
```


```{r eval=!grass_exists}
# setting CRS and extent of the GRASS location, based on 'watercourse_segments'
execGRASS("g.proj", flags = c("c", "quiet"), 
          georef = file.path(datapath, "watercourse_segments"))
  # note: setting the extent should only affect raster computations (so we don't need it)
b_box <- st_bbox(watercourse_segments)
execGRASS("g.region", flags = c("quiet"), 
          n = as.character(b_box["ymax"]), s = as.character(b_box["ymin"]), 
          e = as.character(b_box["xmax"]), w = as.character(b_box["xmin"]), 
          res = "1")
```


```{r}
use_sf()
```

Add `watercourse_segments` data source to GRASS database:

```{r eval=!grass_exists}
execGRASS("v.in.ogr",
          input = file.path(datapath, "watercourse_segments"),
          output = "watercourse_segments")
```

### Generate `watercourse_100mseg` in GRASS

To do this, we use the `v.split` command.

<!-- g.manual v.split -->

```{r eval=!grass_exists}
execGRASS("v.split",
          flags = c("f", "overwrite"),
          input = "watercourse_segments",
          output = "watercourse_100mseg_pre1",
          length = 100)
```

```{r}
system("v.category input=watercourse_100mseg_pre1 option=report")
```

As seen from the above input, in accordance with the `v.split` manual, the categories and the attribute table of the original vector map are simply copied over.
This means that the individual 100 m segments have no unique category, i.e. cannot have their own rows (with ID) in another attribute table.
To get that, as explained in the documentation, we define a second layer^[To do that, the vector map `watercourse_100mseg_pre1` is copied, including layers & the link to the same attribute table.] with a unique category per feature (100 m segment) and create a new attribute table that links to the new layer (an attribute table always has exactly 1 row for each category and uses 1 layer to take categories from).
Next, we copy the category of layer 1 as a mere attribute into a second column in the new attribute table (which is linked to layer 2), in order to define the link between the grouped category `cat_group` (categories of `watercourses_segements`) and the unique category `cat`.

In order to execute the below four GRASS commands (not the outcommented ones), it may be necessary to break them down into separate `execGRASS()` statements.^[
This depends on the availability of the GRASS executables in the shell PATH, which should have been set by `initGRASS()`.
]
The `system()` command directly sends the below commands to the shell as one script for execution by GRASS.

```{r eval=!grass_exists}
system('v.category input=watercourse_100mseg_pre1 option=add layer=2 output=watercourse_100mseg
       # v.category input=watercourse_100mseg option=report
       # v.db.connect -p map=watercourse_100mseg
       v.db.addtable watercourse_100mseg layer=2
       # v.db.connect -p map=watercourse_100mseg
       # v.info -c watercourse_100mseg layer=2
       # v.info -c watercourse_100mseg layer=1
       v.db.addcolumn map=watercourse_100mseg layer=2 column="cat_group int"
       v.to.db map=watercourse_100mseg layer=2 option=query query_layer=1 query_column=cat columns=cat_group --overwrite
       # v.db.select map=watercourse_100mseg layer=2 | head
       ')
```

<!-- Output from the v.to.db command: -->

<!-- WARNING: Values in column <cat_group> will be overwritten -->
<!-- Reading features... -->
<!--  100% -->
<!-- Querying database... -->
<!--  100% -->
<!-- Updating database... -->
<!--  100% -->
<!-- 263871 categories read from vector map (layer 2) -->
<!-- 263871 records selected from table (layer 1) -->
<!-- 263871 categories read from vector map exist in selection from table -->
<!-- 263871 records updated/inserted (layer 2) -->



<!-- v.db.connect -p map=watercourse_100mseg -->
<!-- Vector map <watercourse_100mseg> is connected by: -->
<!-- layer <1/watercourse_segments> table <watercourse_100mseg> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <cat> -->
<!-- layer <2/watercourse_100mseg_2> table <watercourse_100mseg_2> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <cat> -->

### Generate `watercourse_100msegpoints` in GRASS

To do this, we use the `v.to.points` command.

<!-- g.manual v.to.points -->

```{r eval=!grass_exists}
execGRASS("v.to.points",
          flags = c("overwrite"),
          input = "watercourse_100mseg",
          layer = "2",
          output = "watercourse_100msegpoints",
          use = "end")
```

<!-- v.info watercourse_100msegpoints layer=1 -->
<!--  +----------------------------------------------------------------------------+ -->
<!--  | Name:            watercourse_100msegpoints                                 | -->
<!--  | Mapset:          PERMANENT                                                 | -->
<!--  | Location:        grass_watercourses                                        | -->
<!--  | Database:        (...)                                                     | -->
<!--  | Title:                                                                     | -->
<!--  | Map scale:       1:1                                                       | -->
<!--  | Name of creator: floris                                                    | -->
<!--  | Organization:                                                              | -->
<!--  | Source date:     Fri Nov 27 16:14:43 2020                                  | -->
<!--  | Timestamp (first layer): none                                              | -->
<!--  |----------------------------------------------------------------------------| -->
<!--  | Map format:      native                                                    | -->
<!--  |----------------------------------------------------------------------------| -->
<!--  |   Type of map: vector (level: 2)                                           | -->
<!--  |                                                                            | -->
<!--  |   Number of points:       263871          Number of centroids:  0          | -->
<!--  |   Number of lines:        0               Number of boundaries: 0          | -->
<!--  |   Number of areas:        0               Number of islands:    0          | -->
<!--  |                                                                            | -->
<!--  |   Map is 3D:              No                                               | -->
<!--  |   Number of dblinks:      2                                                | -->
<!--  |                                                                            | -->
<!--  |   Projection: Belge 1972 / Belgian Lambert 72                              | -->
<!--  |                                                                            | -->
<!--  |               N:       244044.9074    S:   153062.38093558                 | -->
<!--  |               E:       258444.2968    W:    23006.85298789                 | -->
<!--  |                                                                            | -->
<!--  |   Digitization threshold: 0                                                | -->
<!--  |   Comment:                                                                 | -->
<!--  |                                                                            | -->
<!--  +----------------------------------------------------------------------------+ -->

<!-- v.category input=watercourse_100msegpoints option=report -->
<!-- Layer/table: 1/watercourse_100msegpoints_1 -->
<!-- type       count        min        max -->
<!-- point     263871          1     263871 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       263871          1     263871 -->
<!-- Layer/table: 2/watercourse_100msegpoints_2 -->
<!-- type       count        min        max -->
<!-- point     263871          1     263871 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       263871          1     263871 -->


# Checks on the data source

