# Making the data source

We will create a master dataset of watercourse segments of 100 m (`watercourse_100mseg`), in order to derive sampling frames of lotic types (3260 in particular).

It is a processed data source, derived from the raw data source `watercourses` (version `watercourses_20200807`).
The segments will have a 100 m length, except at the end of the linestrings of `watercourses`.

Further we will also derive a second layer which represent the _endpoints_ of the 100 m segments.

```{r}
local_root <- find_root(has_file("generate_watercourse_100mseg.Rproj"))
datapath <- file.path(local_root, "data")
n2khab_datapath <- find_root_file("n2khab_data",
                                  criterion = has_dir("n2khab_data"))
if (!dir.exists(datapath)) dir.create(datapath)
finalpath <- find_root_file("n2khab_data/20_processed/watercourse_100mseg", 
                            criterion = has_dir("n2khab_data"))
if (!dir.exists(finalpath)) dir.create(finalpath)
```

```{r}
grass_location <- file.path(datapath, "grass_watercourses")
if (!dir.exists(grass_location)) dir.create(file.path(grass_location, "PERMANENT"),
                                            recursive = TRUE)
grassdbase_exists <- file.exists(file.path(grass_location, "PERMANENT/PROJ_INFO"))
if (.Platform$OS.type == "unix") {
  Sys.setenv(GRASS_PYTHON = system("which python3", intern = TRUE))
}
```


## Geoprocessing steps in GRASS

Read the `watercourses` data source.

```{r paged.print=FALSE}
watercourses <- read_sf(file.path(n2khab_datapath, "10_raw/watercourses"))
```

Initialize GRASS projectfolder if non-existant, or link to the already existing one:

```{r}
if (interactive()) {
gisbase_grass <- 
  if (.Platform$OS.type == "windows") link2GI::paramGRASSw()$gisbase_GRASS[1] else {
    link2GI::paramGRASSx()$gisbase_GRASS[1]
  }
}
initGRASS(gisBase = if (interactive()) gisbase_grass else params$gisbase_grass,
          home = tempdir(),
          gisDbase = datapath, location = "grass_watercourses", 
          mapset = "PERMANENT", override = TRUE)
```

Cross-platform function to pass full GRASS command syntax as a string from R:

```{r}
execshell <- function(commandstring, intern = FALSE) {
  if (.Platform$OS.type == "windows") {
    res <- shell(commandstring, intern = TRUE)
  } else {
    res <- system(commandstring, intern = TRUE)
  }
  if (!intern) cat(res, sep = "\n") else return(res)
}
```


```{r eval=!grassdbase_exists}
# setting CRS and extent of the GRASS location, based on 'watercourses'
execGRASS("g.proj", flags = c("c", "quiet"), 
          epsg = 31370)
```


```{r}
use_sf()
```

Add `watercourses` data source to GRASS database:

```{r eval=!grassdbase_exists}
execGRASS("v.in.ogr",
          "o",
          input = file.path(n2khab_datapath, "10_raw/watercourses"),
          output = "watercourses")
```

### Generate `watercourse_100mseg` in GRASS

To do this, we first use the `v.edit` command to flip the direction of the linestrings, because the segments are to be generated from mouth to source, leaving segments of <100 m only at the source side.
After that, we will use the `v.split` command.
It will take each linestring of the raw `watercourses` data source and split it into segments of 100 m length.
This generally results in one segment (for each original linestring) that is shorter than 100 m!

<!-- g.manual v.split -->

```{r eval=!grassdbase_exists}
execshell("g.copy vector=watercourses,watercourses_original")
execshell("v.edit map=watercourses tool=flip where='1=1'")
execshell("v.split -f --overwrite input=watercourses output=watercourse_100mseg_pre1 length=100")
```

```{r}
execshell("v.category input=watercourse_100mseg_pre1 option=report")
```

As seen from the above output, in accordance with the `v.split` manual, the categories and the attribute table of the original vector map are simply copied over.
This means that the individual 100 m segments have no unique category, i.e. they don't have their own rows (with ID) in another attribute table.
To get that, as explained in the documentation, we define a second layer^[To do that, the vector map `watercourse_100mseg_pre1` is copied, including layers & the link to the same attribute table.] with a unique category per feature (100 m segment) and create a new attribute table that links to the new layer (an attribute table always has exactly 1 row for each category and uses 1 layer to take categories from).
Next, we copy the category of layer 1 as a mere attribute into a second column in the new attribute table (which is linked to layer 2), in order to define the link between the grouped category `cat_group` (categories of `watercourses`) and the unique category `cat`.

The `execshell()` command directly sends the below commands to the shell as one script for execution by GRASS.

```{r eval=!grassdbase_exists}
execshell('v.category input=watercourse_100mseg_pre1 option=add layer=2 output=watercourse_100mseg
       # v.category input=watercourse_100mseg option=report
       # v.db.connect -p map=watercourse_100mseg
       v.db.addtable watercourse_100mseg layer=2 key="rank" columns="vhag_code int"
       # v.db.connect -p map=watercourse_100mseg
       # v.info -c watercourse_100mseg layer=2
       # v.info -c watercourse_100mseg layer=1
       v.to.db map=watercourse_100mseg layer=2 option=query query_layer=1 query_column=VHAG columns=vhag_code --overwrite
       # v.db.select map=watercourse_100mseg layer=2 | head
       ')
```

<!-- Output from the v.to.db command: -->

<!-- WARNING: Values in column <vhag_code> will be overwritten -->
<!-- Reading features... -->
<!--  100% -->
<!-- Querying database... -->
<!--  100% -->
<!-- Updating database... -->
<!--  100% -->
<!-- 271676 categories read from vector map (layer 2) -->
<!-- 271676 records selected from table (layer 1) -->
<!-- 271676 categories read from vector map exist in selection from table -->
<!-- 271676 records updated/inserted (layer 2) -->



<!-- v.db.connect -p map=watercourse_100mseg -->
<!-- Vector map <watercourse_100mseg> is connected by: -->
<!-- layer <1/watercourses> table <watercourse_100mseg> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <cat> -->
<!-- layer <2/watercourse_100mseg_2> table <watercourse_100mseg_2> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <rank> -->

### Generate `watercourse_100msegpoints` in GRASS

To do this, we use the `v.to.points` command.
We want each point at the downstream side of its corresponding segment.
We create one point per 100 m segment.
As we flipped the direction of the watercourse lines, we want the _startpoint_ of each segment.

<!-- g.manual v.to.points -->

```{r eval=!grassdbase_exists}
execGRASS("v.to.points",
          flags = c("overwrite"),
          input = "watercourse_100mseg",
          layer = "2",
          output = "watercourse_100msegpoints",
          use = "start")
```

<!--  100% -->
<!-- Building topology for vector map <watercourse_100msegpoints@PERMANENT>... -->
<!-- Registering primitives... -->
<!--     270000 -->
<!-- v.to.points complete. 271676 points written to output vector map. -->


<!-- v.info watercourse_100msegpoints layer=1 -->
 <!-- +----------------------------------------------------------------------------+ -->
 <!-- | Name:            watercourse_100msegpoints                                 | -->
 <!-- | Mapset:          PERMANENT                                                 | -->
 <!-- | Location:        grass_watercourses                                        | -->
 <!-- | Database:        /media/floris/DATA/PROJECTS/09685_NatuurlijkMilieu/160 Be | -->
 <!-- | Title:                                                                     | -->
 <!-- | Map scale:       1:1                                                       | -->
 <!-- | Name of creator: floris                                                    | -->
 <!-- | Organization:                                                              | -->
 <!-- | Source date:     Wed Dec 16 16:54:35 2020                                  | -->
 <!-- | Timestamp (first layer): none                                              | -->
 <!-- |----------------------------------------------------------------------------| -->
 <!-- | Map format:      native                                                    | -->
 <!-- |----------------------------------------------------------------------------| -->
 <!-- |   Type of map: vector (level: 2)                                           | -->
 <!-- |                                                                            | -->
 <!-- |   Number of points:       271676          Number of centroids:  0          | -->
 <!-- |   Number of lines:        0               Number of boundaries: 0          | -->
 <!-- |   Number of areas:        0               Number of islands:    0          | -->
 <!-- |                                                                            | -->
 <!-- |   Map is 3D:              No                                               | -->
 <!-- |   Number of dblinks:      2                                                | -->
 <!-- |                                                                            | -->
 <!-- |   Projection: Belge 1972 / Belgian Lambert 72                              | -->
 <!-- |                                                                            | -->
 <!-- |               N:       243962.6791    S:   153064.35460326                 | -->
 <!-- |               E:   258447.81310536    W:    22979.76203745                 | -->
 <!-- |                                                                            | -->
 <!-- |   Digitization threshold: 0                                                | -->
 <!-- |   Comment:                                                                 | -->
 <!-- |                                                                            | -->
 <!-- +----------------------------------------------------------------------------+ -->


<!-- v.category input=watercourse_100msegpoints option=report -->
<!-- Layer/table: 1/watercourse_100msegpoints_1 -->
<!-- type       count        min        max -->
<!-- point     271676          1     271676 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       271676          1     271676 -->
<!-- Layer/table: 2/watercourse_100msegpoints_2 -->
<!-- type       count        min        max -->
<!-- point     271676          1     271676 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       271676          1     271676 -->

<!-- v.info -c watercourse_100msegpoints layer=1 -->
<!-- Displaying column types/names for database connection of layer <1>: -->
<!-- INTEGER|rank -->
<!-- INTEGER|vhag_code -->

## Exporting resulting objects from GRASS

We already took care of standardized column names within GRASS; so no further changes are needed when exporting the data from GRASS.
We do it this way in order to minimize the number of times the data need to be written forth and back to a geopackage, which takes extra time (> 250e3 features) and file organization.

```{r eval=params$grass_reexport}
execGRASS("v.out.ogr",
          flags = "overwrite",
          input = "watercourse_100mseg",
          layer = "2",
          output = file.path(finalpath, "watercourse_100mseg.gpkg"),
          output_layer = "watercourse_100mseg_lines")
```

```{r eval=params$grass_reexport}
# This export takes a very long time (> 60 min).
# There must be some bug in the export procedure 
# (after all, the data are simpler than previous one).
execGRASS("v.out.ogr",
          "a",
          input = "watercourse_100msegpoints",
          layer = "1",
          output = file.path(finalpath, "watercourse_100mseg.gpkg"),
          output_layer = "watercourse_100mseg_points")
```


# Attribute variables of the processed data source: explanation

Both layers in the GPKG file have exactly the same attribute variables:

- `rank`: a unique integer increment (1-2-3-4-5-...).
It is to be used in a _relative_ way: it ranks the 100 m segments / points along each original linestring in the raw data source.
- `vhag_code`: the unique VHAG code of the original linestring in the raw data source.
It is common to all segments / points in the processed data source that belong to the same original watercourse.

To obtain other attributes, one can make a join on the VHAG code column of the raw data source.


# Checks on the data source

```{r}
filepath <- file.path(finalpath, "watercourse_100mseg.gpkg")
```

Checksums:

```{r}
file(filepath) %>% 
  openssl::md5() %>% 
  str_c(collapse = '') %>% 
  `names<-`("md5sum")
file(filepath) %>% 
  openssl::sha256() %>% 
  str_c(collapse = '') %>% 
  `names<-`("sha256sum")
```


```{r paged.print=FALSE, warning=FALSE}
(seg <- read_sf(filepath, 
                layer = "watercourse_100mseg_lines"))
```

```{r paged.print=FALSE, warning=FALSE}
(pts <- read_sf(filepath, 
                layer = "watercourse_100mseg_points"))
```

```{r}
all.equal(seg$rank, pts$rank)
```

```{r}
all.equal(seg$vhag_code, pts$vhag_code)
```

```{r}
all.equal(unique(seg$vhag_code), unique(watercourses$VHAG))
```

Cartographic display of a row selection:

```{r}
seg %>% 
  filter(rank < 2e4) %>% 
  ggplot() +
  geom_sf(colour = "grey50") +
  geom_sf(data = pts %>% filter(rank < 2e4), colour = "red", size = 0.5) +
  lims(x = c(198e3, 204e3), y = c(183e3, 188e3)) +
  coord_sf(datum = 31370) +
  theme_bw()+
  theme(panel.grid = element_blank())
```


