# Making the data source

We will create a master dataset of watercourse segments of 100 m (`watercourse_100mseg`), in order to derive sampling frames of lotic types (3260 in particular).

It is a processed data source, derived from the raw data source `watercourse_segments` (version `watercourse_segments_20180601`).
The segments will have a 100 m length, except at the end of the linestrings of `watercourse_segments`.

Further we derive a second data source `watercourse_100msegpoints` which represent the endpoints of the 100 m segments.

```{r}
local_root <- find_root(has_file("generate_watercourse_100mseg.Rproj"))
datapath <- file.path(local_root, "data")
if (!dir.exists(datapath)) dir.create(datapath)
finalpath <- find_root_file("n2khab_data/20_processed/watercourse_100mseg", 
                            criterion = has_dir("n2khab_data"))
if (!dir.exists(finalpath)) dir.create(finalpath)
```

```{r}
grass_location <- file.path(datapath, "grass_watercourses")
if (!dir.exists(grass_location)) dir.create(file.path(grass_location, "PERMANENT"),
                                            recursive = TRUE)
grassdbase_exists <- file.exists(file.path(grass_location, "PERMANENT/PROJ_INFO"))
if (.Platform$OS.type == "unix") {
  Sys.setenv(GRASS_PYTHON = system("which python3", intern = TRUE))
}
```


## Geoprocessing steps in GRASS

Read the `watercourse_segments` data source (this step still needs reproducibility).

```{r paged.print=FALSE}
watercourse_segments <- read_sf(file.path(datapath, "watercourse_segments"))
```

Initialize GRASS projectfolder if non-existant, or link to the already existing one:

```{r}
initGRASS(gisBase = link2GI::findGRASS()[1, 1],
          home = tempdir(),
          gisDbase = datapath, location = "grass_watercourses", 
          mapset = "PERMANENT", override = TRUE)
```


```{r eval=!grassdbase_exists}
# setting CRS and extent of the GRASS location, based on 'watercourse_segments'
execGRASS("g.proj", flags = c("c", "quiet"), 
          georef = file.path(datapath, "watercourse_segments"))
  # note: setting the extent should only affect raster computations (so we don't need it)
b_box <- st_bbox(watercourse_segments)
execGRASS("g.region", flags = c("quiet"), 
          n = as.character(b_box["ymax"]), s = as.character(b_box["ymin"]), 
          e = as.character(b_box["xmax"]), w = as.character(b_box["xmin"]), 
          res = "1")
```


```{r}
use_sf()
```

Add `watercourse_segments` data source to GRASS database:

```{r eval=!grassdbase_exists}
execGRASS("v.in.ogr",
          input = file.path(datapath, "watercourse_segments"),
          output = "watercourse_segments")
```

### Generate `watercourse_100mseg` in GRASS

To do this, we use the `v.split` command.

<!-- g.manual v.split -->

```{r eval=!grassdbase_exists}
execGRASS("v.split",
          flags = c("f", "overwrite"),
          input = "watercourse_segments",
          output = "watercourse_100mseg_pre1",
          length = 100)
```

```{r}
system("v.category input=watercourse_100mseg_pre1 option=report")
```

As seen from the above output, in accordance with the `v.split` manual, the categories and the attribute table of the original vector map are simply copied over.
This means that the individual 100 m segments have no unique category, i.e. they don't have their own rows (with ID) in another attribute table.
To get that, as explained in the documentation, we define a second layer^[To do that, the vector map `watercourse_100mseg_pre1` is copied, including layers & the link to the same attribute table.] with a unique category per feature (100 m segment) and create a new attribute table that links to the new layer (an attribute table always has exactly 1 row for each category and uses 1 layer to take categories from).
Next, we copy the category of layer 1 as a mere attribute into a second column in the new attribute table (which is linked to layer 2), in order to define the link between the grouped category `cat_group` (categories of `watercourses_segements`) and the unique category `cat`.

In order to execute the below three GRASS commands (not the outcommented ones), it may be necessary to break them down into separate `execGRASS()` statements.^[
This depends on the availability of the GRASS executables in the shell PATH, which should have been set by `initGRASS()`.
]
The `system()` command directly sends the below commands to the shell as one script for execution by GRASS.

```{r eval=!grassdbase_exists}
system('v.category input=watercourse_100mseg_pre1 option=add layer=2 output=watercourse_100mseg
       # v.category input=watercourse_100mseg option=report
       # v.db.connect -p map=watercourse_100mseg
       v.db.addtable watercourse_100mseg layer=2 key="rank" columns="vhas_code int"
       # v.db.connect -p map=watercourse_100mseg
       # v.info -c watercourse_100mseg layer=2
       # v.info -c watercourse_100mseg layer=1
       v.to.db map=watercourse_100mseg layer=2 option=query query_layer=1 query_column=VHAS columns=vhas_code --overwrite
       # v.db.select map=watercourse_100mseg layer=2 | head
       ')
```

<!-- Output from the v.to.db command: -->

<!-- WARNING: Values in column <vhas_code> will be overwritten -->
<!-- Reading features... -->
<!--  100% -->
<!-- Querying database... -->
<!--  100% -->
<!-- Updating database... -->
<!--  100% -->
<!-- 263871 categories read from vector map (layer 2) -->
<!-- 263871 records selected from table (layer 1) -->
<!-- 263871 categories read from vector map exist in selection from table -->
<!-- 263871 records updated/inserted (layer 2) -->



<!-- v.db.connect -p map=watercourse_100mseg -->
<!-- Vector map <watercourse_100mseg> is connected by: -->
<!-- layer <1/watercourse_segments> table <watercourse_100mseg> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <cat> -->
<!-- layer <2/watercourse_100mseg_2> table <watercourse_100mseg_2> in database <(...)/n2khab-preprocessing/src/generate_watercourse_100mseg/data/grass_watercourses/PERMANENT/sqlite/sqlite.db> through driver <sqlite> with key <rank> -->

### Generate `watercourse_100msegpoints` in GRASS

To do this, we use the `v.to.points` command.

<!-- g.manual v.to.points -->

```{r eval=!grassdbase_exists}
execGRASS("v.to.points",
          flags = c("overwrite"),
          input = "watercourse_100mseg",
          layer = "2",
          output = "watercourse_100msegpoints",
          use = "end")
```

<!-- v.info watercourse_100msegpoints layer=1 -->
<!--  +----------------------------------------------------------------------------+ -->
<!--  | Name:            watercourse_100msegpoints                                 | -->
<!--  | Mapset:          PERMANENT                                                 | -->
<!--  | Location:        grass_watercourses                                        | -->
<!--  | Database:        (...)                                                     | -->
<!--  | Title:                                                                     | -->
<!--  | Map scale:       1:1                                                       | -->
<!--  | Name of creator: floris                                                    | -->
<!--  | Organization:                                                              | -->
<!--  | Source date:     Fri Nov 27 16:14:43 2020                                  | -->
<!--  | Timestamp (first layer): none                                              | -->
<!--  |----------------------------------------------------------------------------| -->
<!--  | Map format:      native                                                    | -->
<!--  |----------------------------------------------------------------------------| -->
<!--  |   Type of map: vector (level: 2)                                           | -->
<!--  |                                                                            | -->
<!--  |   Number of points:       263871          Number of centroids:  0          | -->
<!--  |   Number of lines:        0               Number of boundaries: 0          | -->
<!--  |   Number of areas:        0               Number of islands:    0          | -->
<!--  |                                                                            | -->
<!--  |   Map is 3D:              No                                               | -->
<!--  |   Number of dblinks:      2                                                | -->
<!--  |                                                                            | -->
<!--  |   Projection: Belge 1972 / Belgian Lambert 72                              | -->
<!--  |                                                                            | -->
<!--  |               N:       244044.9074    S:   153062.38093558                 | -->
<!--  |               E:       258444.2968    W:    23006.85298789                 | -->
<!--  |                                                                            | -->
<!--  |   Digitization threshold: 0                                                | -->
<!--  |   Comment:                                                                 | -->
<!--  |                                                                            | -->
<!--  +----------------------------------------------------------------------------+ -->

<!-- v.category input=watercourse_100msegpoints option=report -->
<!-- Layer/table: 1/watercourse_100msegpoints_1 -->
<!-- type       count        min        max -->
<!-- point     263871          1     263871 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       263871          1     263871 -->
<!-- Layer/table: 2/watercourse_100msegpoints_2 -->
<!-- type       count        min        max -->
<!-- point     263871          1     263871 -->
<!-- line           0          0          0 -->
<!-- boundary       0          0          0 -->
<!-- centroid       0          0          0 -->
<!-- area           0          0          0 -->
<!-- face           0          0          0 -->
<!-- kernel         0          0          0 -->
<!-- all       263871          1     263871 -->



## Exporting resulting objects from GRASS

We already took care of standardized column names within GRASS; so no further changes are needed when exporting the data from GRASS.
We do it this way in order to minimize the number of times the data need to be written forth and back to a geopackage, which takes extra time (> 250e3 features) and file organization.

```{r eval=params$grass_reexport}
execGRASS("v.out.ogr",
          flags = "overwrite",
          input = "watercourse_100mseg",
          layer = "2",
          output = file.path(finalpath, "watercourse_100mseg.gpkg"),
          output_layer = "watercourse_100mseg_lines")
```

```{r eval=params$grass_reexport}
# This export takes a very long time (> 60 min).
# There must be some bug in the export procedure 
# (after all, the data are simpler than previous one).
execGRASS("v.out.ogr",
          input = "watercourse_100msegpoints",
          layer = "1",
          output = file.path(finalpath, "watercourse_100mseg.gpkg"),
          output_layer = "watercourse_100mseg_points")
```



# Checks on the data source

```{r}
filepath <- file.path(finalpath, "watercourse_100mseg.gpkg")
```

Checksums:

```{r}
file(filepath) %>% 
  openssl::md5() %>% 
  str_c(collapse = '') %>% 
  `names<-`("md5sum")
file(filepath) %>% 
  openssl::sha256() %>% 
  str_c(collapse = '') %>% 
  `names<-`("sha256sum")
```

We best set the crs explicitly: while the WKT from the gpkg (exported from GRASS using GDAL/OGR) is correct (parameters identical to the ones of `watercourse_segments`), it lost a few names and would otherwise be displayed as 'unknown'.

```{r paged.print=FALSE, warning=FALSE}
(seg <- read_sf(filepath, 
                layer = "watercourse_100mseg_lines",
                crs = 31370)
)
```

```{r paged.print=FALSE, warning=FALSE}
(pts <- read_sf(filepath, 
                layer = "watercourse_100mseg_points",
                crs = 31370))
```

```{r}
all.equal(seg$rank, pts$rank)
```

```{r}
all.equal(seg$vhas_code, pts$vhas_code)
```

```{r}
all.equal(unique(seg$vhas_code), unique(watercourse_segments$VHAS))
```

Cartographic display of a row selection:

```{r}
seg %>% 
  filter(rank < 2e4) %>% 
  ggplot() +
  geom_sf(colour = "grey50") +
  geom_sf(data = pts %>% filter(rank < 2e4), colour = "red", size = 0.5) +
  lims(x = c(198e3, 204e3), y = c(183e3, 188e3)) +
  coord_sf(datum = 31370) +
  theme_bw()+
  theme(panel.grid = element_blank())
```


