# Generate random number based on uniform distribution for each GRTS master cell 

```{r}
grts_master <- read_GRTSmh()

raster_runif <- grts_master

set.seed(3841280)
values(raster_runif) <- runif(n = ncell(raster_runif))

raster_runif@data@names <- "draw_runif"

# writeRaster(raster_runif, 
#             file.path(fileman_up("n2khab_data"), "20_processed/raster_runif/raster_runif.tif"),
#             format = "GTiff")
```

For every grid-cell in the 32m x 32m grts-master we draw a random number between 0 and 1 based on the uniform distribution.

# Application: sample selection for terrestrial types

To draw an equal probability sample for terrestrial types, following approach can be used:

+ select all grts-cells within polygons that contain (partially) a certain type;
+ perform a Bernoulli trial for all grts-cell with the probability of success equal to proportion of the type within the polygon (the phab value);
+ rank the successful grts-cells according to the grts-ranking and select a number of grts-cells according to the desired sample size.

We can use raster_runif to perform the Bernoulli trial: a grts-cell is successful when the corresponding raster_runif value is lower than the phab value. This way we only have to draw one random number per grts-cell, even when the grts-cell contains different possible types each having a different phab-value. 


```{r}
habmap_terr <- read_habitatmap_terr()

habitatmap_terr_polygons <- habmap_terr$habitatmap_terr_polygons 

habitatmap_terr_types <- habmap_terr$habitatmap_terr_types

habpolygon_example <- habitatmap_terr_polygons %>%
  filter(polygon_id == "118623_v2014")

grts_habpolygon_example <- grts_master %>%
  crop(habpolygon_example) %>%
  mask(habpolygon_example) %>%
  rasterToPoints(spatial = TRUE) %>%
  st_as_sf() %>%
  rename(grts_ranking = GRTSmaster_habitats) 
  
st_crs(grts_habpolygon_example) <- 31370

sampling_units_example <- grts_habpolygon_example %>%
  mutate(draw_runif = raster_runif[as(., "Spatial")])%>%
  st_join(dplyr::select(habpolygon_example, polygon_id)) %>%
  left_join(habitatmap_terr_types, by = "polygon_id") %>%
  mutate(draw_bernoulli = draw_runif < phab/100)

```

As example we can take a look at a polygon from the habitatmap containing following types:

```{r}
sampling_units_example %>%
  st_drop_geometry() %>%
  distinct(polygon_id, type, phab) %>%
  arrange(type) %>%
  kable() %>%
  kable_styling() %>%
  collapse_rows(columns = 1)
  
```

In the table bellow we illustrate how the do a Bernoulli trial per grts-cell and per type, based raster_runif value.

```{r}
sampling_units_example %>%
  st_drop_geometry() %>%
  dplyr::select(type, phab, grts_ranking, draw_runif, draw_bernoulli) %>%
  arrange(type) %>%
  kable() %>%
  kable_styling() %>%
  collapse_rows(columns = 1)
```

The result for type 6510_hu is shown bellow. The green circles correspond the centroids of the successful grts-cells. The labels show the relative rank of the successful grts-cells. 

```{r, out.width = '100%'}
sampling_units_example_show <- sampling_units_example %>%
  filter(type == "6510_hu") %>% 
  group_by(draw_bernoulli) %>%
  mutate(volgorde = rank(grts_ranking)) %>%
  ungroup() %>%
  mutate(show_color = ifelse(draw_bernoulli, "green", "red"),
         text = ifelse(draw_bernoulli, str_c("rank: ",volgorde), "")) %>%
  st_transform(4326) 


leaflet(sampling_units_example_show) %>%
  addTiles() %>%
  addCircleMarkers(color = ~show_color,
                   label = ~text, 
                   labelOptions = labelOptions(noHide = TRUE)) %>%
  addPolygons(data = st_transform(habpolygon_example, 4326))
    
  
    
```

`

