# Generate random number based on uniform distribution for each GRTS master cell 

+ We start from the `GRTSmaster_habitats` data source

+ To be sure we will use the correct version of the data source (GRTSmaster_habitats_v2), we will derive the md5 file hashes and compare it to the file hashes in the [data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853).

```{r}
path <- fileman_up("n2khab_data")
file <- "10_raw/GRTSmaster_habitats"
mypath <- file.path(path, file)

hashes <-
    tibble(filepath = str_c(mypath, "/",
        list.files(path = mypath,
            recursive = TRUE)
      )) %>%
    mutate(filename = str_match(filepath, "(.+\\/)*(.+)")[,3],
           md5 = map(filepath, function(x) {
                           file(x) %>% md5 %>% str_c(collapse = '')
                         }) %>% as.character,
           md5_ref = c("20de76e1abfbafd6edcc00e1a9cf87a0"),
           match = md5 == md5_ref) %>%
    select(filename,
           md5,
           md5_ref,
           match)

kable(hashes) %>%
  kable_styling()

if (!all.equal(hashes$md5, hashes$md5_ref)) {
    stop(cat("The source map is NOT up to date ! Please check  the datasource. "))
} else {
  cat("We have a match!")
}
```

```{r}
grts_master <- read_GRTSmh()

```

+ Next we increase the extent of the raster to anticipate on issue [#45](https://github.com/inbo/n2khab-preprocessing/issues/45).

+ We generate a random number between 0 and 1 according to the uniform distribution

+ Finally we mask the raster using a shapefile of Flanders with a buffer of 1024 m: all raster cells outside the buffer get a NA value. We include the Brussels region (by using `fill_holes` function of the `smoothr` package) as the grts_master also covers this region. 

```{r}

flanders <- read_admin_areas(dsn = "flanders")

flanders_and_brussels <- fill_holes(flanders, threshold = units::set_units(1000000, km^2))

flanders_and_brussels_buffer <- flanders_and_brussels %>%
  st_buffer(32 * 32)
```


```{r}
raster_runif <- grts_master

extent_raster <- extent(grts_master)
extent_raster@xmin <- extent_raster@xmin - 32 * 64
extent_raster@ymin <- extent_raster@ymin - 32 * 64
extent_raster@xmax <- extent_raster@xmax + 32 * 64
extent_raster@ymax <- extent_raster@ymax + 32 * 64

raster_runif_buffer <- setExtent(raster_runif, 
                                 ext = extent_raster, 
                                 keepres = TRUE, 
                                 snap = TRUE
                                 )

set.seed(3841280)
values(raster_runif_buffer) <- runif(n = ncell(raster_runif_buffer))

raster_runif_buffer_mask <- raster_runif_buffer %>%
  crop(flanders_and_brussels_buffer) %>%
  mask(flanders_and_brussels_buffer)

#values(raster_runif_buffer_mask) <- values(raster_runif_buffer_mask)/10000

raster_runif_buffer_mask@data@names <- "raster_runif"

```



# Check the result

```{r}
habmap_terr <- read_habitatmap_terr()

habitatmap_terr_polygons <- habmap_terr$habitatmap_terr_polygons 

polygon_show <- habitatmap_terr_polygons %>%
  filter(polygon_id == "580464_v2018")

polygon_show_buffer <- st_buffer(polygon_show, dist = 200)

grts_polygon_example <- grts_master %>%
  crop(polygon_show_buffer) %>%
  #mask(polygon_show_buffer) %>%
  rasterToPoints(spatial = TRUE) %>%
  st_as_sf() %>%
  rename(grts_ranking = GRTSmaster_habitats) 

st_crs(grts_polygon_example) <- 31370

runif_polygon_example <- raster_runif_buffer_mask %>%
  crop(polygon_show_buffer) %>%
  #mask(polygon_show_buffer) %>%
  rasterToPoints(spatial = TRUE) %>%
  st_as_sf()

st_crs(runif_polygon_example) <- 31370


```

The map below shows:

+ a polygon of the habitatmap at the border of Flanders
+ the centroids of the grts_master grid-cells in the surrounding of the habitatmap polygon (larger black circles)
+ the centroids of the raster_runif grid-cells in the surrounding of the habitatmap polygon (smaller yellow circles)
+ the raster_runif values (labels of the yellow circles)

The result shows that:

+ the grid-cells of the grts_master overlap with those of raster_runif
+ the extent of raster_runif exceeds the border of Flanders
+ the values of raster_runif are between 0 and 1

```{r, out.width= "100%"}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = st_transform(polygon_show, crs = 4326)) %>%
  addCircleMarkers(data = st_transform(grts_polygon_example, crs = 4326), radius = 6, color = "black") %>%
  addCircleMarkers(data = st_transform(runif_polygon_example, crs = 4326), radius = 2, color = "yellow", label = ~raster_runif)
```



# Application: sample selection for terrestrial types

To draw an equal probability sample for terrestrial types, following approach can be used:

+ select all grts-cells within polygons that contain (partially) a certain type;
+ perform a Bernoulli trial for all grts-cell with the probability of success equal to proportion of the type within the polygon (the phab value);
+ rank the successful grts-cells according to the grts-ranking and select a number of grts-cells according to the desired sample size.

We can use raster_runif to perform the Bernoulli trial: a grts-cell is successful when the corresponding raster_runif value is lower than the phab value. This way we only have to draw one random number per grts-cell, even when the grts-cell contains different possible types each having a different phab-value. 


```{r}

habitatmap_terr_types <- habmap_terr$habitatmap_terr_types

habpolygon_example <- habitatmap_terr_polygons %>%
  filter(polygon_id == "118623_v2014")

grts_habpolygon_example <- grts_master %>%
  crop(habpolygon_example) %>%
  mask(habpolygon_example) %>%
  rasterToPoints(spatial = TRUE) %>%
  st_as_sf() %>%
  rename(grts_ranking = GRTSmaster_habitats) 
  
st_crs(grts_habpolygon_example) <- 31370

sampling_units_example <- grts_habpolygon_example %>%
  mutate(draw_runif = raster_runif_buffer_mask[as(., "Spatial")])%>%
  st_join(dplyr::select(habpolygon_example, polygon_id)) %>%
  left_join(habitatmap_terr_types, by = "polygon_id") %>%
  mutate(draw_bernoulli = draw_runif < phab/100)

```

As an example we can take a look at a polygon from the habitatmap containing following types:

```{r}
sampling_units_example %>%
  st_drop_geometry() %>%
  distinct(polygon_id, type, phab) %>%
  arrange(type) %>%
  kable() %>%
  kable_styling() %>%
  collapse_rows(columns = 1)
  
```

In the table bellow we illustrate how the do a Bernoulli trial per grts-cell and per type, based raster_runif value.

```{r}
sampling_units_example %>%
  st_drop_geometry() %>%
  dplyr::select(type, phab, grts_ranking, draw_runif, draw_bernoulli) %>%
  arrange(type) %>%
  kable() %>%
  kable_styling() %>%
  collapse_rows(columns = 1)
```

The result for type 6510_hu is shown bellow. The green circles correspond the centroids of the successful grts-cells. The labels show the relative rank of the successful grts-cells. 

```{r, out.width = '100%'}

sampling_units_example_show <- sampling_units_example %>%
  filter(type == "6510_hu") %>% 
  group_by(draw_bernoulli) %>%
  mutate(volgorde = rank(grts_ranking)) %>%
  ungroup() %>%
  mutate(show_color = ifelse(draw_bernoulli, "green", "red"),
         text = ifelse(draw_bernoulli, str_c("rank: ",volgorde), "not selected")) %>%
  st_transform(4326) 


leaflet(sampling_units_example_show) %>%
  addTiles() %>%
  addCircleMarkers(color = ~show_color,
                   label = ~text, 
                   labelOptions = labelOptions(noHide = TRUE)) %>%
  addPolygons(data = st_transform(habpolygon_example, 4326))
    
  
    
```

# Write result

```{r}
dir.create(file.path(path, "10_raw/raster_runif"), recursive = TRUE)

filepath <-  file.path(fileman_up("n2khab_data"), "10_raw/raster_runif/raster_runif.tif")
```


```{r, eval = params$overwrite_result}
writeRaster(raster_runif_buffer_mask,
            filepath,
            format = "GTiff",
            overwrite = TRUE)
```


# Checks on the data source

Checksums:

```{r}
plyr::each(xxh64sum, md5sum, sha256sum)(filepath)

```

+ Check if values are between 0 and 1

```{r}
(raster_runif_buffer_mask@data@min >= 0) & (raster_runif_buffer_mask@data@max <= 1)
```

+ Check distribution

```{r}
tibble(value_runif = values(raster_runif_buffer_mask)) %>%
  filter(!is.na(value_runif)) %>%
  ggplot(aes(x = value_runif)) +
  geom_density()
  
```

