# Strategy for updating attributes, coordinates and even new locations (rows)

1. load current data source as `sf` object with `read_habitatsprings()`
1. write attributes, with xy-coordinates, to:
  - tsv (tab separated values). This one is only meant for git version control.
  - googlesheet
1. after updating googlesheet, read it back in
1. with the result, overwrite the tsv-file
1. use the new dataframe to replace attributes of `sf` object (depending on the specific change, specific handling could be done)
1. write new `sf` object to geojson format RFC7946

New locations can however also be retrieved from another, separate input layer, and added to the existing `sf` object.

See also the `mapedit` project that allows to create / edit spatial `sf` objects in a graphical interface, but not yet attributes.

# Execution

## Load current version

```{r paged.print = FALSE}
path <- fileman_up("n2khab_data")
file <-  "10_raw/habitatsprings/habitatsprings.geojson"
filepath <- file.path(path, file)
habitatsprings <- read_sf(filepath)
habitatsprings
```

## Write attributes

```{r paged.print = FALSE}
oldopt <- options(digits = 15)
hs_attr <- 
  habitatsprings %>%
  mutate(X = st_coordinates(.)[,1],
         Y = st_coordinates(.)[,2]) %>% 
  st_drop_geometry()
```

```{r}
hs_attr %>% 
  write_tsv("data/hs_attr.tsv")
```

Not automatically executed but to be done manually:

```{r eval=FALSE}
hs_gs <- gs_key("1YCClw_1psllIwSNYZhol38dX4yqA674j95u6lfroXmU")
ws_name <- lubridate::today() %>% as.character
hs_gs %>% 
  gs_ws_new(ws_name)
hs_gs <- gs_key("1YCClw_1psllIwSNYZhol38dX4yqA674j95u6lfroXmU")
hs_gs %>% 
  gs_edit_cells(ws = ws_name,
                input = hs_attr)
```

## Manual steps for updating to data from gsheet-tab '2020-02-07'

### Read from googlesheet and straighten data

Setting tab name:

```{r eval=FALSE}
ws_name <- "2020-02-07"
```

Reading the information:

```{r eval=FALSE}
hs_gs <- gs_key("1YCClw_1psllIwSNYZhol38dX4yqA674j95u6lfroXmU")
hs_attr_new <- 
  hs_gs %>% 
  gs_read(ws = ws_name) %>% 
  arrange(id)
```

Compare columns that should match: are they still the same?

```{r}
hs_attr %>% 
  mutate(X = X %>% `attributes<-`(NULL),
         Y = Y %>% `attributes<-`(NULL)
         ) %>% 
  all.equal(hs_attr_new[,1:10])
# TRUE
```

Inspecting and rearranging column order:

```{r eval=FALSE}
hs_attr_new <- 
  hs_attr_new %>% 
  select(1:4, 
         system_type, 
         habitattype, 
         unit_id, 
         everything())
hs_attr_new %>% 
  glimpse
```


### Overwrite tsv file

```{r eval=FALSE}
hs_attr_new %>% 
   write_tsv("data/hs_attr.tsv")
```

### Update sf object

```{r eval=FALSE, paged.print=FALSE}
habitatsprings_new <- 
  habitatsprings %>% 
  bind_cols(hs_attr_new %>% select(system_type, unit_id)) %>% 
  select(1:4, 
         system_type, 
         habitattype, 
         unit_id, 
         everything())
habitatsprings_new
```

## Recheck attributes of sf object

When extracting attributes from new sf object, are they the same as the prepared attributes?

```{r}
hs_attr_new2 <- 
  habitatsprings_new %>%
  mutate(X = st_coordinates(.)[,1],
         Y = st_coordinates(.)[,2]) %>% 
  st_drop_geometry()
hs_attr_new2 %>% 
  mutate(X = X %>% `attributes<-`(NULL),
         Y = Y %>% `attributes<-`(NULL)
         ) %>% 
  all.equal(hs_attr_new)
# TRUE
```


## Write geojson file (format RFC7946)

Using current working directory to write new file this (overwrite in old folder as a manual step outside R):

```{r}
habitatsprings_new %>% 
  st_write("habitatsprings.geojson",
           layer_options = "RFC7946=YES",
           delete_dsn = TRUE)
```


