# Generate standardized habitatmap

## Data source

The shapefile of the BWK and Natura 2000 habitat map of Flanders can be downloaded [here](https://zenodo.org/record/10167695).

To be sure we will use the correct version of the data source (habitatmap_2023), we will derive the md5 file hashes and compare it to the file hashes in the [data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853).

```{r}

path <- fileman_up("n2khab_data")
file <- "10_raw/habitatmap"

mypath <- file.path(path, file)

hashes <-
    tibble(filepath = str_c(mypath, "/",
        list.files(path = mypath,
            recursive = TRUE)
      )) %>%
    mutate(filename = str_match(filepath, "(.+\\/)*(.+)")[,3],
           md5 = map(filepath, function(x) {
                           file(x) %>% md5 %>% str_c(collapse = '')
                         }) %>% as.character,
           md5_ref = c("8d436686b2ad4bc9355bf1c403b5ba43",
                       "358ff672c2fae48eba5bd09f8a671675",
                      "f881f61a6c07741b58cb618d8bbb0b99",
                      "8c6999e914d7f95948d84d5221f58e6f",
                      "8e9c2c1ef888ce085e027d32f7c73b8c",
                      "7ff6cc0efbd34f4cfef52d73e466df0d"),
           match = md5 == md5_ref) %>%
    select(filename,
           md5,
           md5_ref,
           match)

kable(hashes) %>%
  kable_styling()

if (!all.equal(hashes$md5, hashes$md5_ref)) {
    stop(cat("The source map is NOT up to date ! Please check  the datasource. "))
}
```


```{r read_raw_data}

habitatmap_sf <- st_read(file.path(path, file),
                  layer = "habitatmap", crs = 31370)
```


## Processing of the attribute table

Every polygon in the habitat map can consist of maximum 5 different vegetation types. This information is stored in the columns 'HAB1', 'HAB2',..., 'HAB5' of the attribute table. The estimated fraction of each vegetation type within the polygons is stored in the columns 'PHAB1', 'PHAB2', ..., 'PHAB5'.

We will convert the attribute table to a long format, so that every row contains one vegetation type.


```{r select_polygons}

habmap_sf <- habitatmap_sf %>%
            filter(!(HAB1 == "gh" & PHAB1 == 100)) %>%
            mutate(polygon_id = TAG, # unieke id
                   description_orig = str_c(PHAB1, "% ", HAB1,
                                      ifelse(is.na(HAB2), "", str_c("; ", PHAB2, "% ", HAB2)),
                                      ifelse(is.na(HAB3), "", str_c("; ", PHAB3, "% ", HAB3)),
                                      ifelse(is.na(HAB4), "", str_c("; ", PHAB4, "% ", HAB4)),
                                      ifelse(is.na(HAB5), "", str_c("; ", PHAB5, "% ", HAB5)))) 

```


```{r long_format}

habmap_longPHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    gather(PHAB1, PHAB2, PHAB3, PHAB4, PHAB5, key = "phabn", value = "phab") %>%
    mutate(patch_id = str_sub(phabn, 5, 5)) %>%
    select(polygon_id, patch_id, phab) %>%
    filter(!is.na(phab))

habmap_longHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    select(polygon_id, polygon_area = OPPERVL, starts_with("HAB"), -HABLEGENDE) %>%
    gather(HAB1, HAB2, HAB3, HAB4, HAB5, key = "hab", value = "code") %>%
    mutate(patch_id = substr(hab, 4, 4)) %>%
    filter(!is.na(code)) %>%
    filter(! code %in% c("gh", "x"))

```

Here are the vegetation codes used in this version of the habitatmap:

```{r showListHabitats}
habmap_longHAB %>%
  distinct(code) %>% 
  arrange(code) %>% 
  kable(caption = "List codes in habitat map", label = NA) %>%
  kable_styling() %>% 
  scroll_box(height = "300px")
```

## Correction of some of the codes in the Habitat map

Some polygons in the habitat map contain codes that do not correspond with the standardized list of habitat types. 
We correct these codes to make processing of the habitat map more straightforward.
Table \@ref(tab:codeCorrected) shows the corrected codes and the number of polygons for which the correction is applied.


```{r }

habmap_correction <- read_vc("habmap_correction/habmap_correction")

overview_habmap_correction <- habmap_longHAB %>%
  inner_join(habmap_correction, by = "code") %>%
  group_by(code, code_corrected) %>%
  summarise(n_polygons = n()) %>%
  ungroup()

```


```{r codeCorrected}
overview_habmap_correction %>%
  kable(caption = "Corrected codes in habitat map", label = NA) %>%
  kable_styling()
```

```{r}
if(sum(habmap_correction$code %in% habmap_longHAB$code) > 0){

    habmap_longHAB <- habmap_longHAB %>%
        left_join(habmap_correction, by = "code") %>%
        mutate(code_orig = code,
            code = ifelse(is.na(code_corrected),
                             code,
                             code_corrected)) %>%
        select(-code_corrected)
    
    } else {
      
      habmap_longHAB <- habmap_longHAB %>%
        mutate(code_orig = code)
}
```


## Splitting codes that contain different types

In several cases the code contains 2 or 3 possible vegetation types which are separated with a ','. 
We will split the different possible vegetation types and create one row for each of them. 
An additional variable 'certain' will be FALSE if the orginal habitatmap code consists of 2 or 3 possible vegetation types, and TRUE if only one vegetation type is provided.

An exception to this rule are following codes: `3130_rbbmr`, `3140_rbbmr`, `3150_rbbmr` and `3160_rbbmr` (converted in the previous step into `31x0,rbbms`).
These are standing water bodies that contain both `rbbmr` and the habitat type.
Therefore `certain` will be set to `TRUE` for both types included in `code`.

```{r}

habmap_long <- habmap_longHAB %>%
    left_join(habmap_longPHAB, by = c("polygon_id", "patch_id")) %>%
    mutate(certain = !str_detect(code, ","),
           certain = ifelse(code_orig %in% 
                              c("3130_rbbmr", "3140_rbbmr", "3150_rbbmr", "3160_rbbmr"), 
                            TRUE, 
                            certain)) %>%
    separate(code,
             into = c("type1", "type2", "type3"),
             sep = ",",
             remove = FALSE) %>%
    gather(type1, type2, type3, key = "ntype", value = "type") %>%
    filter(!is.na(type)) %>%
    filter(!(type %in% c("gh", "bos"))) %>%
    select(-patch_id)

```

Sometimes you get two records for the same type within the same habitatmap polygon.
We distinguish two cases: 

+ two certain or two uncertain records of the same type, for example:

```{r}
habmap_long %>%
  filter(polygon_id == "357179_v2020")
```

In this case we will sum the phab-values and create one record for each type.

+ a certain and an uncertain record of the same type, for example:

```{r}

example <- habmap_long %>%
  group_by(polygon_id, type) %>%
  filter(n() > 1) %>%
  filter(any(certain) & any(!certain)) %>%
  ungroup()

polygon_id_example <- (example %>%
  slice_head(n = 1))$polygon_id

habmap_long %>%
  filter(polygon_id == polygon_id_example)
```

We will not aggregate the certain and uncertain record, because we will loose some information used to create `habitatmap_terr`. It is some of the uncertain records (such as `9120, gh`) are eliminated when processing `habitatmap_terr`. 

```{r}
check_double_type <- habmap_long %>%
  group_by(polygon_id, type) %>%
  mutate(n_type = n()) %>%
  ungroup() %>%
  filter(n_type > 1)
  
```


```{r}
habmap_long_aggr <- habmap_long %>%
  # group_by(polygon_id, code) %>%
  #     mutate(n = n()) %>%
  #   ungroup() %>%
    group_by(polygon_id, type, certain) %>%
     summarise(phab = sum(phab),
              code_orig = str_c(code_orig, collapse = "; "),
              certain = all(certain)) %>%
    ungroup() 
```
See below the result of the aggregation for the first example. 

```{r}
habmap_long_aggr %>%
  filter(polygon_id == "357179_v2020")
```

## Select vegetation types that belong to the standard list of habitat and rbb types  

Table \@ref(tab:selectTypes) shows the records with habitat types that do not belong to the standard list of habitat and rbb types.

Good news! 
The table is empty

```{r selectTypes}

types <- read_types() %>%
  select(type, typelevel, main_type, typeclass)

habmap_long_aggr <- habmap_long_aggr %>%
  left_join(types, by = "type")

habmap_types <- habmap_long_aggr %>%
  filter(!is.na(typelevel)) %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  mutate(type = factor(type,
                       levels = levels(types$type)
                       )
         ) %>%
  arrange(polygon_id, desc(phab))
  
habmap_other_type <- habmap_long_aggr %>%
  filter(is.na(typelevel))

habmap_other_type %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  kable(caption = "Polygons with types that do not correspond with the standard list of types and rbb",
        label = NA) %>%
  kable_styling()

```


## Select features that contain habitat or rbb types

```{r}

habmap_types_sf <- habmap_sf %>%
  select(polygon_id, description_orig) %>%
  filter(polygon_id %in% habmap_types$polygon_id)

```

## Let 's have a quick look at the result

The ID of each polygon in the habitatmap contains the year in which the polygon
was last updated.
In the table below we show the number of records per type and per update year.

```{r}
habmap_types %>%
  mutate(polygon_version = str_sub(polygon_id, start = -4)) %>%
  group_by(type, polygon_version) %>%
  summarise(n_records = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = "polygon_version", values_from = "n_records", values_fill = 0) %>%
  kable() %>%
  kable_styling()
```

## Control and fix geometries

We want to make sure that the layer has valid geometries so that we can use it 
in spatial analysis in GIS. 

First, we search for polygons with invalid geometries. 

```{r}
habmap_types_geom_invalid <- st_is_valid(habmap_types_sf, reason = TRUE) %>%
  as_tibble() %>%
  mutate(id = rownames(.)) %>% 
  left_join(habmap_types_sf %>% 
              mutate(id = as.character(row_number())), 
            by = "id") %>%
  filter(value != "Valid Geometry") %>% 
  mutate(problem = str_extract(string = value, 
                                pattern = "[^\\[]+"))

cat(paste0("There are ", habmap_types_geom_invalid %>% nrow() , 
           " polygons with invalid geometry"))
```

```{r}
habmap_types_geom_invalid  %>%
  select(id, polygon_id, description_orig, problem) %>% 
  kable(caption = "Polygons with faulty geometries", label = NA) %>%
  kable_styling() %>% 
  scroll_box(height = "300px")
```

```{r}
# show the invalid polygons on a map
habmap_types_geom_invalid_wgs84 <- habmap_types_sf %>% 
  inner_join(habmap_types_geom_invalid %>% 
               select(polygon_id, problem), 
             by = join_by(polygon_id)) %>%
  st_transform(crs = 4326) 

leaflet(height = "600px", width = "700px") %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = habmap_types_geom_invalid_wgs84,
              color = "darkred",
              popup = paste("polygon_id:", 
                          habmap_types_geom_invalid_wgs84$polygon_id, "<br>",
                           "description_orig:",
                          habmap_types_geom_invalid_wgs84$description_orig, 
                          "<br>", 
                          habmap_types_geom_invalid_wgs84$problem)) 
```

We correct the geometry using the function `st_make_valid` and check the results.

```{r}
# correct geometries
habmap_types_sf <- st_make_valid(habmap_types_sf)
```

```{r}
# list the corrected polygons and check the geometry again
habmap_types_geom_corrected <- st_is_valid(habmap_types_sf, reason = TRUE) %>% 
  as_tibble() %>% 
  mutate(id = rownames(.)) %>% 
  left_join(habmap_types_sf %>% 
              mutate(id = as.character(row_number())), 
            by = "id") %>% 
  filter(polygon_id %in% habmap_types_geom_invalid$polygon_id)

habmap_types_geom_corrected %>% 
        select(id, polygon_id, description_orig, problem = value) %>% 
  kable(caption = "Polygons with faulty geometries after correction", 
        label = NA) %>% 
  kable_styling() %>% 
  scroll_box(height = "300px")
```


```{r}
# show the corrected polygons on a map
habmap_types_geom_corrected_wgs84 <- habmap_types_sf %>% 
  filter(polygon_id %in% habmap_types_geom_invalid$polygon_id) %>% 
  st_transform(crs = 4326) 

leaflet(height = "600px", width = "700px") %>%
  addTiles(group = "OSM (default)") %>%
    addPolygons(data = habmap_types_geom_invalid_wgs84, 
                group = "before correction (red)",
                color = "darkred") %>% 
  addPolygons(data = habmap_types_geom_corrected_wgs84,
              group = "after correction (blue)",
              popup = paste("polygon_id:", 
                          habmap_types_geom_corrected_wgs84$polygon_id, "<br>",
                           "description_orig:",
                          habmap_types_geom_corrected_wgs84$description_orig)) %>%
    addLayersControl(overlayGroups = c("before correction (red)", "after correction (blue)"),
                     options = layersControlOptions(collapsed = FALSE))
```


## Write results into a geopackage

```{r}
dir.create(file.path(path, "20_processed/habitatmap_stdized"), recursive = TRUE)
filepath <- file.path(path,"20_processed/habitatmap_stdized/habitatmap_stdized.gpkg")
```

```{r}

st_write(habmap_types_sf, 
         filepath, 
         layer = "habitatmap_polygons", 
         driver = "GPKG", 
         delete_dsn = TRUE)

st_write(habmap_types, 
         filepath, 
         layer = "habitatmap_types", 
         driver = "GPKG", 
         append = TRUE)
```
