# Generate standardized habitatmap

## Data source

The shapefile of the BWK and Natura 2000 habitat map of Flanders can be downloaded [here](https://zenodo.org/record/4428002#.YBqzu-hKigY).

To be sure we will use the correct version of the data source (habitatmap_2020), we will derive the md5 file hashes and compare it to the file hashes in the [data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853).

```{r}

path <- fileman_up("n2khab_data")
file <- "10_raw/habitatmap"

mypath <- file.path(path, file)

hashes <-
    tibble(filepath = str_c(mypath, "/",
        list.files(path = mypath,
            recursive = TRUE)
      )) %>%
    mutate(filename = str_match(filepath, "(.+\\/)*(.+)")[,3],
           md5 = map(filepath, function(x) {
                           file(x) %>% md5 %>% str_c(collapse = '')
                         }) %>% as.character,
           md5_ref = c("f767a12540b2bde435c6709c9c9675ad",
                      "358ff672c2fae48eba5bd09f8a671675",
                      "f881f61a6c07741b58cb618d8bbb0b99",
                      "1da89a5dc267bbd427c8c07fcf63f344",
                      "79ae8b4b3c6d970a1f10e78f3eca9ef7"),
           match = md5 == md5_ref) %>%
    select(filename,
           md5,
           md5_ref,
           match)

kable(hashes) %>%
  kable_styling()

if (!all.equal(hashes$md5, hashes$md5_ref)) {
    stop(cat("The source map is NOT up to date ! Please check  the datasource. "))
}
```


```{r read_raw_data}

habitatmap_sf <- st_read(file.path(path, file),
                  layer = "habitatmap", crs = 31370)
```


## Processing of the attribute table

Every polygon in the habitat map can consist of maximum 5 different vegetation types. This information is stored in the columns 'HAB1', 'HAB2',..., 'HAB5' of the attribute table. The estimated fraction of each vegetation type within the polygons is stored in the columns 'PHAB1', 'PHAB2', ..., 'PHAB5'.

We will convert the attribute table to a long format, so that every row contains one vegetation type.


```{r select_polygons}

habmap_sf <- habitatmap_sf %>%
            filter(!(HAB1 == "gh" & PHAB1 == 100)) %>%
            mutate(polygon_id = TAG, # unieke id
                   description_orig = str_c(PHAB1, "% ", HAB1,
                                      ifelse(is.na(HAB2), "", str_c("; ", PHAB2, "% ", HAB2)),
                                      ifelse(is.na(HAB3), "", str_c("; ", PHAB3, "% ", HAB3)),
                                      ifelse(is.na(HAB4), "", str_c("; ", PHAB4, "% ", HAB4)),
                                      ifelse(is.na(HAB5), "", str_c("; ", PHAB5, "% ", HAB5)))) 

```


```{r long_format}

habmap_longPHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    gather(PHAB1, PHAB2, PHAB3, PHAB4, PHAB5, key = "phabn", value = "phab") %>%
    mutate(patch_id = str_sub(phabn, 5, 5)) %>%
    select(polygon_id, patch_id, phab) %>%
    filter(!is.na(phab))

habmap_longHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    select(polygon_id, polygon_area = OPPERVL, starts_with("HAB"), -HABLEGENDE) %>%
    gather(HAB1, HAB2, HAB3, HAB4, HAB5, key = "hab", value = "code") %>%
    mutate(patch_id = substr(hab, 4, 4)) %>%
    filter(!is.na(code)) %>%
    filter(! code %in% c("gh", "x"))

```

## Correction of some of the codes in the Habitat map

For some polygons in the habitat map the vegetation type is not certain. When 
this is the case, the code contains the possible vegetation types separated with
a ','. However, in some cases this way of coding is not used. In that case we
change the code according to the general coding rules (for example 3130_rbbmr is
changed to 3130,rbbmr). This makes processing of the habitat map more 
straightforward. Table \@ref(tab:codeCorrected) shows the corrected codes and the number of polygons for which the correction is applied.


```{r }

habmap_correction <- read_vc("habmap_correction/habmap_correction")

overview_habmap_correction <- habmap_longHAB %>%
  inner_join(habmap_correction, by = "code") %>%
  group_by(code, code_corrected) %>%
  summarise(n_polygons = n()) %>%
  ungroup()

```


```{r codeCorrected}
overview_habmap_correction %>%
  kable(caption = "Corrected codes in habitat map") %>%
  kable_styling()
```

```{r}
if(sum(habmap_correction$code %in% habmap_longHAB$code) > 0){

    habmap_longHAB <- habmap_longHAB %>%
        left_join(habmap_correction, by = "code") %>%
        mutate(code_orig = code,
            code = ifelse(is.na(code_corrected),
                             code,
                             code_corrected)) %>%
        select(-code_corrected)
    
    } else {
      
      habmap_longHAB <- habmap_longHAB %>%
        mutate(code_orig = code)
}
```


## Splitting uncertain types

When the vegetation type is uncertain, and 2 or 3 possible vegetation types are provided, separated with a ','. 
We will split the different possible vegetation types and create one row for each of them. 
An additional variable 'certain' will be FALSE if the orginal habitatmap code consists of 2 or 3 possible vegetation types, and TRUE if only one vegetation type is provided.
An exception to this rule are following codes: `3130,rbbmr`, `3140,rbbmr`, `3150,rbbmr` and `3160,rbbmr`.
These are standing water bodies that contain both `rbbmr` and the habitat type.
Therefore `certain` will be set to `TRUE` for both types included in `code`.

```{r}

habmap_long <- habmap_longHAB %>%
    left_join(habmap_longPHAB, by = c("polygon_id", "patch_id")) %>%
    mutate(certain = !str_detect(code, ","),
           certain = ifelse(code %in% c("3130,rbbmr", "3140,rbbmr", "3150,rbbmr", "3160,rbbmr"), TRUE, certain)) %>%
    separate(code,
             into = c("type1", "type2", "type3"),
             sep = ",",
             remove = FALSE) %>%
    gather(type1, type2, type3, key = "ntype", value = "type") %>%
    filter(!is.na(type)) %>%
    filter(!(type %in% c("gh", "bos"))) %>%
    select(-patch_id)

```

Sometimes you get two records for the same type within the same habitatmap polygon.
We distinguish two cases: 

+ two certain records of the same type, for example:

```{r}
habmap_long %>%
  filter(polygon_id == "357179_v2020")
```

In this case we will sum the phab-values and create one record for each type.

+ a certain and an uncertain record of the same type, for example:

```{r}
habmap_long %>%
  filter(polygon_id == "639375_v2014")
```

We will not aggregate the certain and uncertain record, because we will loose some information used to create `habitatmap_terr`. It is some of the uncertain records (such as `9120, gh`) are eliminated when processing `habitatmap_terr`. 

```{r}
check_double_type <- habmap_long %>%
  group_by(polygon_id, type) %>%
  mutate(n_type = n()) %>%
  ungroup() %>%
  filter(n_type > 1)
  
```


```{r}
habmap_long_aggr <- habmap_long %>%
  # group_by(polygon_id, code) %>%
  #     mutate(n = n()) %>%
  #   ungroup() %>%
    group_by(polygon_id, type, certain) %>%
     summarise(phab = sum(phab),
              code_orig = str_c(code, collapse = "; "),
              certain = all(certain)) %>%
    ungroup() 
```
See below the result of the aggregation for both examples. 

```{r}
habmap_long_aggr %>%
  filter(polygon_id == "357179_v2020")
```

```{r}
habmap_long_aggr %>%
  filter(polygon_id == "639375_v2014")
```


## Select vegetation types that belong to the standard list of habitat and rbb types  

Table \@ref(tab:selectTypes) shows the records with habitat types that do not belong to the standard list of habitat and rbb types. These records are filtered out. 

```{r selectTypes}

types <- read_types() %>%
  select(type, typelevel, main_type, typeclass)

habmap_long_aggr <- habmap_long_aggr %>%
  left_join(types, by = "type")

habmap_types <- habmap_long_aggr %>%
  filter(!is.na(typelevel)) %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  mutate(type = factor(type,
                       levels = levels(types$type)
                       )
         ) %>%
  arrange(polygon_id, desc(phab))
  
habmap_other_type <- habmap_long_aggr %>%
  filter(is.na(typelevel))

habmap_other_type %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  kable(caption = "Polygons with types that do not coorspond with the standard list of types and rbb") %>%
  kable_styling()

```


## Select features that contain habitat or rbb types

```{r}

habmap_types_sf <- habmap_sf %>%
  select(polygon_id, description_orig) %>%
  filter(polygon_id %in% habmap_types$polygon_id)

```

## Write results into a geopackage

```{r}
dir.create(file.path(path, "20_processed/habitatmap_stdized"), recursive = TRUE)
filepath <- file.path(path,"20_processed/habitatmap_stdized/habitatmap_stdized.gpkg")
```

```{r, eval = FALSE}

st_write(habmap_types_sf, 
         filepath, 
         layer = "habitatmap_polygons", 
         driver = "GPKG", 
         delete_dsn = TRUE)

st_write(habmap_types, 
         filepath, 
         layer = "habitatmap_types", 
         driver = "GPKG", 
         append = TRUE)

con <- dbConnect(SQLite(), filepath)
updatequery <- sprintf("UPDATE gpkg_contents SET last_change = '%s'",
                       gpkg_timestamp)
dbExecute(con, updatequery) %>% 
  message(" rows of the gpkg_contents table have been set with timestamp ",
          gpkg_timestamp)
dbDisconnect(con)
```
