# Generate standardized habitatmap

We start from the (unpublished) interim version of the habitatmap of April 2026 (official name: habitatmap_2025_v99_interim).

The interim habitatmap has not been cleaned up as well as the official published 
versions:
it contains inconsistent field names, the TAG and globalid_BWK fields do not contain unique 
identifiers, the empty fields contain an empty string or one or more spaces, ...
(please refer to miscellaneous > habitatmap.Rmd for a full overview of the 
possible problems and needed corrections).

Because of these problems, it is not a good idea to load the raw version 
of the habitatmap with `n2khab::read_habitatmap()` (there would be many corrections 
needed in `n2khab::read_habitatmap()`, and possibly different corrections for each 
new interim version of the habitatmap).
Therefore we decide to skip the `n2khab::read_habitatmap()` step for the interim versions.

## Data source

- The interim version of the BWK and Natura 2000 habitat map of 
Flanders is not online, and will be imported manually.

To be sure we will use the correct version of the data source (habitatmap_2025_v99_interim), 
we will derive the md5 file checksums and compare it to the file checksums in the 
[data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853).

```{r}
path <- locate_n2khab_data()
file <- "10_raw/habitatmap"

mypath <- file.path(path, file)

checksums <-
  tibble(filepath = str_c(
    mypath, "/",
    list.files(
      path = mypath,
      recursive = TRUE
    )
  )) %>%
  mutate(
    filename = str_match(filepath, "(.+\\/)*(.+)")[, 3],
    md5 = unname(md5sum(filepath)),
    md5_ref = c("b7e88d0687bf27db723aab9845a85a80"),
    match = md5 == md5_ref
  ) %>%
  select(
    filename,
    md5,
    md5_ref,
    match
  )

kable(checksums) %>%
  kable_styling()

if (!all.equal(checksums$md5, checksums$md5_ref)) {
  stop(
    "The data source is not at the right version! ",
    "Please check the data source."
  )
}
```

Because it was found that sf has difficulties with correcting certain (invalid) MULTISURFACE geometries, terra is being used to read the file and the result is converted to sf: terra automatically casts all geometries to polygons on reading; after conversion to sf we get no missing reasons for invalidity and `st_make_valid()` works for all invalid geometries.
See `src/miscellaneous/habitatmap.Rmd` for more details.

```{r read_raw_data}
habitatmap_sf <-
  vect(file.path(mypath, "habitatmap.gpkg")) %>%
  st_as_sf() %>%
  as_tibble() %>%
  st_as_sf(crs = 31370) %>%
  # since we get a combination of POLYGON and MULTIPOLYGON, we entirely cast
  # as MULTIPOLYGON
  st_cast("MULTIPOLYGON")
```

## Clean up the interim version

See `src/miscellaneous/habitatmap.Rmd` for the underlying exploration.

```{r clean_up_interim}
duplic_globalid <-
  habitatmap_sf %>%
  st_drop_geometry() %>%
  count(globalid_BWK) %>%
  filter(n > 1)
# we want to make the interim version as similar as possible to the regular one
# to avoid too many changes in the other chunks
habitatmap_sf <- habitatmap_sf %>%
  rename(
    PHAB1 = pHAB1,
    PHAB2 = pHAB2,
    PHAB3 = pHAB3,
    PHAB4 = pHAB4,
    PHAB5 = pHAB5,
    source = herk_bwk,
    source_hab = HERK_HAB,
    OPPERVL = SHAPE_Area
  ) %>%
  # add unique ID
  mutate(
    TAG = str_c(
      globalid_BWK,
      ifelse(
        globalid_BWK %in% duplic_globalid$globalid_BWK,
        str_c("_", str_pad(row_number(), 2, pad = "0")),
        ""
      )
    ),
    .by = globalid_BWK
  ) %>% 
  # clean up empty fields (empty strings, spaces, "<Null>" strings) and remove
  # leading or trailing whitespaces
  mutate(
    across(-geometry, ~ ifelse(grepl(pattern = "^\\s*$", .), NA, .)),
    across(where(is.character), str_trim),
    info = ifelse(info == "<Null>", NA, info)
  ) %>%
  filter(!is.na(HAB1)) # remove a few errors
```


## Filter the data source according to habitat or rib types

We also add a `description_orig` column to store original codes.

```{r select_polygons}
habmap_sf <- habitatmap_sf %>%
  filter(!(HAB1 == "gh" & PHAB1 == 100)) %>%
  mutate(
    polygon_id = TAG, # unique id
    description_orig = str_c(
      PHAB1, "% ", HAB1,
      ifelse(is.na(HAB2), "", str_c("; ", PHAB2, "% ", HAB2)),
      ifelse(is.na(HAB3), "", str_c("; ", PHAB3, "% ", HAB3)),
      ifelse(is.na(HAB4), "", str_c("; ", PHAB4, "% ", HAB4)),
      ifelse(is.na(HAB5), "", str_c("; ", PHAB5, "% ", HAB5))
    )
  )
```


## Processing the attribute table

### Converting to long format

Every polygon in the habitat map can consist of maximum 5 different vegetation types. This information is stored in the columns 'HAB1', 'HAB2',..., 'HAB5' of the attribute table. The estimated fraction of each vegetation type within the polygons is stored in the columns 'PHAB1', 'PHAB2', ..., 'PHAB5'.

We will convert the attribute table to a long format, so that every row contains one vegetation type.


```{r long_format}
habmap_longPHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    gather(PHAB1, PHAB2, PHAB3, PHAB4, PHAB5, key = "phabn", value = "phab") %>%
    mutate(patch_id = str_sub(phabn, 5, 5)) %>%
    select(polygon_id, patch_id, phab) %>%
    filter(!is.na(phab))

habmap_longHAB <- habmap_sf %>%
    st_drop_geometry() %>%
    select(
      polygon_id,
      starts_with("HAB"),
      -any_of(c("HABLEGENDE"))
    ) %>%
    gather(HAB1, HAB2, HAB3, HAB4, HAB5, key = "hab", value = "code") %>%
    mutate(patch_id = substr(hab, 4, 4)) %>%
    filter(!is.na(code)) %>%
    filter(!code %in% c("gh", "x"))
```

Here are the vegetation codes used in this version of the habitatmap:

```{r showListHabitats}
habmap_longHAB %>%
  distinct(code) %>%
  arrange(code) %>%
  kable(caption = "List codes in habitat map", label = NA) %>%
  kable_styling() %>%
  scroll_box(height = "300px")
```

### Correction of some of the codes in the Habitat map

Some polygons in the habitat map contain codes that do not correspond with the standardized list of habitat types. 
We correct these codes to make processing of the habitat map more straightforward.
Table \@ref(tab:codeCorrected) shows the corrected codes and the number of polygons for which the correction is applied.


```{r }
habmap_correction <- read_vc("habmap_correction/habmap_correction")

overview_habmap_correction <- habmap_longHAB %>%
  inner_join(
    habmap_correction,
    join_by(code),
    relationship = "many-to-one",
    unmatched = "drop"
  ) %>%
  summarise(
    n_polygons = n(),
    .by = c(code, code_corrected)
  ) %>%
  arrange(code_corrected)
```


```{r codeCorrected}
overview_habmap_correction %>%
  kable(caption = "Corrected codes in habitat map", label = NA) %>%
  kable_styling()
```

```{r}
if (sum(habmap_correction$code %in% habmap_longHAB$code) > 0) {
  habmap_longHAB <- habmap_longHAB %>%
    left_join(
      habmap_correction,
      join_by(code),
      relationship = "many-to-one",
      unmatched = "drop"
    ) %>%
    mutate(
      code_orig = code,
      code = ifelse(is.na(code_corrected),
        code,
        code_corrected
      )
    ) %>%
    select(-code_corrected)
} else {
  habmap_longHAB <- habmap_longHAB %>%
    mutate(code_orig = code)
}
```


### Splitting codes that contain different types

In several cases the code contains 2 or 3 possible vegetation types which are separated with a ','. 
We will split the different possible vegetation types and create one row for each of them. 
An additional variable 'certain' will be FALSE if the original habitatmap code consists of 2 or 3 possible vegetation types, and TRUE if only one vegetation type is provided.

An exception to this rule are following codes: `3130_rbbmr`, `3140_rbbmr`, `3150_rbbmr` and `3160_rbbmr` (converted in the previous step into `31x0,rbbmr`).
These are standing water bodies that contain both `rbbmr` and the habitat type.
Therefore `certain` will be set to `TRUE` for both types included in `code`.

```{r}
habmap_long <- 
  habmap_longHAB %>%
  left_join(
    habmap_longPHAB,
    join_by(polygon_id, patch_id),
    relationship = "one-to-one",
    unmatched = "drop"
  ) %>%
  mutate(
    certain = !str_detect(code, ","),
    certain = ifelse(code_orig %in%
      c("3130_rbbmr", "3140_rbbmr", "3150_rbbmr", "3160_rbbmr"),
    TRUE,
    certain
    )
  ) %>%
  separate(
    code,
    into = c("type1", "type2", "type3"),
    sep = ",",
    fill = "right",
    remove = FALSE
  ) %>%
  gather(type1, type2, type3, key = "ntype", value = "type") %>%
  filter(!is.na(type)) %>%
  filter(!(type %in% c("gh", "bos"))) %>%
  select(-patch_id)
```

Sometimes you get two records for the same type within the same habitatmap polygon.
We distinguish two cases: 

+ two certain or two uncertain records of the same type, for example:

```{r}
example <- habmap_long %>%
  add_count(polygon_id, type, certain) %>% 
  filter(n > 1)

polygon_id_example_1 <- example$polygon_id[1]

habmap_long %>%
  filter(polygon_id == polygon_id_example_1)
```

In this case we will sum the phab-values and create one record for each type.

+ a certain and an uncertain record of the same type, for example:

```{r}
example <- habmap_long %>%
  filter(
    n() > 1,
    any(certain),
    any(!certain),
    .by = c(polygon_id, type)
  )

polygon_id_example_2 <- example$polygon_id[1]

habmap_long %>%
  filter(polygon_id == polygon_id_example_2)
```

We will not aggregate the certain and uncertain record, because we will lose some information used to create `habitatmap_terr`. It is some of the uncertain records (such as `9120, gh`) are eliminated when processing `habitatmap_terr`. 

```{r include=FALSE, eval=FALSE}
check_double_type <- habmap_long %>%
  add_count(polygon_id, type, name = "n_type") %>%
  filter(n_type > 1) %>% 
  arrange(polygon_id)
```


```{r}
habmap_long_aggr <- habmap_long %>%
  summarise(
    phab = sum(phab),
    code_orig = str_c(code_orig, collapse = "; "),
    .by = c(polygon_id, type, certain)
  )
```

See below the result of the aggregation for the first example.

```{r}
habmap_long_aggr %>%
  filter(polygon_id == polygon_id_example_1)
```

### Select vegetation types that belong to the standard list of habitat and rbb types  

Table \@ref(tab:selectTypes) shows the records with habitat types that do not belong to the standard list of habitat and rbb types.

Good news! 
The table is empty.

```{r selectTypes}
types <- read_types() %>%
  select(type, typelevel, main_type, typeclass)

habmap_long_aggr <- habmap_long_aggr %>%
  left_join(
    types,
    join_by(type),
    relationship = "many-to-one",
    unmatched = "drop"
  )

habmap_types <- habmap_long_aggr %>%
  filter(!is.na(typelevel)) %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  mutate(type = factor(type, levels = levels(types$type))) %>%
  arrange(polygon_id, desc(phab))
  
habmap_other_type <- habmap_long_aggr %>%
  filter(is.na(typelevel))

habmap_other_type %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  kable(
    caption = "Polygons with types that do not correspond with the standard list of types and rbb",
    label = NA
  ) %>%
  kable_styling()
```


## Transforming the spatial object

### Only select features that contain habitat or rbb types

```{r}
habmap_types_sf <- habmap_sf %>%
  filter(polygon_id %in% habmap_types$polygon_id)
```

### Control and fix geometries

We want to make sure that the layer has valid geometries so that we can use it 
in spatial analysis in GIS. 

First, we search for polygons with invalid geometries (see `src/miscellaneous/habitatmap.Rmd` for more details). 

```{r find_invalid_geoms}
validities <- st_is_valid(habmap_types_sf, reason = TRUE) %>%
  as_tibble() %>%
  mutate(id = rownames(.) ) %>%
  mutate(problem = str_extract(string = value, pattern = "[^\\[]+")) %>%
  left_join(habmap_types_sf %>% 
              mutate(id = as.character(row_number())), 
            by = "id") %>% 
  select(-value)

# There are invalid polygons with a reason mentioned but some may have no reason
# given (NA). Those categories need to be handled separately

# invalid polygons with a clear explanation
habmap_types_geom_invalid <- validities %>%
  filter(problem != "Valid Geometry") 
# invalid polygons without a clear explanation: NA
habmap_types_geom_invalidna <- validities %>% 
  filter(is.na(problem)) 

message(
  "There are ",
  rbind(habmap_types_geom_invalid,habmap_types_geom_invalidna) %>% nrow(),
  " polygons with invalid geometry: \n",
  habmap_types_geom_invalid %>% nrow(),
  " with a clear explanation for the problem and \n",
  habmap_types_geom_invalidna %>% nrow(),
  " without any clear explanation (NA)."
)
```

```{r include=FALSE}
# The code may not work correctly if cases exist with invalid NA (always
# investigate; consult code history for remedies in the past); hence protecting
# with a test
stopifnot(nrow(habmap_types_geom_invalidna) == 0)
```

```{r}
habmap_types_geom_invalid  %>%
  rbind(., habmap_types_geom_invalidna) %>% 
  select(id, polygon_id, description_orig, problem) %>% 
  kable(caption = "Polygons with faulty geometries", label = NA) %>%
  kable_styling() %>% 
  scroll_box(height = "300px")
```

We will correct the geometry using the function `st_make_valid()` and check the results.

Let us take a look at the invalid polygons:

```{r map_invalid_geoms}
# show the invalid polygons on a map
habmap_types_geom_invalid_wgs84 <- habmap_types_geom_invalid %>%
  st_as_sf() %>% 
  mutate(problem = ifelse(is.na(problem), "Unknown (NA)", problem)) %>% 
  st_transform(crs = 4326) 

pal <- colorFactor(palette = c("darkorange", "darkred"), 
                   domain = habmap_types_geom_invalid_wgs84$problem)

leaflet(height = "600px", width = "700px") %>%
  inbospatial::add_wms_be_cartoweb("topo_grey") %>%
  addPolygons(data = habmap_types_geom_invalid_wgs84,
              color = pal(habmap_types_geom_invalid_wgs84$problem),
              popup = paste("polygon_id:", 
                          habmap_types_geom_invalid_wgs84$polygon_id, "<br>",
                           "description_orig:",
                          habmap_types_geom_invalid_wgs84$description_orig, 
                          "<br>", 
                          habmap_types_geom_invalid_wgs84$problem)) %>% 
  addLegend(data = habmap_types_geom_invalid_wgs84,
            pal = pal, values = ~problem,
            opacity = 1)
```

We correct the geometries:

```{r fix_geom}
habmap_types_sf <- st_make_valid(habmap_types_sf)
```

```{r }
# list the corrected polygons and check the geometry again
habmap_types_geom_corrected <- st_is_valid(habmap_types_sf, reason = TRUE) %>% 
  as_tibble() %>% 
  mutate(id = rownames(.)) %>% 
  left_join(habmap_types_sf %>% 
              mutate(id = as.character(row_number())), 
            by = "id") %>% 
  filter(polygon_id %in% habmap_types_geom_invalid$polygon_id)

habmap_types_geom_corrected %>% 
        select(id, polygon_id, description_orig, problem = value) %>% 
  kable(caption = "Polygons with faulty geometries after correction", 
        label = NA) %>% 
  kable_styling() %>% 
  scroll_box(height = "300px")
```


```{r map_corrected_geoms}
# show the corrected polygons on a map
habmap_types_geom_corrected_wgs84 <- habmap_types_sf %>% 
  filter(polygon_id %in% habmap_types_geom_invalid$polygon_id) %>% 
  st_transform(crs = 4326) 

leaflet(height = "600px", width = "700px") %>%
  inbospatial::add_wms_be_cartoweb("topo_grey") %>%
  addPolygons(data = habmap_types_geom_invalid_wgs84, 
              group = "before correction (red)",
              color = "darkred") %>% 
  addPolygons(data = habmap_types_geom_corrected_wgs84,
              group = "after correction (blue)",
              popup = paste("polygon_id:", 
                            habmap_types_geom_corrected_wgs84$polygon_id, "<br>",
                            "description_orig:",
                            habmap_types_geom_corrected_wgs84$description_orig)) %>%
  addLayersControl(overlayGroups = c("before correction (red)", "after correction (blue)"),
                   options = layersControlOptions(collapsed = FALSE))
```

### Add the assessment year and method for each feature

```{r sources}
habmap_sf <- 
  habmap_sf %>% 
  mutate(
    source_for_year = case_when(
      str_detect(source_hab, "^\\d{2}") ~ "hab",
      # only turn to BWK year if the habitat type was translated automatically
      # without further alteration
      str_detect(source, "^\\d{2}") & 
        str_detect(source_hab, "^\\d*?(a|A)") ~ "bwk",
      .default = NA_character_
      ) %>% 
      factor(),
    year_short = recode_values(
      source_for_year,
      "hab" ~ str_sub(source_hab, 1, 2),
      "bwk" ~ str_sub(source, 1, 2),
      ),
    year_short = as.integer(year_short),
    year_assessment = ifelse(
      year_short < 90,
      2000L + year_short,
      1900L + year_short
    ),
    method_abbrev = 
      # extract the methods, not the season-letters [vznw] that may be added to
      # fieldwork years:
      str_extract(source_hab, "^\\d*?([a-mo-ux-yA-MO-UX-Y]{1,2})", group = 1),
    method_assessment = case_when(
      method_abbrev %in% c("A", "a") ~ "Automated translation from BWK",
      method_abbrev == "mk" ~ 
        "Adjustment of BWK-translation, based on field knowledge",
      method_abbrev == "o" ~ 
        "Adjustment of BWK-translation, based on orthophoto interpretation",
      method_abbrev == "l" ~ "Assessment based on literature or extra files",
      method_abbrev == "ex" ~ "Field assessment by an external expert",
      is.na(source_for_year) ~ "Field assessment but year is missing??",
      .default = "Field assessment"
    ) %>% 
      factor()
  ) %>% 
  # checking that no other methods occur
  {
    fieldobs <- filter(
      .,
      str_detect(method_assessment, "Field"),
      !str_detect(method_assessment, "external")
    )
    if (any(!is.na(fieldobs$method_abbrev))) {
      stop("Unsupported method abbreviation detected in source_hab")
    } else {
      .
    }
  } %>% 
  select(-source_for_year, -year_short, -method_abbrev)
```

### Select specific columns

```{r}
habmap_types_sf <- habmap_sf %>%
  select(polygon_id, description_orig, year_assessment, method_assessment)
```


## Write results to a geopackage

```{r}
dir.create(file.path(path, "20_processed/habitatmap_stdized"), recursive = TRUE)
filepath <- file.path(path, "20_processed/habitatmap_stdized/habitatmap_stdized.gpkg")
```

```{r write_gpkg}
st_write(
  habmap_types_sf,
  filepath,
  layer = "habitatmap_polygons",
  driver = "GPKG",
  delete_dsn = TRUE
)

st_write(
  habmap_types,
  filepath,
  layer = "habitatmap_types",
  driver = "GPKG",
  append = TRUE
)
```
