# Generate standardized habitatmap

## Data source

The shapefile of the BWK and Natura 2000 habitat map of Flanders can be downloaded [here](https://doi.org/10.5281/zenodo.3354381).

To be sure we will use the correct version of the data source (habitatmap_2025), we will derive the md5 file checksums and compare it to the file checksums in the [data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853).

```{r}
path <- locate_n2khab_data()
file <- "10_raw/habitatmap"

mypath <- file.path(path, file)

checksums <-
  tibble(filepath = str_c(
    mypath, "/",
    list.files(
      path = mypath,
      recursive = TRUE
    )
  )) %>%
  mutate(
    filename = str_match(filepath, "(.+\\/)*(.+)")[, 3],
    md5 = unname(md5sum(filepath)),
    md5_ref = c(
      "12372851086ee8207407c86905426ee2",
      "59fadca0e8a3ce09fc19b2f0fa72009b",
      "f881f61a6c07741b58cb618d8bbb0b99",
      "1b9a4efa9b94d8906f9ff0903831abe0",
      "aa79a21216c16f7fe0263ed4ba907ed9",
      "7ff6cc0efbd34f4cfef52d73e466df0d"
    ),
    match = md5 == md5_ref
  ) %>%
  select(
    filename,
    md5,
    md5_ref,
    match
  )

kable(checksums) %>%
  kable_styling()

if (!all.equal(checksums$md5, checksums$md5_ref)) {
  stop(
    "The data source is not at the right version! ",
    "Please check the data source."
  )
}
```


```{r read_raw_data}
habitatmap_sf <- read_habitatmap(
  filter_hab = FALSE,
  fix_geom = TRUE,
  version = "habitatmap_2025"
)
```


## Filter the data source according to habitat or rib types

We also add a `description_orig` column to store original codes.

```{r select_polygons}
habmap_sf <- habitatmap_sf %>%
  filter(!(hab1 == "gh" & phab1 == 100)) %>%
  mutate(
    description_orig = str_c(
      phab1, "% ", hab1,
      ifelse(is.na(hab2), "", str_c("; ", phab2, "% ", hab2)),
      ifelse(is.na(hab3), "", str_c("; ", phab3, "% ", hab3)),
      ifelse(is.na(hab4), "", str_c("; ", phab4, "% ", hab4)),
      ifelse(is.na(hab5), "", str_c("; ", phab5, "% ", hab5))
    )
  )
```


## Processing the attribute table

### Converting to long format

Every polygon in the habitat map can consist of maximum 5 different vegetation types. This information is stored in the columns 'HAB1', 'HAB2',..., 'HAB5' of the attribute table. The estimated fraction of each vegetation type within the polygons is stored in the columns 'PHAB1', 'PHAB2', ..., 'PHAB5'.

We will convert the attribute table to a long format, so that every row contains one vegetation type.

```{r long_format}
habmap_longPHAB <- habmap_sf %>%
  st_drop_geometry() %>%
  gather(phab1, phab2, phab3, phab4, phab5, key = "phabn", value = "phab") %>%
  mutate(patch_id = str_sub(phabn, 5, 5)) %>%
  select(polygon_id, patch_id, phab) %>%
  filter(!is.na(phab))

habmap_longHAB <- habmap_sf %>%
  st_drop_geometry() %>%
  select(polygon_id, polygon_area = area_m2, starts_with("hab"), -hab_legend) %>%
  gather(hab1, hab2, hab3, hab4, hab5, key = "hab", value = "code") %>%
  mutate(patch_id = substr(hab, 4, 4)) %>%
  filter(!is.na(code)) %>%
  filter(!code %in% c("gh", "x"))
```

Here are the vegetation codes used in this version of the habitatmap:

```{r showListHabitats}
habmap_longHAB %>%
  distinct(code) %>%
  arrange(code) %>%
  kable(caption = "List codes in habitat map", label = NA) %>%
  kable_styling() %>%
  scroll_box(height = "300px")
```

### Correction of some of the codes in the Habitat map

Some polygons in the habitat map contain codes that do not correspond with the standardized list of habitat types. 
We correct these codes to make processing of the habitat map more straightforward.
Table \@ref(tab:codeCorrected) shows the corrected codes and the number of polygons for which the correction is applied.


```{r }
habmap_correction <- read_vc("habmap_correction/habmap_correction")

overview_habmap_correction <- habmap_longHAB %>%
  inner_join(habmap_correction, by = "code") %>%
  summarise(
    n_polygons = n(),
    .by = c(code, code_corrected)
  ) %>% 
  arrange(code_corrected)
```


```{r codeCorrected}
overview_habmap_correction %>%
  kable(caption = "Corrected codes in habitat map", label = NA) %>%
  kable_styling()
```

```{r}
if (sum(habmap_correction$code %in% habmap_longHAB$code) > 0) {
  habmap_longHAB <- habmap_longHAB %>%
    left_join(habmap_correction, by = "code") %>%
    mutate(
      code_orig = code,
      code = ifelse(is.na(code_corrected),
        code,
        code_corrected
      )
    ) %>%
    select(-code_corrected)
} else {
  habmap_longHAB <- habmap_longHAB %>%
    mutate(code_orig = code)
}
```


### Splitting codes that contain different types

In several cases the code contains 2 or 3 possible vegetation types which are separated with a ','. 
We will split the different possible vegetation types and create one row for each of them. 
An additional variable 'certain' will be FALSE if the original habitatmap code consists of 2 or 3 possible vegetation types, and TRUE if only one vegetation type is provided.

An exception to this rule are following codes: `3130_rbbmr`, `3140_rbbmr`, `3150_rbbmr` and `3160_rbbmr` (converted in the previous step into `31x0,rbbmr`).
These are standing water bodies that contain both `rbbmr` and the habitat type.
Therefore `certain` will be set to `TRUE` for both types included in `code`.

```{r}
habmap_long <- habmap_longHAB %>%
  left_join(habmap_longPHAB, by = c("polygon_id", "patch_id")) %>%
  mutate(
    certain = !str_detect(code, ","),
    certain = ifelse(code_orig %in%
      c("3130_rbbmr", "3140_rbbmr", "3150_rbbmr", "3160_rbbmr"),
    TRUE,
    certain
    )
  ) %>%
  separate(code,
    into = c("type1", "type2", "type3"),
    sep = ",",
    remove = FALSE
  ) %>%
  gather(type1, type2, type3, key = "ntype", value = "type") %>%
  filter(!is.na(type)) %>%
  filter(!(type %in% c("gh", "bos"))) %>%
  select(-patch_id)
```

Sometimes you get two records for the same type within the same habitatmap polygon.
We distinguish two cases: 

+ two certain or two uncertain records of the same type, for example:

```{r}
habmap_long %>%
  filter(polygon_id == "357179_v2020")
```

In this case we will sum the phab-values and create one record for each type.

+ a certain and an uncertain record of the same type, for example:

```{r}
example <- habmap_long %>%
  group_by(polygon_id, type) %>%
  filter(n() > 1) %>%
  filter(any(certain) & any(!certain)) %>%
  ungroup()

polygon_id_example <- (example %>%
  slice_head(n = 1))$polygon_id

habmap_long %>%
  filter(polygon_id == polygon_id_example)
```

We will not aggregate the certain and uncertain record, because we will loose some information used to create `habitatmap_terr`. It is some of the uncertain records (such as `9120, gh`) are eliminated when processing `habitatmap_terr`. 

```{r include=FALSE, eval=FALSE}
check_double_type <- habmap_long %>%
  group_by(polygon_id, type) %>%
  mutate(n_type = n()) %>%
  ungroup() %>%
  filter(n_type > 1)
```


```{r}
habmap_long_aggr <- habmap_long %>%
  summarise(
    phab = sum(phab),
    code_orig = str_c(code_orig, collapse = "; "),
    certain = all(certain),
    .by = c(polygon_id, type, certain)
  )
```

See below the result of the aggregation for the first example. 

```{r}
habmap_long_aggr %>%
  filter(polygon_id == "357179_v2020")
```

### Select vegetation types that belong to the standard list of habitat and rbb types  

Table \@ref(tab:selectTypes) shows the records with habitat types that do not belong to the standard list of habitat and rbb types.

Good news! 
The table is empty

```{r selectTypes}
types <- read_types() %>%
  select(type, typelevel, main_type, typeclass)

habmap_long_aggr <- habmap_long_aggr %>%
  left_join(types, by = "type")

habmap_types <- habmap_long_aggr %>%
  filter(!is.na(typelevel)) %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  mutate(type = factor(type, levels = levels(types$type))) %>%
  arrange(polygon_id, desc(phab))

habmap_other_type <- habmap_long_aggr %>%
  filter(is.na(typelevel))

habmap_other_type %>%
  select(polygon_id, type, certain, code_orig, phab) %>%
  kable(
    caption = "Polygons with types that do not correspond with the standard list of types and rbb",
    label = NA
  ) %>%
  kable_styling()
```

### Let 's have a quick look at the result

The ID of each polygon in the habitatmap contains the year in which the polygon
was last updated.
In the table below we show the number of records per type and per update year.

```{r}
habmap_types %>%
  mutate(polygon_version = str_sub(polygon_id, start = -4)) %>%
  summarise(
    n_records = n(),
    .by = c(type, polygon_version)
  ) %>%
  pivot_wider(
    names_from = "polygon_version",
    names_sort = TRUE,
    values_from = "n_records",
    values_fill = 0
  ) %>%
  arrange(type) %>% 
  kable() %>%
  kable_styling()
```

## Transforming the spatial object

### Only select features that contain habitat or rib types

```{r}
habmap_sf <- habmap_sf %>%
  filter(polygon_id %in% habmap_types$polygon_id)
```

### Add the assessment year and method for each feature

```{r sources}
habmap_sf <- 
  habmap_sf %>% 
  mutate(
    source_for_year = case_when(
      str_detect(source_hab, "^\\d{2}") ~ "hab",
      # only turn to BWK year if the habitat type was translated automatically
      # without further alteration
      str_detect(source, "^\\d{2}") & 
        str_detect(source_hab, "^\\d*?(a|A)") ~ "bwk",
      .default = NA_character_
      ) %>% 
      factor(),
    year_short = recode_values(
      source_for_year,
      "hab" ~ str_sub(source_hab, 1, 2),
      "bwk" ~ str_sub(source, 1, 2),
      ),
    year_short = as.integer(year_short),
    year_assessment = ifelse(
      year_short < 90,
      2000L + year_short,
      1900L + year_short
    ),
    method_abbrev = 
      str_extract(source_hab, "^\\d*?([a-gi-ux-yA-GI-UX-Y]{1,2})", group = 1),
    method_assessment = case_when(
      method_abbrev %in% c("A", "a") ~ "Automated translation from BWK",
      method_abbrev == "mk" ~ "Translation adjusted xxxxxxx",
      method_abbrev == "ex" ~ "External source",
      method_abbrev == "l" ~ "Literature",
      method_abbrev == "o" ~ "Orthophoto interpretation",
      method_abbrev == "n" ~ "???",
      is.na(source_for_year) ~ "Field observation but year is missing??",
      .default = "Field observation"
    ) %>% 
      factor()
  ) %>% 
  # checking that no other methods occur
  {
    fieldobs <- filter(., str_detect(method_assessment, "Field"))
    if (any(!is.na(fieldobs$method_abbrev))) {
      stop("Unsupported method abbreviations detected in source_hab")
    } else {
      .
    }
  } %>% 
  select(-source_for_year, -year_short, -method_abbrev)
```

### Select specific columns

```{r}
habmap_types_sf <- habmap_sf %>%
  select(polygon_id, description_orig, year_assessment, method_assessment)
```


## Write results to a geopackage

```{r}
dir.create(file.path(path, "20_processed/habitatmap_stdized"), recursive = TRUE)
filepath <- file.path(path, "20_processed/habitatmap_stdized/habitatmap_stdized.gpkg")
```

```{r}
st_write(habmap_types_sf,
  filepath,
  layer = "habitatmap_polygons",
  driver = "GPKG",
  delete_dsn = TRUE
)

st_write(habmap_types,
  filepath,
  layer = "habitatmap_types",
  driver = "GPKG",
  append = TRUE
)
```
