## Checks on the data source

Checksums:

```{r}
c("xxh64", "md5", "sha256") %>%
  map_dfr(~ tibble(
    algorithm = .,
    checksum = checksum(filepath, .)
  )) %>%
  kable()
```


```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath, layer = "habitatmap_polygons"))
```

```{r paged.print=FALSE, warning=FALSE}
(types <- read_sf(filepath, layer = "habitatmap_types"))
```

- Do all records in the habitatmap_types table have a corresponding spatial feature in the habitatmap_polygons layer?

```{r}
check_polygon_id_types <- types %>%
  anti_join(pol, by = "polygon_id")

nrow(check_polygon_id_types) == 0
```

- Do all spatial features in the habitatmap_polygons layer have a corresponding record in the habitatmap_types table?

```{r}
check_polygon_id_polygons <- pol %>%
  st_drop_geometry() %>%
  anti_join(types, by = "polygon_id")

nrow(check_polygon_id_polygons) == 0
```

- Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object? 
  
```{r} 
types %>%
  distinct(polygon_id) %>%
  nrow() == nrow(pol)
``` 

- Is the CRS conform EPSG:31370?

```{r} 
st_crs(pol) == st_crs(31370)
```

- Is phab always between 0 and 100?

```{r}
all(between(types$phab, 0, 100))
```

- Is the primary key unique?

```{r}
types %>%
  select(polygon_id, type, certain) %>%
  duplicated() %>%
  negate(any)(.)
```

- Are the geometries valid?

```{r}
validities <- st_is_valid(pol)
sum(!validities | is.na(validities)) == 0
```

### Type x assessment year

In the table below we show the number of records per type and assessment year.

```{r}
pol %>% 
  st_drop_geometry() %>% 
  inner_join(
    types,
    join_by(polygon_id),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>%
  count(type, year_assessment) %>% 
  pivot_wider(
    names_from = year_assessment,
    names_sort = TRUE,
    values_from = n,
    values_fill = 0
  ) %>%
  arrange(type) %>% 
  kable() %>%
  kable_styling() %>%
  scroll_box(height = "600px")
```
