# Check results

```{r}
filepath <- file.path(datapath, "sg_extended.gpkg")
```

- Checksums:

```{r}
c("xxh64", "md5", "sha256") %>% 
  map_dfr(~tibble(algorithm = ., 
                  checksum = checksum(filepath, .))) %>% 
  kable
```

- Contents:

```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath))
glimpse(pol)
```

- Is the CRS conform EPSG:31370?

```{r} 
 st_crs(pol) == st_crs(31370) 
```

- Sensible results?

```{r include=FALSE}
image <- file.path(local_root, "images/qgis_shot.png")
image_exists <- file.exists(image)
```


```{r eval = image_exists, fig.cap="_Appended new areas, in a QGIS screenshot. Zoomed into a specific area. Light colour = original data source, dark colour = appended area._", out.width='80%'}
include_graphics(image)
```

The pattern of locally buffered 'obliggwdep' types is quite dominant in the resulting layer.
In contrast, the original layer has numerous narrow and small patches of shallow groundwater.
So these may be too narrow.
It might be sensible to also add a narrow buffer, e.g. 50 m, around the whole of the original layer.
Or, obtaining a broader area by being more liberal in the selection of soil types.
