# Check results

Checksums:

```{r}
c("xxh64", "md5", "sha256") %>% 
  map_dfr(~tibble(algorithm = ., 
                  checksum = checksum(filepath, .))) %>% 
  kable
```


```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath, 
                layer = "habitatmap_terr_polygons"))
```

```{r paged.print=FALSE, warning=FALSE}
(types <- read_sf(filepath, 
                layer = "habitatmap_terr_types"))
```


- Do all records in the habitatmap_terr_types table have a corresponding spatial feature in the habitatmap_terr_polygons layer?

```{r}
check_polygon_id_types <- types %>%
  anti_join(pol, by = "polygon_id")

nrow(check_polygon_id_types) == 0
```

- Do all spatial features in the habitatmap_terr_polygons layer have a corresponding record in the habitatmap_terr_types table?

```{r}
check_polygon_id_polygons <- pol %>%
  st_drop_geometry() %>%
  anti_join(types, by = "polygon_id")

nrow(check_polygon_id_polygons) == 0
```

- Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object? 
  
```{r} 
 types %>%  
   distinct(polygon_id) %>%  
   nrow == nrow(pol) 
``` 

- Is the CRS conform EPSG:31370?

```{r} 
 st_crs(pol) == st_crs(31370) 
```

- Is phab always between 0 and 100?

```{r}
all(between(types$phab, 0, 100))
```

- Is the primary key unique?

```{r}
types %>% 
  select(polygon_id, type, certain) %>% 
  duplicated() %>% 
  negate(any)(.)
```

- Are the geometries valid?

```{r}
sum(!st_is_valid(pol) | is.na(st_is_valid(pol))) == 0
