---
title: "Miscellaneous code chunks"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(sp)
library(sf)
library(raster)
library(tidyverse)
library(stringr)
library(n2khabutils)
library(plotly)
library(rasterVis)
```
This is a html notebook which you can use to add your own miscellaneous code,
e.g. to explore some datasets before writing a function to read it,
to prepare for writing functions, 
to test existing n2khabutils functions, etc.

# Exploring the GRTS master grid

## Correcting a flaw in the original version


```{r}
GRTSmaster_habitats <- 
    raster("../data/10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif")
GRTSmaster_habitats
```

```{r}
inMemory(GRTSmaster_habitats)
```

```{r}
spplot(GRTSmaster_habitats)
```


Oops, see right upper corner. We need to correct this data layer...

First check that the zero value does indeed only occur in the upper right corner:

```{r}
GRTSmaster_habitats[GRTSmaster_habitats == 0,
                    drop = FALSE] %>% 
    spplot
```

Yes, this seems to be a useful criterium.

<!-- ```{r} -->
<!-- flanders <-  st_read("../data/10_raw/flanders") -->
<!-- flanders_500 <-  -->
<!--     flanders %>%  -->
<!--     st_buffer(500) -->
<!-- ``` -->


<!-- ```{r include=FALSE} -->
<!-- # Inspired by code from <https://datacarpentry.org/r-raster-vector-geospatial/08-vector-plot-shapefiles-custom-legend/index.html>, -->
<!-- # however this was too heavy for plotting: -->
<!-- ggplot() + -->
<!--      geom_raster(data = GRTSmaster_habitats %>%  -->
<!--                      as.data.frame(xy = TRUE),  -->
<!--                  aes(x = x,  -->
<!--                      y = y,  -->
<!--                      fill = GRTSmaster_habitats)) + -->
<!--      geom_sf(data = flanders) + -->
<!--      geom_sf(data = flanders_500, -->
<!--              fill = NA, -->
<!--              colour = "red") + -->
<!--     coord_sf() -->
<!-- # compare with: -->
<!-- gplot(GRTSmaster_habitats) + -->
<!--     geom_tile(fill = value) # however, further extension seems difficult here -->
<!-- ``` -->


```{r}
GRTSmaster_mask <- 
    GRTSmaster_habitats %>% 
    reclassify(rcl = c(1, 1e+9, 1,
                       0, 1, NA),
               right = FALSE,
               filename = "../data/20_processed/GRTSmaster_mask",
               overwrite = TRUE)
```

```{r}
GRTSmaster_mask %>% spplot
```


<!-- ```{r} -->
<!-- GRTSmaster_habitats %>% cellStats('countNA') -->
<!-- ``` -->

```{r}
GRTSmaster_simplified <- 
    GRTSmaster_habitats %>% 
    mask(GRTSmaster_mask,
         filename = "../data/20_processed/GRTSmaster_simplified")
```


```{r}
GRTSmaster_simplified %>% spplot
```

How many cells have values, and how many have no value?

```{r}
valueslogical <- 
    !(GRTSmaster_simplified %>% 
    values %>% 
    is.na)
valueslogical %>% 
    table
```

Let's compare the 'TRUE' number with the original data to check completeness:

```{r}
valueslogical_orig <- 
    (GRTSmaster_habitats %>% 
    values) > 0
valueslogical_orig %>% 
    sum(na.rm = TRUE)
```

OK, there is a complete match.

Range of the values:

```{r}
GRTSmaster_simplified %>% 
    values %>%
    summary
```


Let's write the corrected version as a GeoTIFF.

```{r}
GRTSmaster_simplified %>% 
    writeRaster(filename = "../data/20_processed/GRTSmaster_habitats.tif",
                datatype = "INT4S")
```

And now let's check the raster header and its summary, from the newly created GeoTIFF:

```{r}
GRTSmaster_habitats <- 
    raster("../data/20_processed/GRTSmaster_habitats.tif")
GRTSmaster_habitats
```

```{r}
GRTSmaster_habitats %>% 
    spplot
```

## Exploration

### How to make a brick

```{r}
GRTSmaster_habitats <- 
    raster("../data/10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif")
```


```{r}
GRTSmh_brick <- 
    brick("../data/10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif")
```

```{r}
inMemory(GRTSmh_brick)
```


```{r}
GRTSmh_brick2 <- 
    brick(list(GRTSmaster_habitats,
            GRTSmaster_habitats %>% 
                reclassify(rcl = c(1, 1e+9, 1),
                           right = FALSE)
        )
    ) %>% 
    `names<-`(c("decimal", "base4"))
GRTSmh_brick2
```


```{r}
inMemory(GRTSmh_brick2)
```

<!-- ```{r} -->
<!-- writeRaster(GRTSmh_brick2, -->
<!--             filename = file.path(tempdir(),"GRTSmh_brick2.tif"), -->
<!--             datatype = "INT4S", -->
<!--             overwrite = TRUE) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- GRTSmh_brick2 <-  -->
<!--     brick(file.path(tempdir(),"GRTSmh_brick2.tif")) -->
<!-- GRTSmh_brick2 -->
<!-- ``` -->

<!-- ```{r} -->
<!-- names(GRTSmh_brick2)  <-  -->
<!--     c("decimal", "base4") -->
<!-- ``` -->


### Going from base 10 to base 4: does it make sense?

A dedicated function exists, however it takes so long to load the whole package that it seems a better choice to fork the specific function into n2khabutils, as an internal function.

```{r}
oro.dicom::dec2base(37784, base = 4) %>% as.numeric
```

<!-- For the `len`argument we take 14 digits, otherwise the maximum number in the grid will give an error: -->

<!-- ```{r} -->
<!-- log10(8e+7) / log10(4) -->
<!-- ``` -->

The maximum base 4 address in the grid:

```{r}
oro.dicom::dec2base(67108856, base = 4) %>% as.numeric
```

But we better use this function, which is vectorized and does not give a warning:

```{r}
source("../n2khabutils/R/baseConvert.R")
```

```{r}
baseConvert(67108856, target = 4)
```



Let's look at an extracted portion of the GRTSmaster_habitats grid:

```{r}
GRTS_sub <- 
    GRTSmaster_habitats[2000:2050,3000:3050, drop = FALSE]
GRTS_sub %>% 
    spplot
```

```{r}
GRTS_sub_4 <- 
    GRTS_sub %>% 
    calc(function(x){
            as.numeric(baseConvert(x, target = 4))
        })
GRTS_sub_4 %>% 
    spplot
```

Check whether we can do the inverse and return the original data (extracting a few lines of code from [here](https://github.com/graywh/r-gmisc/blob/master/R/baseConvert.R)):

```{r}
basetodec <- function(x, base){
    characters <- c(seq(0,9), LETTERS)
    numbers <- structure(seq(0,35), names = characters)    
    x <- strsplit(as.character(x), "")
    l <- lapply(x, length)
    f1 <- function(x, l) {
        sum(numbers[x] * base ^ (seq(l - 1,0)))
    }
    mapply(f1, x, l)
}
```

```{r}
basetodec(100, 4)
```

```{r}
GRTS_sub_4 %>% 
    calc(function(x) basetodec(x, base = 4)) %>% 
    all.equal(GRTS_sub)
```

Olé.

But, does it make sense spatially; are the original GRTS addresses still reflected in this base4-conversion? Let's extract the 2nd to 6th digit and look at the pattern:

```{r}
max_digits <- 
    GRTS_sub_4 %>% 
    values %>% 
    max %>% 
    as.character %>% 
    str_split("", simplify = TRUE) %>% 
    length
```

```{r}
extract_digit <- 
    function(x, level, max_digits) {
        withr::with_options(
                  c(scipen = 999), 
                  str_pad(x, 13, pad = "0")
                  ) %>% 
            str_sub(level, level) %>% 
            as.integer
    }
```

```{r}
extract_digit(3333333333320, level = 12, max_digits = max_digits)
```


```{r}
for (i in 2:6) {
GRTS_sub_4 %>% 
    calc(function(x) extract_digit(x, level = i, max_digits = max_digits)) %>% 
        spplot %>% 
        print
}
```

Yeeha. Dare we try this on the whole raster?

```{r}
GRTSmaster_habitats %>% 
    calc(function(x){
            ifelse(is.na(x), NA,
                   as.numeric(baseConvert(x, target = 4)) %>% 
            extract_digit(level = 13, max_digits = 13)
            )
        }) %>% 
    spplot
```


What went wrong?

```{r}
GRTSmaster_habitats %>% 
    values %>% 
    enframe %>% 
    filter(!is.na(value)) %>% 
    select(value) %>% 
    slice(1:200) %>% 
    arrange(value) %>% 
    mutate(base4 = baseConvert(value, target = 4),
           base4_num = as.numeric(base4),
           base4_char = as.character(base4_num),
           formatted = withr::with_options(
                          c(scipen = 999), 
                          str_pad(base4_num, 13, pad = "0")
))
```

```{r}
oro.dicom::dec2base(351070, base = 4)
```






















