---
title: "Comparing CRSes of data sources"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
# library(rgdal)
library(gdalUtils)
```


# Inspecting the CRS of several data sources

```{r}
datapath <- "../../data"
data_crs <- tribble(
  ~name, ~path,
  "GRTSmh", file.path(datapath, "10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif"),
  "habitatmap", file.path(datapath, "10_raw/habitatmap/habitatmap.shp"),
  "flanders", file.path(datapath, "10_raw/flanders/flanders.shp"),
  # "soilmap", file.path(datapath, "10_raw/soilmap/soilmap.sqlite"),
  # ERROR - failed to load SRS definition from ../../data/10_raw/soilmap/soilmap.sqlite
  "habmap_stdized", file.path(datapath, "20_processed/habitatmap_stdized/habitatmap_stdized.gpkg"),
  "GRTSmh_brick", file.path(datapath, "20_processed/GRTSmh_brick/GRTSmh_brick.tif"),
  "GRTSmh_base4frac", file.path(datapath, "20_processed/GRTSmh_base4frac/GRTSmh_base4frac.tif"),
  "GRTSmh_diffres7", file.path(datapath, "20_processed/GRTSmh_diffres/GRTSmh_diffres.7.tif"),
  "GRTSmh_diffres_pol", file.path(datapath, "20_processed/GRTSmh_diffres/GRTSmh_diffres.gpkg")
)
data_crs <- 
  data_crs %>% 
  mutate(crs = map(path, function(x) gdalsrsinfo(x, as.CRS = TRUE)))
```


```{r eval=FALSE, echo=FALSE}
datalist <- list(
# raw data sources
  GRTSmh = raster(file.path(datapath, "10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif")),
  habitatmap = st_read(file.path(datapath, "10_raw/habitatmap/habitatmap.shp"),
                                     quiet = TRUE),
  flanders = st_read(file.path(datapath, "10_raw/flanders/flanders.shp"),
                                     quiet = TRUE),
  soilmap = st_read(file.path(datapath, "10_raw/soilmap/soilmap.sqlite"),
                                     quiet = TRUE),
  # processed data sources
  GRTSmh_brick = read_GRTSmh(datapath, brick = TRUE),
  GRTSmh_base4 = read_GRTSmh_base4frac(datapath),
  GRTSmh_diffres7 = read_GRTSmh_diffres(datapath, level = 7),
  GRTSmh_diffres7_pol = read_GRTSmh_diffres(datapath, level = 7, polygon = TRUE),
  habmap_polygons = st_read(file.path(datapath, "20_processed/habitatmap_stdized/habitatmap_stdized.gpkg"),
                                     "habitatmap_polygons",
                                     quiet = TRUE)
)
```

It appears that all these data sources have exactly the same CRS:

```{r}
data_crs$crs
```

We save it in the object `orig_crs`:

```{r}
orig_crs <- data_crs$crs[[1]]
orig_crs
```

# Belge 1972

What are the existing EPSG CRSes of 'Belge 1972'?

```{r}
system("grep 'Lambert 72' /usr/share/proj/epsg -A 1")
```

The proj.4 strings of the first two can also be requested as:

```{r}
epsg_31300 <- gdalsrsinfo("+init=epsg:31300", as.CRS = TRUE)
epsg_31300
```

```{r}
epsg_31370 <- gdalsrsinfo("+init=epsg:31370", as.CRS = TRUE)
epsg_31370
```

EPSG recommends 31370, and that's what we always want.

Importantly, we see a high resemblance between `orig_crs` and EPSG-code 31300, however not exact.

So it appears that the Flemish data sources are not distributed exactly according to EPSG.

Also the ESRI versions of both Lambert 72 CRSes comply with the EPSG proj.4 strings, the only difference being that the EPSG strings also contain parameters for conversion to WGS84.
So it's not clear where `orig_crs` comes from.

```{r}
system("grep 'Lambert 72' /usr/share/proj/esri -A 1")
```


# Consequences?

Now, let's check what are the consequences of this difference by transforming some point coordinates.

```{r}
gdalinfo(file.path(datapath, "10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif")) %>% 
  .[27:32]
```

```{r}
corners <- 
  gdalinfo(file.path(datapath, "10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif"),
         raw_output = FALSE) %>% 
  .$bbox
corners
```


Making an sf object with these 5 points:

```{r}
orig_crs_sf <- st_crs(orig_crs)
orig_points <- 
  st_sfc(
    st_point(c(corners["s1", "min"], corners["s2", "max"])), # Upper Left
    st_point(c(corners["s1", "min"], corners["s2", "min"])), # Lower Left
    st_point(c(corners["s1", "max"], corners["s2", "max"])), # Upper Right
    st_point(c(corners["s1", "max"], corners["s2", "min"])), # Lower Right
    st_point(c(140445.592, 198542.114)),                     # Center
    crs = orig_crs_sf
  ) %>% 
  st_sf
old <- options(digits = 15)
orig_points %>% 
  st_coordinates
```

Transform to epsg:31300 :

```{r}
st_transform(orig_points, 31300) %>% 
  st_coordinates
```

Transform to epsg:31370 :

```{r}
st_transform(orig_points, 31370) %>% 
  st_coordinates
```

This means that there is a large difference with 31300, and a very tiny one (at the sub-millimeter level) with 31370.
Therefore, and especially for rasters (but also polygons) created in `orig_crs`, transformations to 31370 would do more bad than good.

# Solution

We better _set_ the CRS to 31370 without changing the coordinates of these geospatial objects, rather than transform (reproject) the data.

This will be done as follows by the data-preprocessing functions:

- for sf objects:

```{r}
orig_points_31370 <- orig_points
st_crs(orig_points_31370) <- 31370
orig_points_31370
```

- for raster objects:

In the original CRS:

```{r}
GRTSmh  <- raster(file.path(datapath, "10_raw/GRTSmaster_habitats/GRTSmaster_habitats.tif"))
GRTSmh
```

```{r}
GRTSmh %>% 
  xyFromCell(1:10)
```

With the CRS set to 31370:

```{r}
GRTSmh_31370 <- GRTSmh
crs(GRTSmh_31370) <- epsg_31370
GRTSmh_31370
```

```{r}
GRTSmh_31370 %>% 
  xyFromCell(1:10)
```

So the coordinates were not altered.



