---
title: "Exploring Watina"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
# library(sf)
# library(raster)
library(tidyverse)
library(stringr)
# library(n2khabutils)
# library(plotly)
# library(rasterVis)
# library(stars)
library(DBI)
source("../private_code.R") # just loads connection string for Watina DB
library(dbplyr)
library(dbplot)
```

# Querying the Watina groundwater database and exploring tables

The below explorations are based on code of Ivy Jansen, but use the `DBI` library instead of `RODBC` (`DBI` is standard in RStudio).


```{r}
con <- dbConnect(odbc::odbc(), .connection_string = connectionstring_watina)
dbListTables(con) %>% View("tables")
```

## Locations and filters

Listing relevant tables and showing first records:

```{r}
locations <-  tbl(con, "DimMeetpunt")
locations %>% glimpse
```

Variables of interest: `MeetpuntWID`, `MeetpuntCode`, `MeetpuntStatusCode`, `MeetpuntTypeCode`, `MeetpuntXCoordinaat`, `MeetpuntYCoordinaat`.

```{r}
# dbplot library provides a few basic graph functions for which the 
# calculations are executed inside the database, and a ggplot object is returned.
locations %>% 
    dbplot_raster(MeetpuntXCoordinaat, 
                  MeetpuntYCoordinaat, resolution = 300) +
    coord_fixed()
```


```{r}
# does not work:
dbGetQuery(con, 'SELECT TOP(6) * FROM DimMeetpunt')
```


```{r}
# does not work:
tbl(con, "DimPeilpunt") %>% 
    head
```

```{r}
dbListFields(con, "DimPeilpunt")
```

```{r}
# DOES work:
gwfilters <- 
    dbGetQuery(con, "SELECT MeetpuntWID, PeilpuntCode, ReferentieNiveauMaaiveld, PeilbuisLengte, FilterLengte,
                    PeilpuntOpenbaarheidTypeCode, PeilpuntOpenbaarheidTypeNaam, PeilpuntTypeNaam
                 FROM DimPeilpunt")
gwfilters %>% glimpse
```

The table `DimPeilpunt` will be needed to select the shallow groundwater filters (i.e. with filter base < 3 m below soil surface), and to select the shallowest filter from (remaining) filters duplets.

Let's explore that a little more:

```{r}
gwfilters %>% 
    mutate(filterbase = -(PeilbuisLengte - ReferentieNiveauMaaiveld)) %>% 
    ggplot(aes(x = filterbase)) + 
    geom_histogram(colour = "grey90") +
    scale_y_log10() +
    xlab("filterbase (m below soil surface)")
```

```{r}
gwfilters %>% 
    mutate(filterbase = -(PeilbuisLengte - ReferentieNiveauMaaiveld),
           shallow = filterbase > -3) %>% 
    filter(filterbase > -10) %>% 
    ggplot(aes(x = filterbase, fill = shallow)) +
    geom_histogram(binwidth = 0.5, colour = "grey90") + 
    # scale_y_log10() +
    xlab("filterbase (m below soil surface)")
```

```{r}
gwfilters %>% 
    mutate(filterbase = -(PeilbuisLengte - ReferentieNiveauMaaiveld)) %>% 
    filter(filterbase == -3) %>% 
    nrow
```

Just including the filterdepths of exactly 3 meters deep would however add 433  filters, so let's include that one as well (it probably has to do with rounding effects):

```{r}
gwfilters %>% 
    mutate(filterbase = -(PeilbuisLengte - ReferentieNiveauMaaiveld),
           shallow = filterbase >= -3) %>% 
    filter(filterbase > -10) %>% 
    ggplot(aes(x = filterbase, fill = shallow)) +
    geom_histogram(binwidth = 0.5, colour = "grey90") + 
    # scale_y_log10() +
    xlab("filterbase (m below soil surface)")
```

## Groundwater levels: XG3

```{r}
# "SELECT * FROM ssrs_Precalc"
# the following works:
XG3 <- tbl(con, "ssrs_Precalc")
XG3 %>% glimpse()
```

Variables of interest for MNE: `HydroJaar`, `MeetpuntWID`, `GHG_2`, `GLG_2`, `GVG_2`, `MetingTypeWID`.
Optionally (with a switch?) the `_1` versions of the XG3 variables.
`_1`-variables are the absolute groundwater levels, `_2`-variables are the groundwater levels relative to soil surface.

Nr of rows in this table:

```{r}
XG3 %>% collect %>% nrow
```

Positive values will need to be set to zero:

```{r}
XG3 %>% 
    filter(GHG_2 > -2,
           GHG_2 < 1.5) %>% 
    dbplot_histogram(GHG_2)
```


```{r}
tbl(con, "DimMetingType")
```

Will this table be of use?
May be.
We could replace `MetingTypeWID` in the previous table by a factor `measurement_type`, using the codes from `MetingTypeWID` and translating these (within a function that has a `lang` argument) to a label variable with either Dutch or English (or ...) names by storing those in the `namelist` data source of the R package.


## Chemical data

```{r}
# "SELECT * FROM DimChemVar"
# the following works:
chemvars <- tbl(con, "DimChemVar")
chemvars %>% glimpse()
```

This table provides the chemical variable definitions.

```{r}
# "SELECT * FROM FactChemischeMeting"
# the following works:
chem_measurements <- tbl(con, "FactChemischeMeting")
chem_measurements %>% glimpse
```

This table provides the chemical measurements.
Essential variables are: `MeetpuntWID`, `DatumWID`, `ChemVarWID`, `Meetwaarde`, `MeetwaardeMEQ`, `IsBelowLOQ`, `MetingStatusCode`, `MetingStatus`.

Missing here: electroneutrality (but see Cécile's [remark](https://docs.google.com/a/inbo.be/document/d/1isa_9HYQcxo8OYi0fKYmLY42GGTNxpev6RgW05WNxE4/edit?disco=AAAACuSS6FU)).

```{r}
dbDisconnect(con)
```


