---
title: "Exploring Watina chemistry (extract from complete notebook)"
date: '`r paste("Version",lubridate::now())`'
output:
  html_document:
    df_print: paged
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(stringr)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


```{r}
con <- inbodb::connect_inbo_dbase("W0002_00_Watina")
```

```{r}
chemvars <- tbl(con, "DimChemVar")
chemvars %>% glimpse()
```

```{r}
chem_measurements <- tbl(con, "FactChemischeMeting")
chem_measurements %>% glimpse
```

Essential variables are: `MeetpuntWID`, `DatumWID`, `ChemVarWID`, `Meetwaarde`, `MeetwaardeMEQ`, `IsBelowLOQ`.

# IsBelowLOQ

Which values occur for `IsBelowLOQ`?

```{r}
chem_measurements %>% 
  group_by(IsBelowLOQ) %>% 
  summarise(n = n(),
            mindate = min(DatumWID, na.rm = TRUE),
            maxdate = max(DatumWID, na.rm = TRUE))
```

The interpretation here is that:

- only for LIMS data this is (automatically) imported, i.e. for data since 2014.
This concerns the `TRUE` & `FALSE` values.
- The `NA` values after 2014 are from external sources.

No `NA` values occur for `Meetwaarde` when `IsBelowLOQ == "TRUE"`:

```{r}
chem_measurements %>% 
  filter(IsBelowLOQ == "TRUE") %>% 
  select(Meetwaarde) %>% 
  collect %>% 
  summary
```

This is because LOQ values were used and divided by 2, in those cases:

```{r}
chem_measurements %>% 
  filter(IsBelowLOQ == "TRUE") %>% 
  inner_join(chemvars,
             by = "ChemVarWID") %>% 
  select(ChemVarCode, Meetwaarde) %>% 
  collect %>% 
  split(x = ., f = .$ChemVarCode) %>% 
  purrr::map(.x = ., .f = table)
```

Furthermore, quite a lot of `NA` values are present for `Meetwaarde`:

```{r}
chem_measurements %>% 
  filter(is.na(Meetwaarde)) %>% 
  inner_join(chemvars %>% select(ChemVarWID, ChemVarNaam, ChemVarEenheid),
             by = "ChemVarWID") %>% 
  group_by(IsBelowLOQ, ChemVarNaam) %>% 
  count %>% 
  arrange(desc(n)) %>% 
  kable
```

So we should take into account that `NA` values do occur, when calculating summary statistics.
Update: `NA` values are already removed by `watina::get_chem()`.

# mass vs. meq data

For which variables are these values equal?

```{r}
chem_measurements %>% 
  filter(Meetwaarde == MeetwaardeMEQ,
         IsBelowLOQ == "FALSE") %>% 
  count(ChemVarWID) %>% 
  inner_join(chemvars %>% select(ChemVarWID, ChemVarNaam, ChemVarEenheid),
             by = "ChemVarWID")
```

```{r}
chem_measurements %>% 
  filter(IsBelowLOQ == "FALSE") %>% 
  inner_join(chemvars %>% select(ChemVarWID, ChemVarNaam, ChemVarEenheid),
             by = "ChemVarWID") %>% 
  filter(ChemVarNaam %in% c("Al", "Mn", "Si")) %>% 
  count
```

So _all_ data of Al, Mn and Si show no difference between both columns.
It appears they wrongly have a `Coefficient` = 1 in `chemvars`.

Checking calculations where `Coefficient` > 1, this seems OK:

```{r paged.print=FALSE}
chem_measurements %>% 
  filter(IsBelowLOQ == "FALSE") %>% 
  inner_join(chemvars %>% 
               select(ChemVarWID, ChemVarNaam, ChemVarEenheid, Coefficient) %>% 
               filter(Coefficient > 1),
             by = "ChemVarWID") %>% 
  mutate(test = Meetwaarde / Coefficient) %>% 
  select(ChemVarNaam, Meetwaarde, MeetwaardeMEQ, test, Coefficient)
```

```{r}
chem_measurements %>% 
  filter(IsBelowLOQ == "FALSE") %>% 
  inner_join(chemvars %>% 
               select(ChemVarWID, ChemVarNaam, ChemVarEenheid, Coefficient) %>% 
               filter(Coefficient > 1),
             by = "ChemVarWID") %>% 
  mutate(test = Meetwaarde / Coefficient) %>% 
  select(ChemVarNaam, Meetwaarde, MeetwaardeMEQ, test, Coefficient) %>% 
  filter(abs(test - MeetwaardeMEQ) >= 0.1)
```

Number of observations per chemical variable:

```{r}
chem_measurements %>% 
  inner_join(chemvars %>% select(ChemVarWID, ChemVarNaam, ChemVarEenheid),
             by = "ChemVarWID") %>% 
  count(ChemVarNaam)
```


```{r}
DBI::dbDisconnect(con)
```




