---
title: "Handling the habitatmap"
date: '`r paste("Version",lubridate::now())`'
output:
  html_document:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(sf)
library(dplyr)
library(stringr)
library(tidyr)
library(n2khab)
library(knitr)
library(tmap)
library(leaflet)
library(ggplot2)
library(terra)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# A few checks of the habitatmap data source

We check the interim version of April 2026: **habitatmap_2025_v99_interim**, based on field work until 2025.

Because it was found that sf has difficulties with correcting certain (invalid) MULTISURFACE geometries, terra is being used to read the file and the result is converted to sf: terra automatically casts all geometries to polygons on reading; after conversion to sf we get no missing reasons for invalidity and `st_make_valid()` works for all invalid geometries.

We do keep the geometry-type attribute of the data as read directly by sf, to further check selected MULTISURFACE geometries.

```{r read-habitatmap}
filepath <- file.path(locate_n2khab_data(),
                      "10_raw/habitatmap")

if (length(grep("gpkg", list.files(path = filepath))) > 0) {
  # format = geopackage
  filepath <- file.path(filepath, "habitatmap.gpkg")
}
# all formats
habitatmap_sf <- 
  read_sf(filepath) %>% 
  mutate(geomtype_sf = st_geometry_type(.))
habitatmap <- 
  vect(filepath) %>% 
  st_as_sf() %>%
  as_tibble() %>%
  bind_cols(habitatmap_sf %>% st_drop_geometry() %>% select(geomtype_sf)) %>% 
  st_as_sf() %>% 
  # since we get a combination of POLYGON and MULTIPOLYGON, we entirely cast
  # as MULTIPOLYGON
  st_cast("MULTIPOLYGON")
```

The number of rows is `r nrow(habitatmap)`.

Checksums:
```{r checksums}
filepath %>% 
  n2khab::md5sum() %>% 
  `names<-`("md5sum")
filepath %>% 
  n2khab::sha256sum() %>% 
  `names<-`("sha256sum")
filepath %>% 
  n2khab::xxh64sum() %>% 
  `names<-`("xxh64sum")
```

Compare the geometry types between `habitatmap_sf` and `habitatmap`:

```{r paged.print = FALSE}
habitatmap_sf %>% 
  st_drop_geometry() %>% 
  count(geomtype_sf)
habitatmap %>% 
  mutate(geomtype = st_geometry_type(.)) %>% 
  st_drop_geometry() %>% 
  count(geomtype)
```

Optionally checking that the geometries of type MULTIPOLYGON in `habitatmap_sf` have the exact same geometry in `habitatmap`.

```{r eval = FALSE}
# the comparison at the end takes a lot of time; not running this by default
habitatmap_sf %>%
  select(globalid_BWK) %>%
  mutate(id = row_number()) %>% 
  filter(geomtype_sf == "MULTIPOLYGON") %>% 
  inner_join(
    habitatmap %>%
      # drop sf class but not the geometry column:
      as_tibble() %>% 
      select(globalid_BWK) %>%
      mutate(id = row_number()),
    # it has been checked separately that row order is the same between both
    # objects
    join_by(id)
  ) %>% 
  as_tibble() %>% 
  mutate(equalgeom = geom == geometry) %>% 
  select(id, equalgeom) %>% 
  count(equalgeom)
# # A tibble: 1 × 2
# equalgeom      n
# <lgl>      <int>
#   1 TRUE      696990
```


## Converting column names

```{r column-names}
names(habitatmap)
```

Some names are in uppercase, others in lowercase or mixed case.
This is different from the regular version (where all the column names are in uppercase).
A few columns from the regular version are missing in the interim version: `OIDN`, `UIDN`, `BWKLABEL`, `HABLEGENDE`, `LENGTE`, `OPPERVL`.
Other columns are missing in the regular version: `referentie`, `info`, `globalid_BWK`.
And the following names are different between regular and interim version: 

- `HERK` => `herk_bwk`
- `HERKHAB` => `HERK_HAB`
- `HERKPHAB` => `HERK_pHAB`
- `SHAPE_Leng` => `SHAPE_Length` (probably due to the fact that the previous version was a shapefile)

We convert all the column names to lowercase and remove most underscores, in order to make the code less dependent on the habitatmap version.

```{r lowercase-no-underscore}
names(habitatmap) <- str_to_lower(
  str_replace_all(names(habitatmap), "_(?!sf)", "")
)
```



## A summary of the layer

```{r summary-habitatmap}
habitatmap %>% 
  st_drop_geometry() %>% 
  summary()
```

Let us look for possible errors.

## Are there `NA` values?

There are no `NA`s.

```{r na-values}
sapply(habitatmap, function(x) sum(is.na(x)))
```

## Are there `<Null>` values?

There is one `<Null>` string.

```{r null-values}
sapply(
  habitatmap |> st_drop_geometry(),
  function(x) sum(as.character(x) == "<Null>", na.rm = TRUE)
)
```

```{r}
habitatmap %>% 
  st_drop_geometry() %>% 
  filter(info == "<Null>")
```

## Are there Zero (0) values?

Yes,  in `phab2`, `phab3`, `phab4`, `phab5` (as expected), but also in `herkphab`.

```{r zero-values}
sapply(
  habitatmap |> st_drop_geometry(),
  function(x) sum(as.character(x) == '0', na.rm = TRUE)
)
```

```{r}
habitatmap %>% 
  st_drop_geometry() %>% 
  filter(herkphab == '0')
```

## Are there empty strings instead of NA?

There are many empty strings (apparently used instead of NAs), in most of the cases in the columns where empty values can be expected though.

```{r empty-strings}
sapply(
  habitatmap |> st_drop_geometry(),
  function(x) sum(as.character(x) == '', na.rm = TRUE)
)
```

The following records might need to be checked since we don't expect empty values in the columns `eval`, `eenh1`, `herkbwk`, `hab1`, `herkhab`, `herkphab`.

```{r}
habitatmap |> st_drop_geometry() %>% 
  select(
    objectid,
    tag,
    eval,
    eenh1,
    eenh2,
    herkbwk,
    hab1,
    phab1,
    herkhab,
    herkphab,
    globalidbwk
  ) %>% 
  filter(if_any(c(eval, eenh1, herkbwk, hab1, herkhab, herkphab), \(x) x == ""))
```

<!-- As for `tag`: there are plenty of empty records in this interim versions. -->
<!-- The `tag` will be added for the publication. In the meantime we might use -->
<!-- `globalidbwk`. -->

## Are there spaces instead of NA?

There are records with one or several spaces.

```{r spaces}
sapply(
  habitatmap |> st_drop_geometry(),
  function(x) sum(grepl(pattern = "^\\s+$", x), na.rm = TRUE)
)
```

All rows have a space as the value for `tag`.

As for `eenh2` until `eenh8`, ...: there are too many of them to print the list. 
Here are the cases with spaces for the other variables: 

```{r}
habitatmap |> st_drop_geometry() %>% 
  select(
    objectid,
    tag,
    eval,
    eenh1,
    eenh2,
    herkbwk,
    hab1,
    phab1,
    herkhab,
    herkphab,
    globalidbwk
  ) %>% 
  filter(if_any(
    c(eval, eenh1, herkbwk, hab1, herkhab, herkphab),
    \(x) grepl(pattern = "^\\s+$", x) & nchar(x) > 0
  ))
```


## Are polygon codes unique?

Even if we remove the records with `tag` = 0, there are duplicates.

```{r}
habitatmap$tag %>% unique() %>% length() == nrow(habitatmap)
habitatmap$tag[habitatmap$tag != '0'] %>% unique %>% length == (nrow(habitatmap) - 1)
```

How many repeated TAGs?

```{r}
duplic_tags <- 
  habitatmap %>% 
  st_drop_geometry() %>% 
  count(tag) %>% 
  filter(n > 1)
nrow(duplic_tags)
```

How often do repeated TAGs occur?

```{r}
duplic_tags
```

And is globalidbwk unique?

```{r}
habitatmap$globalidbwk %>% unique() %>% length() == nrow(habitatmap)
```

How many repeated values of `globalidbwk`?

```{r}
duplic_globalid <- 
  habitatmap %>% 
  st_drop_geometry() %>% 
  count(globalidbwk) %>% 
  filter(n > 1)
nrow(duplic_globalid)
```

How often do repeated values of `globalidbwk` occur?

```{r}
quantile(duplic_globalid$n, seq(0, 1, 0.1))
```

### Selecting or constructing a unique code

From here on we will use as identifier: `globalidbwk` with a suffix that makes it unique where needed (in a new `id` column that 
can be used with any version of the habitatmap). 

```{r add-id}
habitatmap <- habitatmap %>% 
  mutate(
    id = str_c(
      globalidbwk,
      ifelse(
        globalidbwk %in% duplic_globalid$globalidbwk,
        str_c("_", str_pad(row_number(), 2, pad = "0")),
        ""
      )
    ),
    .by = globalidbwk
  )
```

## Are the habitat codes correct?

```{r check-types}
# read standard types from n2khab
list_habs <- n2khab::read_types()

# long version with all the habitats in one column
habitatmap_lg <- habitatmap %>% 
  st_drop_geometry() %>% 
  select(id, starts_with("hab") | starts_with("phab"), -any_of(c("HABLEGENDE", "hablegende"))) %>%
  pivot_longer(cols = starts_with("hab"), 
               names_to = "habnr", 
               names_prefix = "hab", 
               values_to = "hab") %>%
  pivot_longer(cols = starts_with("phab"), 
               names_to = "phabnr", 
               names_prefix = "phab", 
               values_to = "phab") %>% 
  filter(habnr == phabnr) %>% 
  select(-phabnr)

# habitats in habitatmap but not in official list n2khab:
habitatmap_lg %>% 
  filter(!is.na(hab) & 
           !hab %in% c("gh", "", '<Null>') & 
           !grepl(pattern = "^\\s*$", hab)) %>%  
  mutate(hab = gsub(hab, pattern = ",gh", replacement = ""),
         hab = gsub(hab, pattern = ",bos", replacement = "")) %>% 
  anti_join(list_habs, by = c("hab" = "type")) %>% 
  count(hab) 
```

Several codes exist with superfluous space at the end, which needs solving.

"x" is used for unknown (for instance private domain).

## Is the sum of the phab 100%?

There should not be any row returned with sum phab < 100% or > 100% 
(except when habitat 1130 is present):

```{r sum-phab-100}
habitatmap %>% 
  st_drop_geometry() %>% 
  mutate(phab_tot = phab1 + phab2 + phab3 + phab4 + phab5) %>% 
  filter(phab_tot < 100 |
           (phab_tot > 100 & hab1 != "1130")) %>% 
  nrow()
```


## Are there habitats with phab = 0%?

Yes: many!
Mostly used for habitats only present on a small surface, for instance as small 
landscape elements.

```{r phab0}
habitatmap_lg %>% 
  filter(
    !(
      is.na(hab) |
        hab %in% c("gh", "", '<Null>') |
        grepl(pattern = "^\\s*$", hab) 
    )
    & 
      phab == 0 ) %>% 
  nrow()
```


## Are HERKHAB codes OK?

The included years look OK.

```{r check-herkhab-1, paged.print = FALSE}
habitatmap %>%
  st_drop_geometry() %>%
  mutate(
    year_hab = ifelse(
      str_detect(herkhab, "^\\d{2}"),
      str_sub(herkhab, 1, 2),
      NA_character_
    ) %>%
      as.integer()
  ) %>%
  count(year_hab) %>%
  print(n = Inf)
```

The included methods look OK.

```{r check-herkhab-2, paged.print = FALSE}
habitatmap %>%
  st_drop_geometry() %>%
  mutate(method = str_extract(
    herkhab,
    # extract the methods, not the season-letters [vznw] that may be added to
    # fieldwork years:
    "^\\d*?([a-mo-ux-yA-MO-UX-Y]{1,2})",
    group = 1
  )) %>%
  count(method)
```


## Is the CRS correct?

Checking the CRS stated in the file as "`r st_crs(habitatmap)$input`": does it actually conform to the EPSG standard (in order to prevent CRS clashes in workflows)?

Does it conform to EPSG:31370?

Yes


```{r check-crs}
st_crs(habitatmap) == st_crs(31370)
st_crs(habitatmap)
```


## Validity of the geometries

### Listing invalid geometries

First we check that no empty geometries are present:

```{r check-empty-geom}
all(!is.na(st_dimension(habitatmap$geometry)))
```

Let's inspect features with invalid or corrupt geometry:

```{r check-geometry}
start_time <- Sys.time()
#st_is_valid(habitatmap) %>% table
validities <- st_is_valid(habitatmap)
# invalid but not NA
invalid_geoms <- habitatmap[!is.na(validities) & !validities, ] %>% 
  mutate(geom_valid = "invalid")
# NA
navalid_geoms <- habitatmap[is.na(validities), ] %>% 
  mutate(geom_valid = "na")
# invalid_geoms <- habitatmap[!validities | is.na(validities), ] # returns all invalid and NA
end_time <- Sys.time()
end_time - start_time
```


```{r count-invalid-geoms}
st_is_valid(rbind(invalid_geoms, navalid_geoms), reason = TRUE) %>%
  as_tibble() %>%
  filter(is.na(value) | value != "Valid Geometry") %>% 
  mutate(problem = str_extract(string = value, 
                                pattern = "[^\\[]+")) %>% 
  bind_cols(
    rbind(invalid_geoms, navalid_geoms) %>%
      st_drop_geometry() %>%
      select(geomtype_sf)
  ) %>% 
  count(geomtype_sf, problem) 
```

The geometry invalidities comprise [self-touching rings](https://gdal.org/en/latest/user/geometry_validity.html#polygon-with-self-touching-ring) and [nested multipolygons](https://gdal.org/en/latest/user/geometry_validity.html#nested-multipolygons), as a consequence of digitization errors.
The nested multipolygons are confined to the geometries that were seen as MULTISURFACE by sf.

There are no polygons that return `NA`.

```{r}
nrow(navalid_geoms)
```

### Inspecting invalid geometries

First, let us take a look at the **invalid polygons**:

```{r graph-invalid-geoms, eval = FALSE}
# only use this if there are not too many invalid polygons
tm_shape(invalid_geoms) + tm_borders() + tm_facets(by = "id")
```

```{r map-invalid-geoms-ring}
# there are many invalid polygons, so we show them with leaflet:
invalid_geoms_wgs84 <- invalid_geoms %>% 
  st_transform(crs = 4326) 

leaflet(height = "600px", width = "700px") %>%
  inbospatial::add_wms_be_cartoweb("topo_grey") %>%
  addPolygons(data = invalid_geoms_wgs84,
              color = "darkred",
              popup = paste("id:", 
                          invalid_geoms_wgs84$id)) 
```

### Fixing invalid geometries and inspecting the result

We can fix the invalid geoms easily with `st_make_valid`.

Let's compare with the same geoms after fixing the rings:

```{r make-valid-ring-geom}
valid_geoms <- st_make_valid(invalid_geoms)
```

```{r graph-corrected-geoms, eval = FALSE}
# only use this if there are not too many invalid polygons
tm_shape(valid_geoms) + tm_borders() + tm_facets(by = "id")
```

```{r map-corrected-geoms}
# there were many invalid polygons, so we show them with leaflet:
valid_geoms_wgs84 <- valid_geoms %>% 
  st_transform(crs = 4326) 

leaflet(height = "600px", width = "700px") %>%
  inbospatial::add_wms_be_cartoweb("topo_grey") %>% 
  addPolygons(data = invalid_geoms_wgs84, 
              group = "before correction (red)",
              color = "darkred") %>% 
  addPolygons(data = valid_geoms_wgs84,
              group = "after correction (blue)",
              popup = paste("id:", valid_geoms_wgs84$id)) %>%
  addLayersControl(overlayGroups = c("before correction (red)", "after correction (blue)"),
                   options = layersControlOptions(collapsed = FALSE))
```

Are all geometries valid now?

```{r recheck-geometry}
all(st_is_valid(valid_geoms))
```

We specifically compare geometries seen as MULTISURFACE geometries by sf, before and after correction.

Before:

```{r}
tm_shape(invalid_geoms %>% filter(geomtype_sf == "MULTISURFACE")) +
  tm_polygons() +
  tm_facets(by = "id")
```

After:

```{r}
tm_shape(valid_geoms %>% filter(geomtype_sf == "MULTISURFACE")) +
  tm_polygons() +
  tm_facets(by = "id")
```

We see that the nested multipolygons have been unioned, which looks fine.

Note: it appeared that directly plotting (valid or invalid) MULTISURFACE geometries in `habitatmap_sf` is not supported by ggplot, tmap or leaflet.

```{r eval=FALSE, include=FALSE}
# invalid geometries
invalid_geoms_sf_multisurface <- 
  habitatmap_sf[!is.na(validities) & !validities, ] %>%
  filter(geomtype_sf == "MULTISURFACE")
ggplot(invalid_geoms_sf_multisurface) +
  geom_sf() +
  facet_wrap(~globalid_BWK)
# sampled valid geometries
valid_geoms_sf_multisurface <- 
  habitatmap_sf[!is.na(validities) & validities, ] %>%
  filter(geomtype_sf == "MULTISURFACE")
tm_shape(valid_geoms_sf_multisurface %>% slice_sample(n = 15)) +
  tm_polygons() +
  tm_facets(by = "globalid_BWK")
ggplot(valid_geoms_sf_multisurface %>% slice_sample(n = 15)) +
  geom_sf() +
  facet_wrap(~globalid_BWK)
```

So the geometry correction works well.
Derived data (habitatmap_stdized etc) will contain the fixed geometries.

For the official version of the habitatmap we also have an optional geometry 
reparation step in `read_habitatmap(fix_geom = TRUE)`.

How long would it take to repair the geometries 'on the fly' while importing the habitatmap? 

```{r fix-geometry, eval = FALSE}
# official version
start_time <- Sys.time()
invisible(st_make_valid(habitatmap))
end_time <- Sys.time()
end_time - start_time
```

Refer to <https://github.com/inbo/n2khab-preprocessing/issues/60>, <https://r-spatial.org/r/2017/03/19/invalid.html> and <https://gdal.org/en/latest/user/geometry_validity.html> for more information!

### Dropping geomtype_sf

We had included geomtype_sf in order to explore the MULTISURFACE geometries, but it is superfluous now.

```{r}
habitatmap <- habitatmap %>% select(-geomtype_sf)
```


# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
habitatmap %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`TAG`                         `polygon_id` (official version)
`globalid_BWK`                `polygon_id` (interim version)
`EVAL`                        `eval`
`EENH1`                       `eenh1`
`V1`                          `v1`
`HERK`                        `source`
`INFO`                        `info `
`BWKLABEL`                    `bwk_label`
`HAB1`                        `hab1`
`PHAB1`                       `phab1`
`HERKHAB`                     `source_hab`
`HERKPHAB`                    `source_phab`
`OPPERVL`                     `area_m2`

# Other considerations for the R object returned by `read_habitatmap()`

- not uptaking `OIDN`, `UIDN`, `HABLEGENDE`, `LENGTE`

# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
if ("sf" %in% si$packages$package) {
  p <- c(p, sf_extSoftVersion())
  names(p)[names(p) == "proj.4"] <- "PROJ"
}
if ("rgrass" %in% si$packages$package) {
  p <- c(p, GRASS = link2GI::findGRASS()[1, "version"])
}
sprintf("- **%s**: %s\n", names(p), p) %>%
  cat(sep = "")
```

```{r results = "asis", echo=FALSE}
si$packages %>%
    as_tibble %>%
    select(package, loadedversion, date, source) %>%
pander::pandoc.table(caption = "Loaded R packages",
                     split.table = Inf)
```
