---
title: "Handling the map for habitat 3260 (habitatstreams)"
date: '`r paste("Version",lubridate::now())`'
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
options(stringsAsFactors = FALSE)
library(sf)
library(tidyverse)
library(n2khab)

# wfs:
library(ows4R)
library(httr)

library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# A few checks of the habitatstreams data source 

We want to add a new version of habitatstreams (1.7) (10.5281/zenodo.4420858)

## Compare version 1.6 and 1.7

(outside R)

- Both are shapefiles
- same prj
- same fieldnames and fieldtypes

CONCLUSION: we can use the existing read_habitatstreams as a quick tool to explore the new version and compare it with version 1.6

## Data exploration

```{r data-import}

# with version 1.7 in n2khab_data\10_raw\habitatstreams
# load new version 1.7
habitatstreams17 <- read_habitatstreams(source_text = TRUE)

# manually change the current version in n2khab_data\10_raw\habitatstreams 
# (because there is no management system for the versions yet)
# load old version 1.6 (10.5281/zenodo.3386246)
# habitatstreams16 <- read_habitatstreams()

```

**A quick look**

```{r quick-look}
habitatstreams17$lines %>%
  st_drop_geometry %>% 
  summary
# there are some NA in the river_name (n= 26/561), but no NA in source_id and type
# type is always 3260 as expected

# habitatstreams16 %>% 
#   st_drop_geometry %>% 
#   summary # in version 1.6 there were 25 NA's in river_name 
```
There are some NA's for river_name: 26 in version 1.7 versus 25 in version 1.6. This difference seems consistent.

**Look for inconsistencies in source_id**

```{r data-exploration}
# we check the source_ids
habitatstreams17$lines %>%
  count(source_id) # seems ok, no NA, names seems fine
```

**Let's plot the streams as a map**

```{r plot}

# plot habitatstreams 1.7
 p <- ggplot() +
  geom_sf(data = habitatstreams17$lines, aes(), color = "blue")

# wfs Flanders
wfs_vrbg <- "https://geoservices.informatievlaanderen.be/overdrachtdiensten/VRBG/wfs"

sf_vl <- wfs_vrbg %>% 
  parse_url() %>% 
  list_merge(query = list(service = "wfs",
                          #version = "1.1.0", # optional
                          request = "GetFeature",
                          typeName = "VRBG:Refgew",
                          srsName = "EPSG:31370")) %>% 
  build_url() %>% 
  read_sf(crs = 31370)%>% 
  st_cast(to = "GEOMETRYCOLLECTION")

p <- p + 
  geom_sf(data = sf_vl, fill = NA)

print(p)

```

**Are all the streams in Flanders?**

```{r streams-in-fl}
in_vl <- st_contains(x = sf_vl, y = habitatstreams17$lines)# patience

not_contained <- habitatstreams17$lines %>% 
  st_drop_geometry() %>% 
  mutate(orig_rowname = rownames(.)) %>% 
  filter(!rownames(.) %in% in_vl[[1]])

# 32 streams (segments)
print(not_contained %>% nrow())

not_contained %>% 
  distinct(river_name) %>% # suggestion for An: homogeneous lettercase?
arrange(river_name)
```
32 stream sections are not 'contained' within the polygon for Flanders (see list above)
These 7 rivers are close to a boundary with another region, it seems plausible.
