---
title: "Handling the habitatmap"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# A few checks of the habitatmap_stdized data source

```{r}
result <- read_habitatmap_stdized("../../data")
```


Do all polygons from the sf object have a unique ID? (Beware, this calculation takes approx. 2-3 minutes; it has been disabled)

```{r eval=FALSE}
result$habitatmap_polygons %>% 
  count(polygon_id) %>% 
  filter(n > 1) %>% 
  nrow == 0
```

Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object?

```{r}
result$habitatmap_patches %>% 
  distinct(polygon_id) %>% 
  nrow == nrow(result$habitatmap_polygons)
```

Do all polygon IDs from the dataframe coincide with those of the sf object?

```{r message=FALSE}
result$habitatmap_patches %>% 
  inner_join(result$habitatmap_polygons) %>% 
  nrow == nrow(result$habitatmap_patches)
```

Number of polygons with phab = 0:

```{r}
result$habitatmap_patches %>% 
  filter(phab == 0) %>% 
  nrow()
```

