---
title: "Is clipping `watersurfaces` out of `habitatmap_stdized` a sensible step in generating a derived data source `habitatmap_terr`?"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(stringr)
library(sf)
library(sp)
library(rgrass7)
library(knitr)
library(tmap)
library(n2khab)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# Reading basic datasets

Components of `habitatmap_stdized`:

```{r}
habmap_std <- read_habitatmap_stdized()
habmap_std_pol <- habmap_std$habitatmap_polygons
habmap_std_patch <- habmap_std$habitatmap_patches
```

Components of `watersurfaces`:

```{r warning=FALSE}
watsurf_pol <- 
    st_read("../../n2khab_data/10_raw/watersurfaces/",
            crs = 31370,
            quiet = TRUE)
```


```{r}
glimpse(watsurf_pol)
```

Let's just keep geometries:

```{r}
watsurf_pol <- 
  watsurf_pol %>% 
  st_geometry
```

```{r}
watsurf_pol
```

**The aim here is to remove all areas from `habitatmap_stdized` which are covered by watersurfaces, and see if that actually makes sense (inspect what is being removed).**

# Preliminary experiments with sf, sp, rgeos, RQGIS

## Trials to do the 'difference' operation with the sf package

Did't work out due to topological faults in the data sources!

See commit 0f881bb for the details.


## Trials to use sp and rgeos

Same topological errors - not surprising as GEOS is used in all cases.

See commit d6029af for the details.


## Experimenting with the QGIS bridge

On a system with GDAL and GRASS installed, QGIS offers at least:

- access to an OGR algorithm `clipvectorsbypolygon`, which however resulted in an error.

```{r eval=FALSE}
library(RQGIS)
find_algorithms()
find_algorithms("clip", name_only = TRUE)
open_help("gdalogr:clipvectorsbypolygon")
get_usage("gdalogr:clipvectorsbypolygon")
# get_options("gdalogr:clipvectorsbypolygon")
```

```{r eval=FALSE}
habmap_integr_pol <-
    run_qgis("gdalogr:clipvectorsbypolygon",
             INPUT = habmap_std_pol,
             INPUT2 = watsurf_pol,
                 OUTPUT = file.path(tempdir(), "habmap_integr_pol.shp"),
                 load_output = TRUE)
```

- the GRASS algorithm `v.overlay`; however for some reason the transportation of the data is not well done, so error again:

```{r eval=FALSE}
find_algorithms("overlay", name_only = TRUE)
get_usage("grass:v.overlay")
# get_options("grass:v.overlay")
```

```{r eval=FALSE}
habmap_overlap_diff <- 
    run_qgis("grass:v.overlay", 
             ainput = habmap_overlap, 
             binput = watsurf_pol_u, 
             operator = "not",
                 output = "habmap_overlap_diff", # file.path(tempdir(), "habmap_overlap_diff.shp"),
                 load_output = TRUE)
# Error in UseMethod("st_as_sf") : no applicable method for 'st_as_sf' applied to an object of class "c('sfc_MULTIPOLYGON', 'sfc')"
```


# Using GRASS directly from R

[GRASS](https://grass.osgeo.org) is a powerful open source GIS specialized in geospatial processing, originally developed by the USA-CERL (1982-1995), and with development currently led by the GRASS Development Team.
It is now at version 7.6 (stable release).
The `rgrass7` R-package allows to directly send GRASS commands to the GRASS GIS environment.
See paragraph 9.4 from the [book](https://geocompr.robinlovelace.net/gis.html#rgrass) 'Geocomputation with R'.

The process goes smooth:

```{r}
# Set up a temporary GRASS project:
link2GI::linkGRASS7(habmap_std_pol, # sets the CRS etc.
                    search_path = if (.Platform$OS.type == "unix") {
                      "/usr/bin"} else NULL)
```

```{r}
writeVECT(SDF = as(habmap_std_pol, "Spatial"), 
          vname = "habmap_std_pol")
```

It appears that mistakes are taken care of!
Looks good.

Note, a few marginally overlapping polygons exist.

While importing data sources can be done directly from the R objects, using the `writeVECT()` function (as above), it is in fact more efficient for larger datasets (such as `watersurfaces`) to let GRASS import the shapefiles directly if those already exist.
`writeVECT()` first writes an sp object as a temporary shapefile but that is not needed if you have a file already.
Furthermore, `writeVECT()` led to an R error in the case of `watersurfaces`, so that's why we do it as follows:

```{r}
execGRASS("v.in.ogr", 
          input = normalizePath("../../n2khab_data/10_raw/watersurfaces/watersurfaces.shp"),
          output = "watsurf_pol",
          flags = "o" # assume CRS is same as from GRASS, i.e. set CRS as epsg:31370
          )
```

Contents of the GRASS database:

```{r}
execGRASS("g.list", type = "vector,raster", flags = "p")
```

So now let's do the job.
The manual of `v.overlay` in the current stable release 7.6 is [here](https://grass.osgeo.org/grass76/manuals/v.overlay.html).

```{r}
system.time(
execGRASS("v.overlay",
          ainput = "habmap_std_pol", 
          binput = "watsurf_pol", 
          operator = "not",
          output = "habmap_diff")
)
```

Vector layers in the GRASS database:

```{r}
execGRASS("g.list", type = "vector", flags = "p")
```

Reading result back into R as an sf object:

```{r}
system.time(
habmap_diff <- 
    readVECT("habmap_diff") %>% 
    st_as_sf
)
```

Note that we follow the standard advice of `readVECT()` not to use the c-flag (`with_c=FALSE`).

Result:

```{r}
habmap_diff
```

Housekeeping:

```{r}
habmap_diff <- 
    habmap_diff %>% 
    select(polygon_id = a_polygon_id,
           description_orig = a_description_orig)
```

```{r}
habmap_diff
```

## Inspecting the result

Duplicate ID's:

```{r}
habmap_diff %>% 
    st_drop_geometry() %>% 
    count(polygon_id) %>% 
    count(n)
```

Duplicate polygon IDs probably have to do with clipping along semi-matching borders...

There are no empty ID's

```{r}
habmap_diff %>% 
    st_drop_geometry() %>% 
    filter(is.na(polygon_id))
```

Identifying the completely removed polygons, by their ID:

```{r}
habmap_std_pol %>% 
    anti_join(habmap_diff %>% 
                st_drop_geometry(),
              by = "polygon_id") %>% 
  nrow
```


```{r warning=FALSE}
habmap_std_patch %>% 
    anti_join(habmap_diff %>% 
                st_drop_geometry(),
              by = "polygon_id") %>% 
    group_by(type) %>% 
    summarise(count = n(),
              avg_phab = mean(phab)) %>% 
    ggplot(aes(x = type, y = count)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


Calculating the actual surface areas of the above:

```{r warning=FALSE}
whiped_areas <- 
  habmap_std_pol %>% 
    anti_join(habmap_diff %>% 
                st_drop_geometry(),
              by = "polygon_id") %>% 
  mutate(area = st_area(geom)) %>% 
  st_drop_geometry %>% 
  inner_join(habmap_std_patch, by = "polygon_id") %>% 
  mutate(area = phab * area / 100) %>% 
  select(polygon_id, type, area) %>% 
  group_by(type) %>% 
  summarise(area_ha = (sum(area) / 1e+4) %>% units::drop_units())
whiped_areas %>% 
  ggplot(aes(x = type, y = area_ha)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Total area (ha) of omitted polygons from `habitatmap_stdized`:

```{r}
whiped_areas$area_ha %>% sum
```

# Further conclusion for the workflow

After some discussion, it appears that 'watersurfaces' also effectively includes semi-terrestrial types such as 'rbbmr', which is the type most often clipped by the above operation.
This is not what we want --
we do not want those to disappear from the data!

So, it seems better to follow a conservative approach, both in `watersurfaces_hab` and in `habitatmap_terr`:

- `watersurfaces_hab` combines the currently available information on the presence of watersurfaces and the presence of types 31xx / 2190_a / (7220?) / rbbah, in order to provide the areas where aquatic types will probably or potentially be present.
This is a good basis for aquatic setups.
To that aim, `watersurfaces_hab` takes a _conservative_ approach in defining the aquatic areas, meaning that in reality quite some (semi-)terrestrial areas are covered as well!
A further interpreted form can be made, where code 3130 is interpreted as 3130_aom.
Current intention is to provide the option of interpretation as an argument of `read_watersurfaces_hab()`.
- `habitatmap_terr` delineates all other types (but still excluding 3260 and 1130), which is useful for terrestrial + semi-terrestrial setups.
It contains the interpreted terrestrial + semi-terrestrial habitats.
To this aim, when generating `habitatmap_terr`, we want to only omit polygons of which we are quite sure that they are aquatic!

The consequence of this is that there will be a deliberate overlap between `watersurfaces_hab` and `habitatmap_terr`, which perfectly makes sense.



