---
title: "Exploration of generating line segments with GRASS"
date: '`r paste("Version",lubridate::now())`'
output:
  html_document:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

# Data preparation and GRASS setup

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(sf)
library(n2khab)
library(rgrass7)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
if (.Platform$OS.type == "unix") {
  Sys.setenv(GRASS_PYTHON = reticulate::py_discover_config() %>% 
                          as.character() %>% 
                          .[1])
}
```

## A simple multilinestring

```{r}
one_multiline <- 
  read_sf("test_vector.gpkg")
```

```{r paged.print=FALSE}
one_multiline
```


```{r}
st_length(one_multiline)
```


```{r}
one_multiline %>% 
  ggplot() +
  geom_sf() + 
  coord_sf(datum = st_crs(31370)) +
  theme_bw()
```


```{r}
one_multiline %>% 
  ggplot() +
  geom_sf(data = 
            read_admin_areas(dsn = "provinces") %>% 
            filter(name == "Limburg"),
          colour = NA) +
  geom_sf(colour = "red") + 
  coord_sf(datum = st_crs(31370)) +
  theme_bw()
```

## GRASS temporary project setup & data import in GRASS database

For more tedious geoprocessing we can use [GRASS](https://grass.osgeo.org/), through the packages `link2GI` en `rgrass7`.

```{r results="hide"}
# Set up a temporary GRASS project:
link2GI::linkGRASS7(one_multiline, # sets the CRS etc.
                    search_path = if (.Platform$OS.type == "unix") {
                      "/usr/bin"} else NULL,
                    use_home = TRUE)
use_sf()
```

Some GRASS project metadata:

```{r}
system("g.region -p
        g.proj -j")
```

Note that there's also:

```{r}
rgrass7::gmeta()
```

Add the sf object to the GRASS database:

```{r}
one_multiline %>% 
  writeVECT("one_multiline",
              v.in.ogr_flags = c("overwrite", "o"))
```

Check presence of the layer:

```{r}
execGRASS("g.list", type = "vector", flags = "p")
```

## Working interactively with GRASS from R

Launch the GUI:

```{r eval=FALSE}
system("g.gui")
```

Launch the grass terminal from R (in Linux):

```{r eval=FALSE}
mapset_path <- 
  system("g.gisenv get=GISDBASE,LOCATION_NAME,MAPSET sep='/'", 
         intern = TRUE)
system(paste0("gnome-terminal --working-directory=",
              mapset_path,
              " -e 'grass --text .'")
)
```

Read manual for `v.split`:

```{r eval=FALSE}
system("g.manual v.split")
```

That's the same as (potentially more generic):

```{r eval=FALSE}
execGRASS("g.manual",
          entry = "v.split")
```


# Geoprocessing

## v.split

```{r}
system("v.split -f input=one_multiline output=split1 length=100")
```

```{r}
split1 <- readVECT("split1")
split1 %>% 
  cbind(col = factor(1:4)) %>% 
  ggplot(aes(colour = col)) + geom_sf(size = 1.5)
```

```{r paged.print=FALSE}
split1 %>% as_tibble() %>% st_as_sf()
```


# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
sprintf("- **%s**:\n %s\n", names(p), p) %>%
  cat()
```

```{r results = "asis", echo=FALSE}
si$packages %>%
    as_tibble %>%
    select(package, loadedversion, date, source) %>%
pander::pandoc.table(caption = "(\\#tab:sessioninfo)Loaded R packages",
                     split.table = Inf)
```




