---
title: "Exploration of generating line segments with GRASS"
date: '`r paste("Version",lubridate::now())`'
params:
  gisbase_grass: !r if (.Platform$OS.type == "windows") link2GI::paramGRASSw()$gisbase_GRASS[1] else link2GI::paramGRASSx()$gisbase_GRASS[1]
output:
  html_document:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

# Data preparation and GRASS setup

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(sf)
library(n2khab)
library(rgrass7)
library(knitr)
library(mapview)
mapviewOptions(fgb = FALSE)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
if (.Platform$OS.type == "unix") {
  Sys.setenv(GRASS_PYTHON = system("which python3", intern = TRUE))
}

```

## A simple multilinestring

```{r}
one_multiline <- 
  read_sf("test_vector.gpkg")
```

```{r paged.print=FALSE}
one_multiline
```


```{r}
st_length(one_multiline)
```


```{r}
one_multiline %>% 
  ggplot() +
  geom_sf() + 
  coord_sf(datum = st_crs(31370)) +
  theme_bw()
```


```{r}
one_multiline %>% 
  ggplot() +
  geom_sf(data = 
            read_admin_areas(dsn = "provinces") %>% 
            filter(name == "Limburg"),
          colour = NA) +
  geom_sf(colour = "red") + 
  coord_sf(datum = st_crs(31370)) +
  theme_bw()
```

## A longer multilines object

We shorten the file in order to inspect results more easily:

```{r}
more_multilines <- 
  read_sf("habitatmap3260_v14.gpkg") %>% 
  mutate(length = st_length(.) %>% units::drop_units()) %>% 
  filter(provincie == "Limburg",
         length > 1500)
```

```{r paged.print=FALSE}
more_multilines
```

We better cast it to linestrings in order to distinguish single linestrings (one multilinestring was composed of two linestrings):

```{r paged.print=FALSE}
linestrings <- 
  more_multilines %>% 
  st_cast("LINESTRING")
linestrings
```




```{r}
summary(linestrings$length)
```

```{r}
linestrings %>% 
  ggplot() +
  geom_sf(data = 
            read_admin_areas(dsn = "provinces") %>% 
            filter(name == "Limburg"),
          colour = NA) +
  geom_sf(colour = "red") + 
  coord_sf(datum = st_crs(31370)) +
  theme_bw()
```


## GRASS temporary project setup & data import in GRASS database

For more tedious geoprocessing we can use [GRASS](https://grass.osgeo.org/), using the package `rgrass7`.

```{r results="hide"}
# Set up a temporary GRASS project:
if (interactive()) {
gisbase_grass <- 
  if (.Platform$OS.type == "windows" && nchar(Sys.getenv("OSGEO4W_ROOT")) > 0) {
    # needed only until https://github.com/r-spatial/link2GI/issues/48 is fixed
  "C:/OSGeo4W64/apps/grass/grass78"
} else {
  if (.Platform$OS.type == "windows") link2GI::paramGRASSw()$gisbase_GRASS[1] else {
    link2GI::paramGRASSx()$gisbase_GRASS[1]
  }
}
}
initGRASS(gisBase = if (interactive()) gisbase_grass else params$gisbase_grass,
          home = tempdir(),
          mapset = "PERMANENT", 
          override = TRUE)
execGRASS("g.proj", c("c", "quiet"), 
          georef = file.path("test_vector.gpkg"))
use_sf()
```

Wrapper function to pass full GRASS command syntax as a string from R:

```{r}
execshell <- function(commandstring, intern = FALSE) {
  if (.Platform$OS.type == "windows") {
    res <- shell(commandstring, intern = TRUE)
  } else {
    res <- system(commandstring, intern = TRUE)
  }
  if (!intern) cat(res, sep = "\n") else return(res)
}
```


Some GRASS project metadata:

```{r}
execshell("g.region -p
           g.proj -w")
```

Note that there's also:

```{r}
rgrass7::gmeta()
```

Add the sf objects to the GRASS database:

```{r}
one_multiline %>% 
  writeVECT("one_multiline",
              v.in.ogr_flags = c("overwrite", "o"))
linestrings %>% 
  writeVECT("linestrings",
              v.in.ogr_flags = c("overwrite", "o"))
```

Check presence of the layers:

```{r}
execGRASS("g.list", type = "vector", flags = "p")
```

## Working interactively with GRASS from R

Launch the GUI:

```{r eval=FALSE}
execshell("g.gui")
```

Launch the grass terminal from R (in Linux):

```{r eval=FALSE}
mapset_path <- 
  system("g.gisenv get=GISDBASE,LOCATION_NAME,MAPSET sep='/'", 
         intern = TRUE)
system(paste0("gnome-terminal --working-directory=",
              mapset_path,
              " -- grass --text .")
)
```

Read manual for `v.split`:

```{r eval=FALSE}
execshell("g.manual v.split")
```


# Geoprocessing

## GRASS: v.split + v.to.points

### one_multiline

```{r}
execshell("v.split -f input=one_multiline output=split1 length=100")
```

```{r}
split1 <- readVECT("split1")
split1 %>% 
  cbind(col = factor(1:4)) %>% 
  ggplot(aes(colour = col)) + geom_sf(size = 1.5)
```

```{r paged.print=FALSE}
split1 %>% as_tibble() %>% st_as_sf()
```

### linestrings

Now using the rgrass7 function:

```{r}
execGRASS("v.split",
          flags = c("f", "overwrite"),
          input = "linestrings",
          output = "split2",
          length = 100)
```

Also generate points at the segment ends:

```{r}
execGRASS("v.to.points",
          flags = c("overwrite"),
          input = "split2",
          output = "split2points",
          use = "end")
```


```{r out.width='800px', out.height='500px'}
split2 <- readVECT("split2")
split2points <- readVECT("split2points")
segments <- 
  split2 %>% 
  cbind(col = factor(1:2)) %>% 
  mutate(segment_id = 1:nrow(.))
endpoints <- 
  split2points %>% 
  cbind(col = factor(1:2)) %>% 
  mutate(point_id = 1:nrow(.))
mapview(segments,
        zcol = "col",
        label = "segment_id",
        color = c("green3", "purple"),
        lwd = 4) +
  mapview(endpoints,
        zcol = "col",
        label = "point_id",
        cex = 5,
        col.regions = c("green3", "purple"),
        color = c("green3", "purple"))
```

```{r paged.print=FALSE}
split2 %>% as_tibble() %>% st_as_sf()
```

We can see that each linestring was segmented **independently**, resulting in as many segments shorter than 100 m as there are linestrings.

```{r}
split2 %>% 
  mutate(length = st_length(.) %>% round(2)) %>% 
  st_drop_geometry() %>% 
  count(length) %>% 
  mutate(rank = 1:nrow(.)) %>% 
  kable
```

## Comparison with `sp::spsample()`

We do this only for the `linestrings` object.

We make a `pointsample1` and a `pointsample2` object.
The difference is that `pointsample2` misses the watercourse where `OBJECTID==413`.
In this way, stability of points can be verified toward changes in the dataset.

```{r paged.print=FALSE, warning=FALSE}
length_total <- 
  linestrings %>% 
  st_length() %>% 
  sum %>% 
  units::drop_units()
pointsample1 <- 
  sp::spsample(as_Spatial(linestrings),
               n = ceiling(length_total/100),
               type="regular",
               cellsize=100,
               offset=0) %>% 
  st_as_sf
```

```{r paged.print=FALSE, warning=FALSE}
length_total2 <- 
  linestrings %>% 
  filter(OBJECTID != 413) %>% 
  st_length() %>% 
  sum %>% 
  units::drop_units()
pointsample2 <- 
  sp::spsample(as_Spatial(linestrings %>% 
                            filter(OBJECTID != 413)),
               n = ceiling(length_total2/100),
               type="regular",
               cellsize=100,
               offset=0) %>% 
  st_as_sf
```

Plot the points on top of the previous map:


```{r out.width='800px', out.height='500px'}
mapview(segments,
        zcol = "col",
        label = "segment_id",
        color = c("green3", "purple"),
        lwd = 4) +
  mapview(endpoints,
        zcol = "col",
        label = "point_id",
        cex = 5,
        col.regions = c("green3", "purple"),
        color = c("green3", "purple")) +
  mapview(pointsample1,
        cex = 5,
        col.regions = "red") +
  mapview(pointsample2,
        cex = 5,
        col.regions = "blue")
```


# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
sprintf("- **%s**:\n %s\n", names(p), p) %>%
  cat()
```

```{r results = "asis", echo=FALSE}
si$packages %>%
    as_tibble %>%
    select(package, loadedversion, date, source) %>%
pander::pandoc.table(caption = "(\\#tab:sessioninfo)Loaded R packages",
                     split.table = Inf)
```




