---
title: "Handling the soilmap"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# Exploring the soilmap data source

```{r}
soilmap <- read_sf(file.path(fileman_up("n2khab_data"), "10_raw/soilmap"),
                   stringsAsFactors = TRUE) # for exploring unique values
```


```{r}
soilmap %>% 
  st_drop_geometry %>% 
  summary
```

Number of unique values of numeric variables:

```{r}
ul <- function(x) x %>% unique %>% length
list(soilmap$gid,
     soilmap$id_kaartvl,
     soilmap$Kaartbldnr) %>% 
  lapply(ul)
```

Look at this:

```{r}
all(soilmap$gid == soilmap$id_kaartvl)
```


Number of NA's of some numeric variables:

```{r}
nal <- function(x) x %>% is.na %>% sum
list(soilmap$gid,
     soilmap$id_kaartvl,
     soilmap$Kaartbldnr,
     soilmap$codeid) %>% 
  lapply(nal)
```

Inspecting the number of levels of variables.
Note that in several cases, the number of different codes does not match the number of different meanings/explanations.

```{r}
soilmap %>% st_drop_geometry %>% lapply(ul)
```

String inconsistencies do happen:

```{r}
soilmap$Type_class %>% unique %>% as.matrix
```

`codeid` refers to `Bodemtype` according to the metadatafile "20140814Uitleg_bij_veldnamen_bodemkaart.docx", and indeed it has the same number of levels.
Somehow the order of variables is a bit confusing here.

Explanations of `Fase_c` are not data, but metadata:

```{r}
soilmap %>% 
  st_drop_geometry %>% 
  select(contains("Fase")) %>% 
  distinct
```

## Exploring the relation between `Bodemtype` and `Unitype`

```{r}
soilmap %>% 
  mutate(same = as.character(Bodemtype) == as.character(Unitype)) %>% 
  .$same %>% summary
```

```{r}
soilmap %>% 
  st_drop_geometry %>% 
  mutate(same = as.character(Bodemtype) == as.character(Unitype)) %>% 
  filter(!same) %>% 
  count(Bodemtype, Textuur_c, Bodemser_c, Type_c, Unitype)
```


```{r}
soilmap %>% 
  st_drop_geometry %>% 
  mutate(same = as.character(Bodemtype) == as.character(Unitype)) %>% 
  filter(!same) %>% 
  .$Bodemtype %>% str_sub(1,1) %>% unique
```
