---
title: "Handling scheme_types in relation to the habitatmap"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# Intro

Some inspiration is [here](https://github.com/inbo/n2khab-mne-design/blob/master/310_design_atmosphere/110_vlops_datapreparation/20_verkenning.Rmd#L33) already.

```{r}
habitatmap_patches <- 
  read_habitatmap_stdized("../../data") %>% 
  .$habitatmap_patches
```

```{r}
habitatmap_polygons <- 
  read_habitatmap_stdized("../../data") %>% 
  .$habitatmap_polygons
```

```{r}
scheme_types <- read_scheme_types()
```

Let's prepare for a function that expands the types in a dataframe in some appropriate way, in order to better join with the habitatmap.

```{r}
types <- read_types()
types
```

# Main type - subtype equivalence?

Which subtypes are so rare within the main type that they may be ignored when abstracting from subtypes to main types (i.e. taking a group of subtypes and adding the main type code)? ^[
However, the further conclusion shows that we will actually only need to abstract from subtypes to main types for those main types where the reverse abstraction from main types to _one specific subtype_ was not possible.
Plausible reverse abstraction from main types to _one specific subtype_ will already be made available in a derived data source `terr_habitatmap`, which is to be used as the data source for the envisaged function.]

```{r fig.height=10, fig.width=7, warning=FALSE}
types_sub <-
  types %>% 
  filter(typelevel == "subtype") %>% 
  distinct(main_type) %>% 
  semi_join(types, .) %>% 
  pull(type)
patches_sub <- 
  habitatmap_patches %>% 
  filter(type %in% types_sub,
         !str_detect(code_orig, "bos"),
         !(code_orig %in% c("6510,gh",
                            "9120,gh"))) %>% 
                      ## only a tiny fraction of xxx,gh or xxx,bos 
                      ## is expected to be habitat type or rbb
  select(polygon_id, phab, type, certain)
types_sub_areas <- 
  habitatmap_polygons %>% 
  semi_join(patches_sub) %>% 
  mutate(area = st_area(.)) %>% 
  select(polygon_id, area) %>% 
  st_drop_geometry %>% 
  inner_join(patches_sub, .) %>% 
  inner_join(types %>% select(1:3)) %>% 
  mutate(area = area * phab / 100) %>% 
  group_by(main_type, type, certain) %>% 
  summarise(area = sum(area))
types_sub_areas %>% 
  group_by(main_type) %>% 
  mutate(rel_area = area / sum(area)) %>% 
  ggplot(aes(x = type, y = rel_area, fill = certain)) +
    geom_bar(stat = "identity") +
    facet_wrap(~main_type, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Several things are to be noted:

- several codes of `types_sub` (i.e. the list of main type & subtype codes that correspond to the main types that do have subtypes) are not present in the habitatmap. Some are main type codes, some are subtype codes. Only in case of 2190 this poses an interpretation challenge.

```{r}
types_sub[!(types_sub %in% types_sub_areas$type)] %>% 
  as.character %>% 
  as.matrix
```

- 9120, rbbkam, rbbvos and rbbzil currently have an inconsistent meaning.
These codes can refer to a main type or to an implicit subtype, i.e. to the main type minus the single explicit subtype they possess.
This confusion is a real problem and should be solved in later typologies.
The habitatmap currently uses this for the implicit subtype!
  - For these, `scheme_types` _currently_ only holds main type codes (table below).
Therefore no problem is _currently_ to be expected.

```{r}
scheme_types %>% 
  distinct(type) %>% 
  filter(str_detect(type, "9120|rbbvos|rbbzil|rbbkam"))
```

Which of `types_sub` are present in `scheme_types`?

```{r}
scheme_types %>% 
  inner_join(types %>% select(1:3)) %>% 
  distinct(main_type, type) %>% 
  filter(type %in% types_sub) %>% 
  arrange(main_type, type)
```

## Conclusion based on discussion with Steven De Saeger

### Habitatmap interpretation

Regarding **interpretation** of the habitatmap, which best goes into `terr_habitatmap` (the interpreted and augmented form of `habitatmap_stdized`) and `watersurfaces_interpr`:

- when we only consider records of `habitatmap_patches` _with considerable probability_ of effectively holding a habitat type or RIB (RBB) ^[
This selection can be based on the following condition: `!str_detect(code_orig, "bos") & !(code_orig %in% c("6510,gh", "9120,gh"))`.
]:
for some main types, the main type code _mostly_ refers to a specific subtype -- so _that_ code can be used instead when interpreting the `habitatmap_stdized` data source:
  - 3130 -> 3130_aom
  - 6410 -> 6410_mo
  - 6430 -> 6430_hf
  - 6510 -> 6510_hu
  - 7140 -> 7140_meso
  - 9130 -> 9130_end
- considering code **2190** in the habitatmap, this code refers to the sum of 2190_a and 2190_overig.
It is necessary to use BWK fields from the raw `habitatmap` data source in order to distinguish 2190_a.
Even better, use the `watersurfaces` data source for this, within the dunes ecoregion (data source `ecoregions`).
So it will probably be better to apply this at once in the `terr_habitatmap` and in an interpreted form of `watersurfaces` (`watersurfaces_interpr`), which should be spatially exclusive (and together will be part of `integrated_habitatmap`).
- not having to do with subtypes, but with the previous point: also rbbhfl is a code which needs interpretation from BWK + habitatmap, in order to make it 'appear' (these polygons currently appear as rbbhf).

So `terr_habitatmap` should:

- exclude aquatic types and all areas covered by `watersurfaces`.
- _not_ include the following 'uncertain' values of `code_orig` (has been discussed; these only very rarely will be habitat type or rbb):
  - containing "bos"
  - equal to "6510,gh" or to "9120,gh"
- provide the interpreted values (see above), plus a column that makes it apparent what the source of a value is:
  - habitatmap_stdized
  - habitatmap_stdized + interpretation
  - habitatdune
  
### Function to expand a type column

So this means that a **function** that works on a type column ^[e.g. the output of `read_scheme_types()`, filtered for a specific scheme], 
expanding main types into (all) subtypes and adding the main type for sets of subtypes, would be reasonable, 
considering that `terr_habitatmap` will effectively present those cases where this approach is indeed reasonable.
The function, then, is intended to make its result useful only when based on `terr_habitatmap` (and further derived data sources).

The aspect of 'adding the main type for sets of subtypes' would in general require the 'complete set of subtypes' to be available, but:

- we will relax this for:
  - 6230: only needs 6230_ha, 6230_hmo and 6230_hnk in the type column, not the rare 6230_hnk
  - 91E0: only needs 91E0_va, 91E0_vm, 91E0_vn, not the rarer sf, vc and vo subtypes.
- we simply won't do it for 1310, 1330, 2130, 2190, 3130, 6210, 6410, 6430, 6510, 7140, 9130.
That is because `terr_habitatmap` will not have those codes.
- we should neither do this for rbbkam, rbbvos, rbbzil & 9120, as the main type also has another subtype which has no existant code; hence the sole distinguished subtype does not cover the whole main type!
- That means it _will_ be done only for 2330, 6230 & 91E0.

The addition of the main type code based on subtype codes could _optionally_ also be relaxed by the function (e.g. by an argument `strict = FALSE`, with default `TRUE`):
it could allow the addition of a main type code when just _one not too rare_ (see above) subtype code is present.
That would introduce more errors, but it would allow a 'maximal' definition of the target population.
That a selected location effectively belongs to the target population would then have to be checked on-site!
By the way, the latter always has to be done as there are intrinsic uncertainties in the habitatmap.
When checking on-site, always the strict list of target population defining types has to be used, not the expanded one.

# Some further queries on habitatmap_stdized

Which values of `code_orig` exist for uncertain records (i.e., containing a 'comma')?

```{r}
habitatmap_patches %>% 
  filter(!certain) %>%
  count(code_orig) %>% 
  kable()
```

Which records have phab == 0?

```{r}
habitatmap_patches %>% 
  filter(phab == 0) %>% 
  count(type, certain) %>% 
  arrange(-n) %>% 
  kable
```

Note that these generally are real occurrences of the type, as 'line'- or 'point'-shaped polygons in the field (KLE).
For some types this is even an important aspect (e.g. rbbsp), despite the non-registered surface area.

