---
title: "Handling the watersurfaces data source"
date: '`r paste("Version",lubridate::now())`'
output:
  html_notebook:
    number_sections: yes
    code_folding: hide
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```


# Exploring the watersurfaces data source

```{r}
ws_10 <- read_sf(file.path(fileman_up("n2khab_data"), "10_raw/watersurfaces"),
                   stringsAsFactors = TRUE) # for exploring unique values

ws_11 <- read_sf(file.path(fileman_up("n2khab_data"), "10_raw/watersurfaces/watersurfaces.gpkg"),
                   stringsAsFactors = TRUE) 



```


```{r}
ws %>% 
  st_drop_geometry %>% 
  summary
```

Number of unique values of numeric variable `HYLAC`:

```{r}
ws$HYLAC %>% unique %>% length
```

```{r eval=FALSE}
ul <- function(x) x %>% unique %>% length
list(ws$HYLAC) %>% 
  lapply(ul)
```

How many `NA` values for numeric variable `HYLAC`?

```{r}
ws$HYLAC %>% is.na %>% sum
```

Potential issues:

- `WTRLICHC`: 10 `<Null>` values which should be `NA`
- `HYLAC`: `0` values should be `NA` (empty values, i.e. no code available)
- `NAAM`: 5 `<Null>` values which should be `NA`
- `CONNECT`, `KRWTYPES`: levels in the dataset do not yet conform to the codes in the metadata report (Packet et al 2018)
- There seems to be a mistake in the (explanation of) classes of `CONNECT` in the metadata report (Packet et al 2018): 'inzonder' vs. 'uitmet'

These aspects will ± be taken care of by `read_watersurfaces()`.

# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
ws %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`WVLC`                        `polygon_id` -- note: these are currently not unique [^note]
`WTRLICHC`                    `wfd_code`
`HYLAC`                       `hyla_code`
`NAAM`                        `name`
`GEBIED`                      `area_name`
`KRWTYPE`                     `wfd_type `
`KRWTYPES`                    `wfd_type_certain`
`DIEPKL`                      `depth_class`
`CONNECT`                     `connectivity`
`FUNCTIE`                     `usage`

Other considerations for the R object returned by `read_watersurfaces()`:

- not uptaking `OPPWVL` & `OMTWVL` (area & perimeter are easily calculated etc) -- OK
- sort by `polygon_id` -- OK
- add translations to long text for `wfd_type `,
`connectivity` and `usage`? -- OK, but not by default
- converting null / 0 values to `NA` -- OK
- combining polygons with same `WVLC` value into a multipolygon -- no (see note with `polygon_id`)

[^note]: A new version of the data source is now underway and will effectively 
provide unique codes.
So while it is not unique in the current version, we will apply this variable name anyway.


