---
title: "Handling the watersurfaces data source"
date: "`r paste("Version",lubridate::now())`"
  output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    number_sections: yes
    code_folding: hide
    includes:
      in_header: ../header.html
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
# library(sp)
library(sf)
library(raster)
library(tidyverse)
library(n2khab)
# # library(plotly)
# library(rasterVis)
# library(stars)
# library(units)
# library(tmap)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```

# Exploring the watersurfaces data source

The previous version was 1.0 (10.5281/zenodo.3386859), we want to explore the changes in the most recent version: v 1.1 (10.5281/zenodo.4117543)

Import the geometrical layer:

```{r geometrical-layer-v10}
# manually adapt the path
# ws_10 <- read_sf(file.path(fileman_up("n2khab_data"), "10_raw/watersurfaces/versies/v1.0"),
#                    stringsAsFactors = TRUE) # n = 86026

```

```{r v11}

filepath_11 <- file.path(fileman_up("n2khab_data"), "10_raw/watersurfaces/watersurfaces.gpkg")

st_layers(filepath_11)

```
```{r geometrical-layer-v11}

ws_11 <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "Watervlakken")  # n = 88713

```

The other (non geometrical) layers:

```{r non-geometrical-layers}
# only available in version 1.1

ws_peilbeheer <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktPEILBEHEER") 
ws_krwtypes <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktKRWTYPES")
ws_krwtype <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktKRWTYPE")
ws_functie <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktFUNCTIE")
ws_diepkl <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktDIEPKL")
ws_connect <- read_sf(filepath_11, stringsAsFactors = TRUE, layer = "LktCONNECT")

```

### Encoding problems (Windows only)

For Windows users, we will need to correct some encoding problems, as shown in the examples hereunder:

```{r encoding-problems}
osVersion
sessionInfo()$R.version$version.string
sessionInfo()$R.version$platform

ws_diepkl

ws_connect

# 1) geometrical layer:

# sf_utf8 <- sf::st_read(filepath_11, stringsAsFactors = TRUE, layer = "Watervlakken", options = "ENCODING=UTF-8")%>% 
#   filter( WVLC =="ANTMOL0152")
# options = "ENCODING=UTF-8" cannot be used with a geopackage


# as imported by default:
ws_11 %>% st_drop_geometry %>% 
  filter( WVLC =="ANTMOL0152") %>% select(WVLC, DIEPKL, CONNECT)

#conversion latin1 - UTF-8
ws_11_2 <- n2khab:::convertdf_enc(ws_11 , from = "UTF-8", to = "latin1")
ws_11_2 %>% st_drop_geometry %>% 
  filter( WVLC =="ANTMOL0152") %>% select(WVLC, DIEPKL, CONNECT)

ws_11_3 <- n2khab:::convertdf_enc(ws_11 , from = "latin1", to = "UTF-8")
ws_11_3 %>% st_drop_geometry %>% 
  filter( WVLC =="ANTMOL0152") %>% select(WVLC, DIEPKL, CONNECT)

#conversion WINDOWS-1252 - UTF-8
ws_11_4 <- n2khab:::convertdf_enc(ws_11 , from = "UTF-8", to = "WINDOWS-1252")
ws_11_4 %>% st_drop_geometry %>% 
  filter( WVLC =="ANTMOL0152") %>% select(WVLC, DIEPKL, CONNECT)

ws_11_5 <- n2khab:::convertdf_enc(ws_11 , from = "WINDOWS-1252", to = "UTF-8")
ws_11_5 %>% st_drop_geometry %>% 
  filter( WVLC =="ANTMOL0152") %>% select(WVLC, DIEPKL, CONNECT)

# 2) non geometrical layers:

# latin1 to UTF-8: not OK:
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktCONNECT"), from =  "latin1", to ="UTF-8")) 
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktDIEPKL"), from =  "latin1", to ="UTF-8")) 

# UTF-8 to latin1: 
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktCONNECT"), from = "UTF-8", to = "latin1"))
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktDIEPKL"), from = "UTF-8", to = "latin1"))

# UTF-8 to WINDOWS-1252: 
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktCONNECT"), from = "UTF-8", to = "WINDOWS-1252"))
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktDIEPKL"), from = "UTF-8", to = "WINDOWS-1252"))
  
# UTF-8 to UTF-8: 
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktCONNECT"), from = "UTF-8", to = "UTF-8"))
print(n2khab:::convertdf_enc(read_sf(filepath_11, layer = "LktDIEPKL"), from = "UTF-8", to = "UTF-8"))
   
```

This is a Windows-related problem, so we can add a condition `if (.Platform$OS.type == "windows") { }` to deal with it.

```{r correct-encoding-windows}
if (.Platform$OS.type == "windows") {
  
# only really needed for CONNECT + a correction for the >= <= in DIEPKL
  
# ws_peilbeheer <- n2khab:::convertdf_enc(ws_peilbeheer, from = "UTF-8", to = "latin1")
# ws_krwtypes <- n2khab:::convertdf_enc(ws_krwtypes, from = "UTF-8", to = "latin1")
# ws_krwtype <- n2khab:::convertdf_enc(ws_krwtype, from = "UTF-8", to = "latin1")
# ws_functie <- n2khab:::convertdf_enc(ws_functie, from = "UTF-8", to = "latin1")
  ws_connect <- n2khab:::convertdf_enc(ws_connect, from = "UTF-8", to = "latin1")

ws_diepkl <- ws_diepkl %>% 
  mutate(Omschr = case_when(
    Code == "2 - 4 m" ~ ">= 2 - < 4 m",
    Code == "4 - 6 m" ~ ">= 4 - < 6 m",
   # Code == "> 6 m" ~ as.character(expression("\u2265 6 m")),
  # Code == "> 6 m" ~ "\u2265 6 m",
   Code == "> 6 m" ~ ">= 6 m",
    TRUE ~ Omschr
  )) 

# How does this look like in html?
# Using one of the following options 
# Code == "> 6 m" ~ as.character(expression("\u2265 6 m")),
# Code == "> 6 m" ~ "\u2265 6 m",
# would look like a "=" in html, so we might be better using an alternative such as ">="

# We will have to do the same for ws_11 (see below)

  }
```

### A summary of the geometrical table for version 1.1

```{r}
# ws_10 %>% 
#   st_drop_geometry %>% 
#   summary
```

```{r}
ws_11 %>% 
  st_drop_geometry %>% 
  summary
```
Let us look for a few typical errors more systematically in the new version 1.1 (geometrical layer)

### Are there `NA` values?

There are plenty of NA's but only in fields where we expect them.

```{r}
sapply(ws_11, function(x) sum(is.na(x)))

```

### Are there `<Null>` values?

Corrected since v1.0: no  `<Null>` values anymore in `WTRLICHC` or `NAAM` or any of the fields

```{r}
# slow - 25 sec
sapply(ws_11, function(x) sum(as.character(x) == '<Null>', na.rm = TRUE))
```

### Are there Zero (0) values?

Really plenty of zeroes for `HYLAC`

```{r}
# slow - 25 sec
sapply(ws_11, function(x) sum(as.character(x) == '0', na.rm = TRUE))
```

Number of unique values of numeric variable `HYLAC`:

```{r}
ws_11$HYLAC %>% unique %>% length
```

```{r eval=FALSE}
ul <- function(x) x %>% unique %>% length
list(ws_11$HYLAC) %>% 
  lapply(ul)
```

How many `NA` values for numeric variable `HYLAC`?

None!

```{r}
ws_11$HYLAC %>% is.na %>% sum
```
### Are WVLC codes unique?

Yes. It was not the case in version 1.0, but it has been corrected in v1.1

```{r}
ws_11$WVLC %>% unique %>% length
```


### Levels for each factor

We can compare the levels for each factor variable with the information given in 
[Leyssen et al. 2020](https://pureportal.inbo.be/nl/publications/watervlakken-versie-11-polygonenkaart-van-stilstaand-water-in-vla) and with the extra layers in the geopackage.

I do not check `NAAM` and `GEBIED` since there are many possible options.

**KRWTYPE**: the codes are correct but there are more codes mentioned in the metadata report and in the geopackage (layer LktKRWTYPE)

```{r}
levels(ws_11$KRWTYPE)

ws_krwtype 
```
**KRWTYPES** (status): the codes in the geometrical layer are the same as in the metadata report but not as in the geopackage (layer LktKRWTYPES)

```{r}
levels(ws_11$KRWTYPES)

ws_krwtypes 
```
**DIEPKL**: there are extra codes in the dataset ("> 6 m" and "0 - 2 m") and the ordering of 
the levels could be more logical. 

```{r}
levels(ws_11$DIEPKL)

ws_diepkl

# for Windows users we will recode the levels with => instead of ≥/\u2265 :
# (otherwise "≥" in ws_11$DIEPKL are rendered as "=" in the html output) 

if (.Platform$OS.type == "windows") {
  
  levels(ws_11$DIEPKL) <- gsub(pattern = "\u2265", ">=", levels(ws_11$DIEPKL))
  
  # ws_11$DIEPKL <- ordered(ws_11$DIEPKL, levels = c("0 - < 2 m", "0 - 2 m", ">= 2 - < 4 m", ">= 4 - < 6 m", ">= 6 m", "> 6 m"))
  
levels(ws_11$DIEPKL)
  }

```



**CONNECT**: the codes are correct

```{r}
levels(ws_11$CONNECT)

ws_connect 
```

And here are the categories in the metadata report for `CONNECT`

- geïsoleerd: niet verbonden met een waterloop 
- permanent: het watervlak staat permanent in verbinding met minstens één
waterloop 
- periodiek: het watervlak staat tijdelijk (door peilbeheer of droogte) in verbinding
met minstens één waterloop

**FUNCTIE**: the code "veedrenk" is not mentioned in the metadata report and 
there are many more codes in the report and layer of the geopackage than in the 
dataset

```{r}
levels(ws_11$FUNCTIE)

ws_functie
```

And here are the categories in the metadata report for `FUNCTIE`:

functie                       toewijzing
----------------------        ---------------------
natuur                        doelstelling natuurbehoud
hengelintensief               intensief hengelen (met infrastructuur, bepoting of gebruikt voor wedstrijdhengelen)
hengelextensief               extensief hengelen (geen infrastructuur, bepoting of wedstrijdhengelen)
jacht                         jagen
tuin/park                     esthetisch (verblijfsrecreatie, tuin- en parkvijvers)
vogel                         waterpartij voor gedomesticeerde watervogels
viskweek                      opkweken van vis
zwemmen                       zwemmen
duiken                        duiken
zachterecreatie               niet gemotoriseerde watersport
motorrecreatie                gemotoriseerde watersport
berging                       waterberging ten behoeve van overstromings- of peilbeheer
opslag                        reservoir voor water (industrie, landbouw, bluswater, waterkracht…)
drinkwater                    drinkwaterwinning
zuivering                     (kleinschalige) waterzuivering, infiltratie
bezinking                     bezinking van proceswater
drinkplaats                   watervoorziening voor vee
geen                          geen specifieke functie

## Potential issues:

- (`PEILBEHEER`: mentioned in report Leyssen et al. 2020 but not available in the dataset)
- `HYLAC`: `0` values should be `NA` (empty values, i.e. no code available)
- `KRWTYPE`, `FUNCTIE`: more levels in the metadata report (Leyssen et al. 2020) than in the dataset 
- `KRWTYPES`: can not use the corresponding layer LktKRWTYPES because the codes are not the same
- `DIEPKL`, `FUNCTIE`: extra code in the dataset that is not present in Leyssen et al. 2020

The problem with `HYLAC` = 0 will be taken care of by `read_watersurfaces()`. 
The extra codes will be discussed with the authors of the layers.


## Let's plot the watersurfaces as a map

```{r plot}

# plot watersurfaces v1.1
 p <- ggplot() +
  geom_sf(data = ws_11, aes(), color = "blue")

# Flanders
sf_vl <- read_admin_areas()

p <- p + 
  geom_sf(data = sf_vl, fill = NA)

print(p)

```

<!-- ### Are all the watersurfaces in Flanders? -->

<!-- ```{r streams-in-fl} -->
<!-- in_vl <- st_contains(x = sf_vl, y = ws_11) -->

<!-- not_contained <- ws_11 %>%  -->
<!--   st_drop_geometry() %>%  -->
<!--   mutate(orig_rowname = rownames(.)) %>%  -->
<!--   filter(!rownames(.) %in% in_vl[[1]]) -->

<!-- # 61 watersurfaces -->
<!-- print(not_contained %>% nrow()) -->

<!-- not_contained %>%  -->
<!--   distinct(WVLC, NAAM) %>%  -->
<!--   kable() -->
<!-- ``` -->
<!-- 61 watersurfaces are not 'contained' within the polygon for Flanders (see list above) -->

# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
ws_11 %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`WVLC`                        `polygon_id` 
`WTRLICHC`                    `wfd_code`
`HYLAC`                       `hyla_code`
`NAAM`                        `name`
`GEBIED`                      `area_name`
`KRWTYPE`                     `wfd_type `
`KRWTYPES`                    `wfd_type_certain`
`DIEPKL`                      `depth_class`
`CONNECT`                     `connectivity`
`FUNCTIE`                     `usage`

**Other considerations for the R object returned by `read_watersurfaces()`**:

- not uptaking `OPPWVL` & `OMTWVL` (area & perimeter are easily calculated etc) -- OK
- converting 0 values in `HYLAC` to `NA` -- OK
- sort by `polygon_id` -- OK
- add translations to long text for `wfd_type ` and `connectivity`  -- OK but not by default
- add translations to long text for  `usage`? -- for a later version (as more 
codes will be used)

