---
title: "Handling the watersurfaces data source"
date: "`r paste('Version',lubridate::now())`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
---

```{r setup, message=FALSE, echo=FALSE}
library(sf)
library(dplyr)
library(stringr)
library(tmap)
library(n2khab)
library(ggplot2)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```

# Exploring the watersurfaces data source

We want to explore the changes in the most recent version: watersurfaces_2026 (DOI 10.5281/zenodo.22671253).

```{r}
filepath <- file.path(locate_n2khab_data(), "10_raw/watersurfaces/watersurfaces.gpkg")
```

Checksums:

```{r}
filepath %>% 
  n2khab::md5sum() %>% 
  `names<-`("md5sum")
filepath %>% 
  n2khab::sha256sum() %>% 
  `names<-`("sha256sum")
filepath %>% 
  n2khab::xxh64sum() %>% 
  `names<-`("xxh64sum")
```

Layers:

```{r, paged.print = FALSE}
st_layers(filepath)
```

Read the spatial layer:

```{r spatial-layer, paged.print = FALSE}
(ws <- read_sf(filepath, stringsAsFactors = TRUE, layer = "Watervlakken"))
```

The number of rows is `r nrow(ws)`.

Other (aspatial) layers were available in version 1.1, and again in this version.

## Step-by-step exploration

### Encoding problems (Windows only)

This has been explored for version 1.1, and handled where needed by `n2khab::read_watersurfaces()`.
Given the factor levels explored below, it does not appear that more changes are needed.
Hence this exploratory code has been dropped (2023-11-09).

### A summary of the spatial layer

```{r}
ws %>% 
  st_drop_geometry() %>% 
  summary()
```

Let us look for a few typical errors more systematically.

### Are there `NA` values?

There are no `NA` values at all but ... see below! 

```{r}
sapply(ws, function(x) sum(is.na(x)))
```

### Are there `"NA"` values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "NA", na.rm = TRUE))
```

### Are there `<Null>` values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '<Null>', na.rm = TRUE))
```

### Are there any "/" values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "/", na.rm = TRUE))
```

### Are there any "-" values?

Yes, in `KRWTYPEA`, but it's expected (described in the metadata):

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "-", na.rm = TRUE))
```

### Are there empty (`""`) values?

Yes, plenty of them. 
The empty fields are actually coded as an empty string ("")
(or more accurately, since all those fields are factors: as a factor level "") 

The "" are in fields where `NA` can be expected.

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "", na.rm = TRUE))
```

### Are there Zero (0) values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '0', na.rm = TRUE))
```

### Are WVLC codes unique?

Yes:

```{r}
ws$WVLC %>% unique() %>% length() == nrow(ws)
```

### Levels for each factor

We can compare the levels for each factor variable with the information given in 
[Leyssen et al. (2026)](https://pureportal.inbo.be/nl/publications/watervlakken-2026-polygonenkaart-van-stilstaand-water-in-vlaander).

We do not check `NAAM` and `GEBIED` since there are many possible options.

**WVLC**:

We check whether the structure of `WVLC` is correct.

- Does `WVLC`  start with a province code?

```{r}
substr(levels(ws$WVLC),1 ,3) %>% unique()
```

- How many records with an unexpected length? (not 10 characters)?

```{r}
ws %>% filter(str_length(WVLC) != 10) %>% nrow()
```

**KRWTYPE**: the codes in the spatial layer comply with the metadata report, except for the empty string.

```{r}
levels(ws$KRWTYPE)
```

**KRWTYPEA** (alternative type): the codes in the spatial layer comply with the metadata report, except for the empty string.

```{r}
levels(ws$KRWTYPEA)
```

**KRWTYPES** (status): the codes in the spatial layer are the same as in the metadata report, except for the empty string.

```{r}
levels(ws$KRWTYPES)
```

`KRWTYPE` / `KRWTYPEA` / `KRWTYPES` consistency:

```{r}
ws |> 
  st_drop_geometry() |> 
  count(KRWTYPE, KRWTYPES) |> 
  print()
ws |> 
  st_drop_geometry() |> 
  filter(!is.na(KRWTYPEA), KRWTYPEA != "-") |> 
  count(KRWTYPE, KRWTYPEA) |> 
  print(n = Inf)
```

One record has `KRWTYPES` set to 'voorlopig':

```{r}
ws |> filter(KRWTYPES == "voorlopig") |> as.data.frame()
```

After discussion with the authors, this seems to be correct.

**DIEPKL**: the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$DIEPKL)
```

Notes:

- in version 1.1, an R chunk in the 'Windows encoding' section already reduced the number of levels, which resulted in fewer extra categories.
However this was just for exploratory purposes as it was not the purpose to always implement such changes.
- also in the future we will not rectify this (by default) in reading functions for raw data sources: problems in the data will be returned as-is and should be solved in a future version of the data source.
By default we just streamline column names and variable types, we make sure that values referring to `NA` are effectively returned as `NA` and we try to avoid some encoding problems.

Because on Windows the ≥/\u2265 character is not well displayed, we had to recode it into '=>'
(otherwise "≥" in ws$DIEPKL was rendered as "=" in the html output). 
Note that this is not needed anymore since version 2024, which does not contain the character ≥ anymore, 
but we will keep this in the read function for the previous versions.

```{r}
levels(ws$DIEPKL) <- gsub(pattern = "\u2265", ">=", levels(ws$DIEPKL))
levels(ws$DIEPKL)
```

**CONNECT**: the codes in the spatial layer are the same as in the metadata report, except for the empty string.

```{r}
levels(ws$CONNECT)
```

**FUNCTIE**: the codes in the spatial layer comply with the metadata report, except for the empty string.

```{r}
levels(ws$FUNCTIE)
```

Here are the categories in the metadata report for `FUNCTIE`:

functie                       toewijzing
----------------------        ---------------------
natuur                        doelstelling natuurbehoud
hengelIntensief               intensief hengelen (met infrastructuur, bepoting of gebruikt voor wedstrijdhengelen)
hengelextensief               extensief hengelen (geen infrastructuur, bepoting of wedstrijdhengelen)
jacht                         jagen
tuin_park                     esthetisch (verblijfsrecreatie, tuin- en parkvijvers)
vogel                         waterpartij voor gedomesticeerde watervogels
viskweek                      opkweken van vis
zwemmen                       zwemmen
duik                          duiken
zachteRecreatie               niet gemotoriseerde waterrecreatie
motorrecreatie                gemotoriseerde waterrecreatie
waterberging                  waterberging ten behoeve van overstromings- of peilbeheer
opslag                        reservoir voor water (industrie, landbouw, bluswater, waterkracht…)
drinkwater                    drinkwaterwinning
zuivering                     (kleinschalige) waterzuivering, infiltratie
bezinking                     bezinking van proceswater
veedrenk                      watervoorziening voor vee
geen                          geen specifieke functie

**PEILBEHEER**: the codes in the spatial layer comply with the metadata report, except for the empty string.

```{r}
levels(ws$PEILBEHEER)
```


## Potential issues

- All the missing values for the factors are coded as level empty string "".

## Validity of the geometries

Let's inspect features with invalid or corrupt geometry:

```{r}
validities <- st_is_valid(ws)
validities %>% table
invalid_geoms <- ws[!validities | is.na(validities), ]
st_is_valid(invalid_geoms, reason = TRUE)
```

```{r}
tm_shape(invalid_geoms) + tm_borders() + tm_facets(by = "OBJECTID")
```

The geometry invalidity is the consequence of self-intersecting rings, as a consequence of digitalization errors.

Let's compare with the same geoms after fixing the self-intersecting rings:

```{r}
valid_geoms <- st_make_valid(invalid_geoms)
tm_shape(valid_geoms) + tm_borders() + tm_facets(by = "OBJECTID")
```

Are all geometries valid now?

```{r}
all(st_is_valid(valid_geoms))
```

So this works well; in derived data we will fix these geometries.
We can use the optional geometry reparation step in `read_watersurfaces()`.

We also check that no empty geometries are present:

```{r}
all(!is.na(st_dimension(ws$geom)))
```


Refer to <https://github.com/inbo/n2khab-preprocessing/issues/60> and <https://r-spatial.org/r/2017/03/19/invalid.html> for more information!

## Let's plot the watersurfaces as a map

```{r plot}

# plot watersurfaces v2024
 p <- ggplot() +
  geom_sf(data = ws, aes(), color = "blue")

# Flanders
sf_vl <- read_admin_areas()

p <- p + 
  geom_sf(data = sf_vl, fill = NA)

print(p)

```

<!-- ### Are all the watersurfaces in Flanders?  -->

<!-- ```{r streams-in-fl} -->
<!-- in_vl <- st_contains(x = sf_vl, y = ws) -->

<!-- not_contained <- ws %>% -->
<!--   st_drop_geometry() %>% -->
<!--   mutate(orig_rowname = rownames(.)) %>% -->
<!--   filter(!rownames(.) %in% in_vl[[1]]) -->

<!-- # 62 watersurfaces -->
<!-- print(not_contained %>% nrow()) -->

<!-- not_contained %>% -->
<!--   distinct(WVLC, NAAM) %>% -->
<!--   kable() -->
<!-- ``` -->

<!-- 62 watersurfaces are not 'contained' within the polygon for Flanders (see list above) -->

# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
ws %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`WVLC`                        `polygon_id` 
`WTRLICHC`                    `wfd_code`
`NAAM`                        `name`
`GEBIED`                      `area_name`
`KRWTYPE`                     `wfd_type `
`KRWTYPEA`                    `wfd_type_alternative`
`KRWTYPES`                    `wfd_type_certain`
`DIEPKL`                      `depth_class`
`CONNECT`                     `connectivity`
`FUNCTIE`                     `usage`
`PEILBEHEER`                  `water_level_management`

# Other considerations for the R object returned by `read_watersurfaces()`

- not uptaking `OPPWVL`, `OMTWVL`, `SHAPE_Length`, `SHAPE_Area` (area & perimeter are easily calculated etc) -- OK
- sort by `polygon_id` -- OK
- add translations to long text for `wfd_type `, `wfd_type_alternative ` and `connectivity`  -- OK but not by default
- add translations to long text for  `usage`? -- for a later version (as more 
codes will be used)
- converting null / 0 / "" values to `NA` -- OK
- support variable `water_level_management` (since 1.2) – OK
- support variable `wfd_type_alternative` (since v2024) – OK
- make variable `hyla_code` optional (dropped in v2024) – OK
- apply a natural order to the `depth_class` levels

# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
if ("sf" %in% si$packages$package) {
  p2 <- sf::sf_extSoftVersion()
  p2 <- p2[names(p2) != "proj.4"]
  names(p2) <- paste("sf", names(p2))
  p <- c(p, p2)
}
if ("terra" %in% si$packages$package) {
  p2 <- terra::gdal(lib = "all")
  names(p2) <- paste("terra", names(p2) %>% toupper())
  p <- c(p, p2)
}
if ("rgrass" %in% si$packages$package) {
  p <- c(p, GRASS = link2GI::findGRASS()[1, "version"])
}
sprintf("- **%s**: %s\n", names(p), p) %>%
  cat(sep = "")
```

```{r results = "asis", echo=FALSE}
si$packages %>%
  as_tibble %>%
  select(package, loadedversion, date, source) %>%
  pander::pandoc.table(caption = "Loaded R packages",
                     split.table = Inf)
```
