---
title: "Handling the watersurfaces data source"
date: "`r paste('Version',lubridate::now())`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    number_sections: yes
    code_folding: hide
    includes:
      in_header: ../header.html
---

```{r setup, message=FALSE, echo=FALSE}
options(stringsAsFactors = FALSE)
library(sf)
library(tidyverse)
library(tmap)
library(n2khab)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```

# Exploring the watersurfaces data source

The previous versions were 1.0 (10.5281/zenodo.3386859), 1.1 (10.5281/zenodo.4117543)
and 1.2 (10.5281/zenodo.7440931)

We want to explore the changes in the most recent version: 2024 (10.5281/zenodo.14203168)

```{r v2024}
filepath <- file.path(fileman_up("n2khab_data"), "10_raw/watersurfaces/watersurfaces.gpkg")
```

Checksums:

```{r}
filepath %>% 
  n2khab::md5sum() %>% 
  `names<-`("md5sum")
filepath %>% 
  n2khab::sha256sum() %>% 
  `names<-`("sha256sum")
filepath %>% 
  n2khab::xxh64sum() %>% 
  `names<-`("xxh64sum")
```

Import the spatial layer:

```{r}
st_layers(filepath)
```

```{r spatial-layer-v2024}
(ws <- read_sf(filepath, stringsAsFactors = TRUE, layer = "Watervlakken"))
```

The number of rows is `r nrow(ws)`.

Other (aspatial) layers were only available in version 1.1.

## Step-by-step exploration

### Encoding problems (Windows only)

This has been explored for version 1.1, and handled where needed by `n2khab::read_watersurfaces()`.
Given the factor levels explored below, it does not appear that more changes are needed.
Hence this exploratory code has been dropped (2023-11-09).

### A summary of the spatial layer for version 2024

```{r}
ws %>% 
  st_drop_geometry %>% 
  summary
```

`PEILBEHEER` was a new column in version 1.2 and is still present in version 2024.
The column HYLAC was dropped in version 2024 (since the Hyla-codes are not used anymore by Natuurpunt)

Let us look for a few typical errors more systematically in version 2024

### Are there `NA` values?

There are no `NA` values at all in version 2024 but ... see below! 


```{r}
sapply(ws, function(x) sum(is.na(x)))
```

### Are there `""` values?

Yes, plenty of them. 
The empty fields are actually coded as an empty string ("")
(or more accurately, since all those fields are factors: as a factor level  "") 

The "" are in fields where `NA` can be expected.

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '', na.rm = TRUE))
```

### Are there `<Null>` values?

None: 

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '<Null>', na.rm = TRUE))
```

### Are there Zero (0) values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '0', na.rm = TRUE))
```

### Are there any "/" values?

None 

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "/", na.rm = TRUE))
```


### Are there any "-" values?

Yes there are "-" in `KRWTYPEA`, but it is expected (described in the metadata). 

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "-", na.rm = TRUE))
```

### Are WVLC codes unique?

Yes. It was not the case in version 1.0, but it has been corrected since v1.1.

```{r}
ws$WVLC %>% unique %>% length == nrow(ws)
```


### Levels for each factor

We can compare the levels for each factor variable with the information given in 
[Leyssen et al. (2024)](https://pureportal.inbo.be/nl/publications/watervlakken-2024-polygonenkaart-van-stilstaand-water-in-vlaander).

We do not check `NAAM` and `GEBIED` since there are many possible options.

**WVLC** 

Similar problem for the `WVLC` but we can check whether the structure of the codes is correct:

Does `WVLC`  start with a province code?

```{r}
substr(levels(ws$WVLC),1 ,3) %>% unique()
```

How many records with an unexpected length? (not 10 characters)?

```{r}
ws %>% filter(str_length(WVLC) != 10) %>% nrow()
```

There is one problem: 

```{r}
ws |> filter(str_detect(WVLC, pattern = "d")) |> as.matrix() |> t()
```

**KRWTYPE**: the codes are correct but there are more codes mentioned in the metadata report.

```{r}
levels(ws$KRWTYPE)
```
**KRWTYPEA**: the codes are correct but there are more codes mentioned in the metadata report.

```{r}
levels(ws$KRWTYPEA)
```


**KRWTYPES** (status): the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$KRWTYPES)
```

`KRWTYPE` / `KRWTYPEA` / `KRWTYPES` consistency:

```{r}
ws |> 
  st_drop_geometry() |> 
  count(KRWTYPE, KRWTYPES) |> 
  print()
```

```{r}
ws |> 
  st_drop_geometry() |> 
  filter(!is.na(KRWTYPEA), KRWTYPEA != "-") |> 
  count(KRWTYPE, KRWTYPEA) |> 
  print(n = Inf)
```

One record has `KRWTYPES` set to ‘voorlopig’:

```{r}
ws |> filter(KRWTYPES == "voorlopig") |> as.matrix() |> t()
```

After discussion with the authors, this seems to be correct.

**DIEPKL**: the codes in the spatial layer are the same as in the metadata report but the ordering of the levels could be more logical.

```{r}
levels(ws$DIEPKL)
```

Notes:

- in version 1.1, an R chunk in the 'Windows encoding' section already reduced the number of levels, which resulted in fewer extra categories.
However this was just for exploratory purposes as it was not the purpose to always implement such changes.
- also in the future we will not rectify this (by default) in reading functions for raw data sources: problems in the data will be returned as-is and should be solved in a future version of the data source.
By default we just streamline column names and variable types, we make sure that values referring to `NA` are effectively returned as `NA` and we try to avoid some encoding problems.

Because on Windows the ≥/\u2265 character is not well displayed, we had to recode it into '=>'
(otherwise "≥" in ws$DIEPKL was rendered as "=" in the html output). 
Note that this is not needed anymore for version 2024 that does not contain the character ≥ anymore 
but we will keep this in the read fucntion for the previous versions

```{r}
levels(ws$DIEPKL) <- gsub(pattern = "\u2265", ">=", levels(ws$DIEPKL))
levels(ws$DIEPKL)
```

**CONNECT**: the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$CONNECT)
```

**FUNCTIE**: there are many more codes in the report than in the dataset

```{r}
levels(ws$FUNCTIE)
```

Here are the categories in the metadata report for `FUNCTIE`:

functie                       toewijzing
----------------------        ---------------------
natuur                        doelstelling natuurbehoud
hengelintensief               intensief hengelen (met infrastructuur, bepoting of gebruikt voor wedstrijdhengelen)
hengelextensief               extensief hengelen (geen infrastructuur, bepoting of wedstrijdhengelen)
jacht                         jagen
tuin_park                     esthetisch (verblijfsrecreatie, tuin- en parkvijvers)
vogel                         waterpartij voor gedomesticeerde watervogels
viskweek                      opkweken van vis
zwemmen                       zwemmen
duik                          duiken
zachteRecreatie               niet gemotoriseerde waterrecreatie
motorrecreatie                gemotoriseerde waterrecreatie
waterberging                  waterberging ten behoeve van overstromings- of peilbeheer
opslag                        reservoir voor water (industrie, landbouw, bluswater, waterkracht…)
drinkwater                    drinkwaterwinning
zuivering                     (kleinschalige) waterzuivering, infiltratie
bezinking                     bezinking van proceswater
veedrenk                      watervoorziening voor vee
geen                          geen specifieke functie

**PEILBEHEER**: one code less than in the report.

```{r}
levels(ws$PEILBEHEER)
```


## Potential issues

- there is a new field: `KRWTYPEA` and one field was dropped `HYLAC`
- `KRWTYPE`, `KRWTYPEA`, `FUNCTIE`, `PEILBEHEER`: more levels in the metadata report than in the dataset.
This is not necessarily a problem.
- All the missing values for the factors are coded as level empty string ""  
- One wrong WVLC code ("d") in de Kleiputten van Heist (Palingpot)


## Validity of the geometries

Let's inspect features with invalid or corrupt geometry:

```{r}
validities <- st_is_valid(ws)
validities %>% table
invalid_geoms <- ws[!validities | is.na(validities), ]
st_is_valid(invalid_geoms, reason = TRUE)
```

```{r}
tm_shape(invalid_geoms) + tm_borders() + tm_facets(by = "OBJECTID")
```

The geometry invalidity is the consequence of self-intersecting rings, as a consequence of digitalization errors.

Let's compare with the same geoms after fixing the self-intersecting rings:

```{r}
valid_geoms <- st_make_valid(invalid_geoms)
tm_shape(valid_geoms) + tm_borders() + tm_facets(by = "OBJECTID")
```

Are all geometries valid now?

```{r}
all(st_is_valid(valid_geoms))
```

So this works well; in derived data we will fix these geometries.
We can use the optional geometry reparation step in `read_watersurfaces()`.

We also check that no empty geometries are present:

```{r}
all(!is.na(st_dimension(ws$geom)))
```


Refer to <https://github.com/inbo/n2khab-preprocessing/issues/60> and <https://r-spatial.org/r/2017/03/19/invalid.html> for more information!

## Let's plot the watersurfaces as a map

```{r plot}

# plot watersurfaces v2024
 p <- ggplot() +
  geom_sf(data = ws, aes(), color = "blue")

# Flanders
sf_vl <- read_admin_areas()

p <- p + 
  geom_sf(data = sf_vl, fill = NA)

print(p)

```

<!-- ### Are all the watersurfaces in Flanders?  -->

<!-- ```{r streams-in-fl} -->
<!-- in_vl <- st_contains(x = sf_vl, y = ws) -->

<!-- not_contained <- ws %>% -->
<!--   st_drop_geometry() %>% -->
<!--   mutate(orig_rowname = rownames(.)) %>% -->
<!--   filter(!rownames(.) %in% in_vl[[1]]) -->

<!-- # 62 watersurfaces -->
<!-- print(not_contained %>% nrow()) -->

<!-- not_contained %>% -->
<!--   distinct(WVLC, NAAM) %>% -->
<!--   kable() -->
<!-- ``` -->

<!-- 62 watersurfaces are not 'contained' within the polygon for Flanders (see list above) -->

# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
ws %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`WVLC`                        `polygon_id` 
`WTRLICHC`                    `wfd_code`
`NAAM`                        `name`
`GEBIED`                      `area_name`
`KRWTYPE`                     `wfd_type `
`KRWTYPEA`                    `wfd_type_alternative`
`KRWTYPES`                    `wfd_type_certain`
`DIEPKL`                      `depth_class`
`CONNECT`                     `connectivity`
`FUNCTIE`                     `usage`
`PEILBEHEER`                  `water_level_management`

# Other considerations for the R object returned by `read_watersurfaces()`

- not uptaking `OPPWVL`, `OMTWVL`, `SHAPE_Length`, `SHAPE_Area` (area & perimeter are easily calculated etc) -- OK
- sort by `polygon_id` -- OK
- add translations to long text for `wfd_type `, `wfd_type_alternative `, `connectivity`  -- OK but not by default
- add translations to long text for  `usage`? -- for a later version (as more 
codes will be used)
- converting null / 0 / "" values to `NA` -- OK
- support variable water_level_management (since 1.2) – OK
- support variable wfd_type_alternative (since v2024) – OK
- make variable hyla_code optional (dropped in v2024) – OK
- there is one wrong WVLC code ("d" should be "WVLKNO0072") in de Kleiputten van Heist (Palingpot) - should we stay true to source?  see <https://github.com/inbo/n2khab/pull/192>

# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
if ("sf" %in% si$packages$package) {
  p <- c(p, sf_extSoftVersion())
  names(p)[names(p) == "proj.4"] <- "PROJ"
}
if ("rgrass" %in% si$packages$package) {
  p <- c(p, GRASS = link2GI::findGRASS()[1, "version"])
}
sprintf("- **%s**: %s\n", names(p), p) %>%
  cat(sep = "")
```

```{r results = "asis", echo=FALSE}
si$packages %>%
    as_tibble %>%
    select(package, loadedversion, date, source) %>%
pander::pandoc.table(caption = "Loaded R packages",
                     split.table = Inf)
```




