---
title: "Inspecting an internal draft of the watersurfaces data source"
date: "`r paste('Version',lubridate::now())`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    number_sections: yes
    code_folding: show
    includes:
      in_header: ../header.html
---

```{r setup, message=FALSE, echo=FALSE}
library(sf)
library(dplyr)
library(tmap)
library(n2khab)
library(ggplot2)
library(knitr)
opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```

# Exploring an internal draft of the watersurfaces data source

Prior to publication, updates of the watersurfaces happen in a MS SQL Server database.

Here we do some checks from R in order to prevent possible problems in a released version of the data source.

Make the database connection and the queries:

```{r dev-after-v1.2}
con <- inbodb::connect_inbo_dbase("G0102_00_AquaMorf")
query <- sql("SELECT P.CodePlas AS WVLC, 
                  plasnaamalt AS WTRLICHC,
                  plasnaam AS NAAM, 
                  P.gebied AS GEBIED,
                  P.watertype AS KRWTYPE,
                  P.watertypeAlt AS KRWTYPEA,
                  P.statuswatertype AS KRWTYPES,
                  P.DiepteKlasse AS DIEPKL, 
                  P.connectiviteit11 AS CONNECT, 
                  P.peilbeheer AS PEILBEHEER,
                  shape.STArea() AS OPPWVL,
                  shape.STLength() AS OMTWVL,
                  P.DiepteGem_m AS diepte_gem, 
                  P.DiepteMax_m AS diepte_max, 
                  P.gebiedalternatief AS gebied_alt,
                  connectiviteit AS connect_old,
                  P.jaaraanleg AS aanleg,
                  P.status AS status, 
                  P.globalid AS id_plas,
                  shape.STAsBinary() AS geometry
          FROM AquaMorf.PLAS P
          WHERE GDB_TO_DATE >= '9999' AND status IS NULL")
# show all columns of the database table with:
# tbl(con, I("AquaMorf.PLAS")) |> glimpse()
query_functie <- sql("SELECT PG.gebruiksfunctie AS FUNCTIE,
                      PG.globalid AS id_functie, 
                      PG.CodePlas AS codeplas, 
                      PG.globalid_plas AS id_plas
                      FROM AquaMorf.PLAS_GEBRUIKSFUNCTIE PG
                      WHERE GDB_TO_DATE >= '9999'")
```

Note that we map variable names to uppercase column names from the last released version where possible.
Some other available names have been kept lowercase.

Import the spatial layer:

```{r spatial-layer}
ws <- read_sf(con, query = query, crs = 31370) |> 
  mutate(across(where(is.character), factor))
glimpse(ws)
```

The number of rows is `r nrow(ws)`.

Query the values of `FUNCTIE` that are available for specific watersurfaces:

```{r}
ws_functie <- tbl(con, query_functie) |> 
  collect() |> 
  mutate(across(where(is.character), factor))
glimpse(ws_functie)
```

The number of rows is `r nrow(ws_functie)`.

Sidenote: `read_sf()` downloads the data, but it is also possible to define a lazy object, execute database-level queries using dplyr verbs on this object (before collecting), and then convert to sf after collecting a filtered result.
Example below:

```{r paged.print=FALSE, collapse=TRUE}
ws_db <- tbl(con, query)
# without conversion to sf
ws_db |> 
  filter(row_number() <= 25) |> 
  select(WVLC, geometry) |> 
  collect()
# one can still convert the binary geometries to WKB; see
# https://hydroecology.net/reading-spatial-data-from-sql-server-without-sf/
ws_db |> 
  filter(row_number() <= 25) |> 
  select(WVLC, geometry) |> 
  collect() |> 
  mutate(geometry = wk::as_wkb(geometry))
# conversion to sf:
ws_db |> 
  filter(row_number() <= 25) |> 
  select(WVLC, geometry) |> 
  collect() |>
  st_as_sf(crs = 31370)
```

```{r}
DBI::dbDisconnect(con)
```


## Step-by-step exploration

### A summary of the spatial layer

```{r}
ws %>% 
  st_drop_geometry %>% 
  summary
```

Compared to version 2024:

- `FUNCTIE` is absent from the query (it has to be retrieved via a separate query).
- Several columns are available (lowercase names in the query) which haven't been part of the official version yet (except for `connect_old`, which was `CONNECT` in older versions).

Let us look for a few typical errors more systematically.

### Are there `NA` values?

There are plenty of `NA`'s but only in fields where we expect them.

```{r}
sapply(ws, function(x) sum(is.na(x)))
```

### Are there `"NA"` values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "NA", na.rm = TRUE))
```

### Are there `<Null>` values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '<Null>', na.rm = TRUE))
```

### Are there any "/" values?

None :

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "/", na.rm = TRUE))
```

### Are there `-` values?

Yes, in `KRWTYPEA`, but it's expected (described in the metadata):

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '-', na.rm = TRUE))
```

### Are there empty (`""`) values?

Yes, in `KRWTYPEA`, see further:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == "", na.rm = TRUE))
```

### Are there Zero (0) values?

None:

```{r}
sapply(ws |> st_drop_geometry(), function(x) sum(as.character(x) == '0', na.rm = TRUE))
```

### Are WVLC codes unique?

Yes:

```{r}
ws$WVLC %>% unique() %>% length() == nrow(ws)
```

<!-- Which codes are duplicated? -->

<!-- ```{r} -->
<!-- ws |>  -->
<!--   st_drop_geometry() |>  -->
<!--   count(WVLC) |>  -->
<!--   filter(n > 1) -->
<!-- ``` -->

### Levels for each factor

We can compare the levels for each factor variable with the information given in 
a draft metadata report.

We do not check `NAAM` and `GEBIED` since there are many possible options.

**KRWTYPE**: the codes are correct but there are more codes mentioned in the metadata report.

```{r}
levels(ws$KRWTYPE)
```

**KRWTYPEA** (alternative type): the codes in the spatial layer comply with the metadata report, except for the empty string.

```{r}
levels(ws$KRWTYPEA)
```

**KRWTYPES** (status): the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$KRWTYPES)
```

`KRWTYPE` / `KRWTYPEA` / `KRWTYPES` consistency:

```{r}
ws |> 
  st_drop_geometry() |> 
  count(KRWTYPE, KRWTYPES) |> 
  print()
ws |> 
  st_drop_geometry() |> 
  filter(!is.na(KRWTYPEA), KRWTYPEA != "-") |> 
  count(KRWTYPE, KRWTYPEA) |> 
  print(n = Inf)
```

Note that one record has `KRWTYPEA` equal to `""` while it should be a missing value:

```{r}
ws |> filter(KRWTYPEA == "") |> as.data.frame()
```

One record has `KRWTYPES` set to 'voorlopig':

```{r}
ws |> filter(KRWTYPES == "voorlopig") |> as.data.frame()
```

After discussion with the authors, this seems to be correct.

**DIEPKL**: the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$DIEPKL)
```

Notes:

- in version 1.1, an R chunk in the 'Windows encoding' section already reduced the number of levels, which resulted in fewer extra categories.
However this was just for exploratory purposes as it was not the purpose to always implement such changes.
- also in the future we will not rectify this (by default) in reading functions for raw data sources: problems in the data will be returned as-is and should be solved in a future version of the data source.
By default we just streamline column names and variable types, we make sure that values referring to `NA` are effectively returned as `NA` and we try to avoid some encoding problems.

Because on Windows the ≥/\u2265 character is not well displayed, we had to recode it into '=>'
(otherwise "≥" in ws$DIEPKL was rendered as "=" in the html output). 
Note that this is not needed anymore since version 2024, which does not contain the character ≥ anymore, 
but we will keep this in the read function for the previous versions.

```{r}
levels(ws$DIEPKL) <- gsub(pattern = "\u2265", ">=", levels(ws$DIEPKL))
levels(ws$DIEPKL)
```

**CONNECT**: the codes in the spatial layer are the same as in the metadata report.

```{r}
levels(ws$CONNECT)
```

**FUNCTIE**: the codes are correct but there are more codes mentioned in the metadata report.

```{r}
levels(ws_functie$FUNCTIE)
```

Here are the categories in the metadata report for `FUNCTIE`:

functie                       toewijzing
----------------------        ---------------------
natuur                        doelstelling natuurbehoud
hengelIntensief               intensief hengelen (met infrastructuur, bepoting of gebruikt voor wedstrijdhengelen)
hengelextensief               extensief hengelen (geen infrastructuur, bepoting of wedstrijdhengelen)
jacht                         jagen
tuin_park                     esthetisch (verblijfsrecreatie, tuin- en parkvijvers)
vogel                         waterpartij voor gedomesticeerde watervogels
viskweek                      opkweken van vis
zwemmen                       zwemmen
duik                          duiken
zachteRecreatie               niet gemotoriseerde waterrecreatie
motorrecreatie                gemotoriseerde waterrecreatie
waterberging                       waterberging ten behoeve van overstromings- of peilbeheer
opslag                        reservoir voor water (industrie, landbouw, bluswater, waterkracht…)
drinkwater                    drinkwaterwinning
zuivering                     (kleinschalige) waterzuivering, infiltratie
bezinking                     bezinking van proceswater
veedrenk                      watervoorziening voor vee
geen                          geen specifieke functie

**PEILBEHEER**: the codes are correct but there is one code less than in the report.

```{r}
levels(ws$PEILBEHEER)
```

## Potential issues

- One record has an empty string for `KRWTYPEA`; given that `KRWTYPE` is a missing value this should also be the case for `KRWTYPEA`.
- `KRWTYPE`, `PEILBEHEER`, `FUNCTIE`: more levels in the metadata report than in the dataset.
This is not necessarily a problem.

These will be reported to the authors of the layers.


## Validity of the geometries

Let's inspect features with invalid or corrupt geometry:

```{r}
ws_validity <- st_is_valid(ws)
ws_validity %>% table
invalid_geoms <- ws[!ws_validity | is.na(ws_validity), ]
```

Identifying the invalid geometries:

```{r}
invalid_geoms |> 
  select(WVLC) |> 
  mutate(reason = st_is_valid(invalid_geoms, reason = TRUE)) |> 
  st_drop_geometry()
```

```{r}
tm_shape(invalid_geoms) + tm_borders() + tm_facets(by = "WVLC")
```

The geometry invalidity is the consequence of self-intersecting rings, as a consequence of digitalization errors.

Let's compare with the same geoms after fixing the self-intersecting rings:

```{r}
valid_geoms <- st_make_valid(invalid_geoms)
tm_shape(valid_geoms) + tm_borders() + tm_facets(by = "WVLC")
```

Are all geometries valid now?

```{r}
all(st_is_valid(valid_geoms))
```

So this works well; in derived data we will fix these geometries.
Note that `n2khab::read_watersurfaces()` also has a `fix_geom` argument to apply this to the raw data source.

We also check that no empty geometries are present:

```{r}
all(!is.na(st_dimension(ws$geometry)))
```


Refer to <https://github.com/inbo/n2khab-preprocessing/issues/60> and <https://r-spatial.org/r/2017/03/19/invalid.html> for more information!

## Let's plot the watersurfaces as a map

```{r plot}

# plot watersurfaces
 p <- ggplot() +
  geom_sf(data = ws, aes(), color = "blue")

# Flanders
sf_vl <- read_admin_areas()

p <- p + 
  geom_sf(data = sf_vl, fill = NA)

print(p)

```

<!-- ### Are all the watersurfaces in Flanders? -->

<!-- ```{r streams-in-fl} -->
<!-- in_vl <- st_contains(x = sf_vl, y = ws) -->

<!-- not_contained <- ws %>%  -->
<!--   st_drop_geometry() %>%  -->
<!--   mutate(orig_rowname = rownames(.)) %>%  -->
<!--   filter(!rownames(.) %in% in_vl[[1]]) -->

<!-- # 61 watersurfaces -->
<!-- print(not_contained %>% nrow()) -->

<!-- not_contained %>%  -->
<!--   distinct(WVLC, NAAM) %>%  -->
<!--   kable() -->
<!-- ``` -->
<!-- 61 watersurfaces are not 'contained' within the polygon for Flanders (see list above) -->

# Tidyverse-styled, internationalized column names when using the data source in R

```{r eval = FALSE}
ws %>% colnames %>% cat(sep = "\n")
```

data source variable          data frame variable
----------------------        ---------------------
`WVLC`                        `polygon_id` 
`WTRLICHC`                    `wfd_code`
`NAAM`                        `name`
`GEBIED`                      `area_name`
`KRWTYPE`                     `wfd_type `
`KRWTYPEA`                    `wfd_type_alternative`
`KRWTYPES`                    `wfd_type_certain`
`DIEPKL`                      `depth_class`
`CONNECT`                     `connectivity`
`FUNCTIE`                     `usage`
`PEILBEHEER`                  `water_level_management`

# Other considerations for the R object returned by `read_watersurfaces()`

- not uptaking `OPPWVL`, `OMTWVL`, `SHAPE_Length`, `SHAPE_Area` (area & perimeter are easily calculated etc) -- OK
- sort by `polygon_id` -- OK
- add translations to long text for `wfd_type `, `wfd_type_alternative ` and `connectivity`  -- OK but not by default
- add translations to long text for  `usage`? -- for a later version (as more 
codes will be used)
- converting null / 0 / "" values to `NA` -- OK
- support variable `water_level_management` (since 1.2) – OK
- support variable `wfd_type_alternative` (since v2024) – OK
- make variable `hyla_code` optional (dropped in v2024) – OK
- apply a natural order to the `depth_class` levels

# Used environment

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
if ("sf" %in% si$packages$package) {
  p2 <- sf::sf_extSoftVersion()
  p2 <- p2[names(p2) != "proj.4"]
  names(p2) <- paste("sf", names(p2))
  p <- c(p, p2)
}
if ("terra" %in% si$packages$package) {
  p2 <- terra::gdal(lib = "all")
  names(p2) <- paste("terra", names(p2) %>% toupper())
  p <- c(p, p2)
}
if ("rgrass" %in% si$packages$package) {
  p <- c(p, GRASS = link2GI::findGRASS()[1, "version"])
}
sprintf("- **%s**: %s\n", names(p), p) %>%
  cat(sep = "")
```

```{r results = "asis", echo=FALSE}
si$packages %>%
    as_tibble %>%
    select(package, loadedversion, date, source) %>%
pander::pandoc.table(caption = "Loaded R packages",
                     split.table = Inf)
```




