# Generate map of watersurfaces that contain aquatic habitat

We will create a map with watersurfaces that contain aquatic habitat and rib. The structure of the map will be similar to processed habitatmap of Flanders: it will consist of spatial file that contains the polygons and a table with the different habitat 'patches' within the polygons.  

## Data sources

* The processed habitatmap of Flanders

* The water surfaces map of Flanders


```{r read processed habmap}

habmap <- read_habitatmap_stdized() 

```

The watersurface map does not have attribute that contains a unique id. We create one based on the WVLC attribute. Watersurfaces with an identical value WVLC get a suffix equal to the rank of the watersurface based on its area. 

```{r read watersurfaces map}

watersurfaces <- st_read(dsn = "../../n2khab_data/10_raw/watersurfaces",
                  layer = "watersurfaces",
                  crs = 31370)

unique_id <- nrow(watersurfaces) == n_distinct(watersurfaces$WVLC)
#FALSE

watersurfaces <- watersurfaces %>%
    group_by(WVLC) %>%
    mutate(n_id = n(),
           polygon_id_ws = ifelse(n_id > 1,
                                  str_c(as.character(WVLC), "_", rank(OPPWVL)),
                                  as.character(WVLC))
           ) %>%
    ungroup()

unique_id <- nrow(watersurfaces) == n_distinct(watersurfaces$polygon_ws_id)

```

## Habitatmap 2190_a

For habitat sutbtype 2190_a we select all water surfaces within habitat map polygons that contain coastal dune habitat types. In a next step, this selection will have to be manually screened, as some of the larger (artificial) water surfaces within dune habitat polygons might not correspond with the 2190_a subtype.

### Selection of dune habitat polygons

```{r select dune habitat polygons}

habitatmap_21xx_patches <- habmap$habitatmap_patches %>%
  filter((str_sub(type, 1, 2) == "21")) 

habitatmap_21xx_polygons<- habmap$habitatmap_polygons %>%
  filter(polygon_id %in% habitatmap_21xx_patches$polygon_id)

```

## Selection of water surfaces that overlap with dune habitat polygons in habitat map

```{r}

watersurfaces_21xx <- watersurfaces[habitatmap_21xx_polygons,]

```


## Create 2190_a map 

For each watersurface (that is situated in coastal dune habitat) we create following attributes:

* polygon_id_ws: the id of the watersurface
* polygon_id_habitatmap: the id of all habitatmap polygons that overlap with the watersurfaces. The different id's are separated with a '+'
* description_orig: all unique habitat descriptions (description_orig attribute in habitatmap_stdized) of habitatmap polygons that overlap with the watersurfaces. The different descriptions are separated with a '+'.
* polygon_id: we will use polygon_id_ws as polygon_id
* polygon_type: to indicate the source of the polygon, either the watersurface map or the habitatmap

To create the attributes 'polygon_id_habitatmap' and 'description_orig', we make a spatial intersection between the watersurfaces within  coastal dunes and the habitatmap. Next we summarise for each watersurface and paste the different polygon_id_habitatmap's and description_orig's.  

```{r}

intersection_ws_21xx <- watersurfaces_21xx %>%
    st_intersection(habitatmap_21xx_polygons) %>%
    select(polygon_id_ws, polygon_id, description_orig)

watersurfaces_2190_a_patches <- intersection_ws_21xx %>%
    st_drop_geometry() %>%
    # add 'description_orig' attribute
    left_join(habitatmap_21xx_patches, by = "polygon_id") %>% 
    rename(polygon_id_habitatmap = polygon_id) %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig) %>%
    mutate(type = "2190_a") %>%
    group_by(polygon_id_ws, type) %>%
    summarise(polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
           description_orig =  str_c(description_orig, collapse = "+")) %>%
    ungroup() 

# we will add the description_orig attribute to the object that contains the polygons (analogous to habitatmap_stdized)
description <- watersurfaces_2190_a_patches %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig)

watersurfaces_2190_a_polygons <- watersurfaces_21xx %>%
    left_join(description, by = "polygon_id_ws") %>%
    mutate(polygon_type = "watersurface map",
           polygon_id = polygon_id_ws) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, description_orig, polygon_type)

watersurfaces_2190_a_patches <- watersurfaces_2190_a_patches %>%
    select(-description_orig) %>%
    mutate(patch_id = 1,
           code_orig = NA,
           certain = TRUE,
           polygon_id = polygon_id_ws) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, patch_id, code_orig, certain, type)


```

## Create standing water (excl. 2190) map

```{r}

standing_water <- read_types() %>%
    filter(tag_1 == "SW") %>%
    filter(main_type != "2190")

habitatmap_standw_patches <- habmap$habitatmap_patches %>%
  filter(type %in% standing_water$type) 

habitatmap_standw_polygons <- habmap$habitatmap_polygons %>%
  filter(polygon_id %in% habitatmap_standw_patches$polygon_id)

watersurfaces_standw <- watersurfaces[habitatmap_standw_polygons,] 

check_id <- n_distinct(watersurfaces_standw$polygon_id_ws) == nrow(watersurfaces_standw)

```


### Intersection water surfaces and standing water habitat

We create the water surface map with standing water habitat in a very similar way than the 2190_a map.

For each water surface we will create a patch for every standing water habitat type that occurs in the water surface according to the habitat map. We do not attempt to calculate a phab for the different patches. 

```{r}

intersection_ws_standw <- watersurfaces_standw %>%
    st_intersection(habitatmap_standw_polygons) %>%
    select(polygon_id_ws, polygon_id, description_orig)

watersurfaces_standw_patches <- intersection_ws_standw %>%
    st_drop_geometry() %>%
    left_join(habitatmap_standw_patches, by = "polygon_id") %>%
    rename(polygon_id_habitatmap = polygon_id) %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig, code_orig, certain, type) %>%
    group_by(polygon_id_ws) %>%
    mutate(patch_id = rank(code_orig),
           polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
           description_orig =  str_c(description_orig, collapse = "+")
           ) %>%
    ungroup()

description <- watersurfaces_standw_patches %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig)

watersurfaces_standw_polygons <- watersurfaces_standw %>%
    # filter(polygon_id_ws %in% watersurfaces_standw_patches$polygon_id_ws) %>%
    left_join(description, by = "polygon_id_ws") %>%
    mutate(polygon_type = "watersurface map") %>%
    select(polygon_id_ws, polygon_id_habitatmap, description_orig, polygon_type)
    
watersurfaces_standw_patches <- watersurfaces_standw_patches %>%
    select(-description_orig)

```


### Standing water habitat that does not intersect with water surface map

We also select all habitat map polyons that do not intersect with the water surface map. 

```{r}

habitatmap_standw_polygons_no_ws <- habitatmap_standw_polygons %>%
    filter(!polygon_id %in% intersection_ws_standw$polygon_id) %>%
    rename(polygon_id_habitatmap = polygon_id, geometry = geom) %>%
    mutate(polygon_id_ws = NA,
           polygon_type = "habitatmap")

habitatmap_standw_patches_no_ws <- habitatmap_standw_patches %>%
    filter(!polygon_id %in% intersection_ws_standw$polygon_id) %>%
    mutate(polygon_id_habitatmap = polygon_id,
           polygon_id_ws = NA) %>%
    select(-phab)

```

```{r}
library(leaflet)

leaflet(habitatmap_standw_polygons_no_ws %>%
            st_transform(crs = 4326)) %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(popup = ~htmltools::htmlEscape(description_orig)) %>% 
  addLayersControl(
    baseGroups = c("OSM (default)", "Esri.WorldImagery"),
    options = layersControlOptions(collapsed = FALSE)) 



```




### Combine habitatmap polygons and water surface polygons

```{r}

habmap_standw_patches <- watersurfaces_standw_patches %>%
    bind_rows(habitatmap_standw_patches_no_ws) %>%
    mutate(polygon_id = ifelse(is.na(polygon_id_ws),
                               as.character(polygon_id_habitatmap),
                               as.character(polygon_id_ws))) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, patch_id, everything())

habmap_standw_polygons <- watersurfaces_standw_polygons %>%
    rbind(habitatmap_standw_polygons_no_ws) %>%
    mutate(polygon_id = ifelse(is.na(polygon_id_ws),
                               as.character(polygon_id_habitatmap),
                               as.character(polygon_id_ws))) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, everything())
    
```

## Combine 31xx and 2190_a map

```{r}
watersurfaces_hab_patches <- habmap_standw_patches %>%
    bind_rows(watersurfaces_2190_a_patches) 

watersurfaces_hab_polygons <- habmap_standw_polygons %>%
    rbind(habitatmap_2190_a_polygons) 
```


## Write results into a geopackage

```{r}

st_write(watersurfaces_hab_polygons, "../../n2khab_data/20_processed/watersurfaces_hab/watersurfaces_hab.gpkg", layer = "watersurfaces_hab_polygons", driver = "GPKG")

con = dbConnect(SQLite(),
                dbname = "../../n2khab_data/20_processed/watersurfaces_hab/watersurfaces_hab.gpkg")

dbWriteTable(con, "watersurfaces_hab_patches", watersurfaces_hab_patches)

dbDisconnect(con)

```
