# Generate map of watersurfaces that contain aquatic habitat

We will create a map with watersurfaces that contain aquatic habitat and rib. The structure of the map will be similar to the processed habitatmap of Flanders: it will consist of spatial file that contains the polygons and a table with the different types within the polygons.  

## Data sources

* The processed habitatmap of Flanders

* The water surfaces map of Flanders

To be sure we will use the correct version of the data sources (version 2020 for the processed habitatmap and version 1.1 for watersurfaces), we will first derive the xxh64 file hashes and compare them to the file hashes in the [data source version overview table](https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853) and [here](https://github.com/inbo/n2khab/pull/122#issuecomment-805940739)

### Processed habitatmap

Control the file hashes and read habitatmap_stdized.

```{r read processed habmap}
path <- fileman_up("n2khab_data")
file <- "20_processed/habitatmap_stdized"
filepath <- str_c(file.path(path, file), "/",
        # list.files(path = file.path(path, file), recursive = TRUE)
        setdiff(list.files(path = file.path(path, file)), 
                list.dirs(path = file.path(path, file), full.names = FALSE))
      )

hashes <-
    tibble(filepath) %>%
    mutate(filename = basename(filepath),
           xxh64 = map(filepath, function(x) {
                           x %>% xxh64sum() %>% str_c(collapse = '')
                         }) %>% as.character) %>% 
           mutate(xxh64_ref = c("3109c26f0a27a0f3"),
           match = xxh64 == xxh64_ref) %>%
    select(filename,
           xxh64,
           xxh64_ref,
           match)

kable(hashes) %>%
  kable_styling()

if (!all.equal(hashes$xxh64, hashes$xxh64_ref)) {
    stop(cat("The source map is NOT up to date ! Please check the datasource. "))
}

habmap <- read_habitatmap_stdized() 

```

### Watersurfaces map

Control the file hashes and read watersurfaces.

In previous versions (until v1.0) the watersurface map did not have an attribute that contained a unique id. In v1.1 the WVLC attribute is unique and can be used as identifier. 

```{r read watersurfaces map}

path <- fileman_up("n2khab_data")
file <- "10_raw/watersurfaces" 
filepath <- str_c(file.path(path, file), "/",
        # list.files(path = file.path(path, file), recursive = TRUE)
        setdiff(list.files(path = file.path(path, file)), 
                list.dirs(path = file.path(path, file), full.names = FALSE))
      )

hashes <-
    tibble(filepath) %>%
    mutate(filename = basename(filepath),
           xxh64 = map(filepath, function(x) {
                           x %>% xxh64sum() %>% str_c(collapse = '')
                         }) %>% as.character) %>% 
           mutate(xxh64_ref = c("8a21a68a3f4d225f"), # version 1.1
           match = xxh64 == xxh64_ref) %>%
    select(filename,
           xxh64,
           xxh64_ref,
           match)

kable(hashes) %>%
  kable_styling()

if (!all.equal(hashes$xxh64, hashes$xxh64_ref)) {
    stop(cat("The source map is NOT up to date ! Please check the datasource. "))
}

watersurfaces <- st_read(filepath,
                  layer = "Watervlakken",
                  crs = 31370)

unique_id <- nrow(watersurfaces) == n_distinct(watersurfaces$WVLC)
#TRUE
# we will call this id polygon_id_ws

watersurfaces <- watersurfaces %>% 
  rename(polygon_id_ws = WVLC)

```

## Habitatmap 2190_a

For habitat sutbtype 2190_a we select all water surfaces within habitat map polygons that contain coastal dune habitat types. In a next step, this selection will have to be manually screened, as some of the larger (artificial) water surfaces within dune habitat polygons might not correspond with the 2190_a subtype.

### Selection of dune habitat polygons

```{r select dune habitat polygons}

habitatmap_21xx_types <- habmap$habitatmap_types %>%
  filter((str_sub(type, 1, 2) == "21")) 

habitatmap_21xx_polygons <- habmap$habitatmap_polygons %>%
  filter(polygon_id %in% habitatmap_21xx_types$polygon_id) 
```

### Selection of water surfaces that overlap with dune habitat polygons in habitat map

```{r}

watersurfaces_21xx <- watersurfaces[habitatmap_21xx_polygons,]

```


### Create 2190_a map 

For each watersurface (that is situated in coastal dune habitat) we create following attributes:

* polygon_id_ws: the id of the watersurface
* polygon_id_habitatmap: the id of all habitatmap polygons that overlap with the watersurfaces. The different id's are separated with a '+'
* description_orig: all unique habitat descriptions (description_orig attribute in habitatmap_stdized) of habitatmap polygons that overlap with the watersurfaces. The different descriptions are separated with a '+'.
* polygon_id: we will use polygon_id_ws as polygon_id
* polygon_type: to indicate the source of the polygon, either the watersurface map or the habitatmap

To create the attributes 'polygon_id_habitatmap' and 'description_orig', we make a spatial intersection between the watersurfaces within  coastal dunes and the habitatmap. Next we summarise for each watersurface and paste the different polygon_id_habitatmap's and description_orig's.  

#### Tidying feature geometries

Unfortunately the intersection step would be hampered by a faulty geometry in habitatmap_21xx_polygons
"Error in CPL_geos_op2(op, x, y) : 
  Evaluation error: TopologyException: Input geom 1 is invalid: Ring Self-intersection at or near point 78082.684000000358 228483.7210999988 at 78082.684000000358 228483.7210999988.".

We solve the problem based on this [article: Tidying feature geometries with sf](https://www.r-spatial.org/r/2017/03/19/invalid.html).

```{r inspect geometry habmap, out.width='800px', out.height='500px'}

geom_validity_habmap <- 
  st_is_valid(habitatmap_21xx_polygons %>% 
                   st_transform(crs = 4326), 
              reason = TRUE) %>% 
  as_tibble() %>% 
  mutate(id = rownames(.)) %>% 
  left_join(habitatmap_21xx_polygons %>% 
              # rownames_to_column(var = "id") %>% 
              mutate(id = as.character(row_number())), 
            by = "id") %>% 
  filter(value != "Valid Geometry") %>% 
  mutate(lng_lat = str_extract(string = value, 
                                pattern = "(?<=\\[).*(?=\\])"),
         lng = str_extract(string = lng_lat, 
                                pattern = "^.*\\s") %>% as.double(),
         lat = str_extract(string = lng_lat, 
                                pattern = " .*") %>% as.double(),
         problem = str_extract(string = value, 
                                pattern = "[^\\[]+")) %>% 
  select(-lng_lat)

kable(geom_validity_habmap %>% 
        select(id, value, polygon_id, description_orig, geom)) %>% 
  kable_styling()

library(leaflet)

leaflet() %>%
  setView(lng = geom_validity_habmap[1,]$lng, lat = geom_validity_habmap[1,]$lat, 
          zoom = 9) %>% 
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = habitatmap_21xx_polygons %>%
            st_transform(crs = 4326),
            popup = paste("polygon_id:", habitatmap_21xx_polygons$polygon_id, "<br>",
                           "description_orig:", habitatmap_21xx_polygons$description_orig)) %>%
  addCircleMarkers(data = geom_validity_habmap, 
                   lng  = ~lng, lat = ~lat, 
                   color = "red",
                   label = paste(geom_validity_habmap$polygon_id,
                                 geom_validity_habmap$problem))
  
```

```{r correct geometry habmap, out.width='800px', out.height='500px'}

habitatmap_21xx_polygons <- st_make_valid(habitatmap_21xx_polygons)

st_is_valid(habitatmap_21xx_polygons, reason = TRUE) %>% 
  .[. != "Valid Geometry"]

# let's compare this map with the previous map: no difference at first sight
leaflet() %>%
  setView(lng = geom_validity_habmap[1,]$lng, lat = geom_validity_habmap[1,]$lat, 
          zoom = 9) %>% 
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = habitatmap_21xx_polygons %>%
            st_transform(crs = 4326),
            popup = paste("polygon_id:", habitatmap_21xx_polygons$polygon_id, "<br>",
                           "description_orig:", habitatmap_21xx_polygons$description_orig)) %>%
  addCircleMarkers(data = geom_validity_habmap, 
                   lng  = ~lng, lat = ~lat, 
                   color = "black",
                   label = paste(geom_validity_habmap$polygon_id,
                                 geom_validity_habmap$problem))

```

#### Create 2190_a map 

Now we can proceed to the intersect and add the attributes listed above.

```{r}
intersection_ws_21xx <- watersurfaces_21xx %>%
    st_intersection(habitatmap_21xx_polygons) %>%
    select(polygon_id_ws, polygon_id, description_orig)

watersurfaces_2190_a_types <- intersection_ws_21xx %>%
    st_drop_geometry() %>%
    # add 'description_orig' attribute
    left_join(habitatmap_21xx_types, by = "polygon_id") %>% 
    rename(polygon_id_habitatmap = polygon_id) %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig) %>%
    mutate(type = "2190_a") %>%
    group_by(polygon_id_ws, type) %>%
    summarise(polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
           description_orig =  str_c(description_orig, collapse = "+")) %>%
    ungroup() 

# we will add the description_orig attribute to the object that contains the polygons (analogous to habitatmap_stdized)
description <- watersurfaces_2190_a_types %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig)

watersurfaces_2190_a_polygons <- watersurfaces_21xx %>%
    left_join(description, by = "polygon_id_ws") %>%
    mutate(polygon_type = "watersurface map",
           polygon_id = polygon_id_ws) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, description_orig, polygon_type)

check_unique_polygon <- nrow(watersurfaces_2190_a_polygons) == n_distinct(watersurfaces_2190_a_polygons$geom)
check_unique_polygon

watersurfaces_2190_a_types <- watersurfaces_2190_a_types %>%
    select(-description_orig) %>%
    mutate(code_orig = NA,
           certain = TRUE,
           polygon_id = polygon_id_ws) %>%
    select(polygon_id, code_orig, certain, type)


```

## Create standing water (excl. 2190)

```{r}

standing_water <- read_types() %>%
    filter(tag_1 == "SW") %>%
    filter(main_type != "2190")

habitatmap_standw_types <- habmap$habitatmap_types %>%
  filter(type %in% standing_water$type) 

habitatmap_standw_polygons <- habmap$habitatmap_polygons %>%
  filter(polygon_id %in% habitatmap_standw_types$polygon_id)

watersurfaces_standw <- watersurfaces[habitatmap_standw_polygons,] 

check_id <- n_distinct(watersurfaces_standw$polygon_id_ws) == nrow(watersurfaces_standw)
check_id
```


### Intersection water surfaces and standing water habitat

We create the water surface map with standing water habitat in a very similar way than the 2190_a map.

For each water surface we will create a record for every standing water habitat type that occurs in the water surface according to the habitat map. We do not attempt to calculate a phab for the different types.

Once again, the intersection step is hampered by a faulty geometry, this time in watersurfaces_standw. 


#### Tidying feature geometries

```{r inspect geometry, out.width='800px', out.height='500px'}
geom_validity_ws_standw <- 
  st_is_valid(watersurfaces_standw %>% 
                   st_transform(crs = 4326), 
              reason = TRUE) %>% 
  as_tibble() %>% 
  mutate(id = rownames(.)) %>% 
  left_join(watersurfaces_standw %>% 
              # rownames_to_column(var = "id") %>% 
              mutate(id = as.character(row_number())), 
                     by = "id") %>% 
  filter(value != "Valid Geometry") %>% 
  mutate(lng_lat = str_extract(string = value, 
                                pattern = "(?<=\\[).*(?=\\])"),
         lng = str_extract(string = lng_lat, 
                                pattern = "^.*\\s") %>% as.double(),
         lat = str_extract(string = lng_lat, 
                                pattern = " .*") %>% as.double(),
         problem = str_extract(string = value, 
                                pattern = "[^\\[]+")) %>% 
  select(-lng_lat)

kable(geom_validity_ws_standw %>% 
        select(id, value, polygon_id_ws, geom)) %>% 
  kable_styling()

leaflet() %>%
  setView(lng = geom_validity_ws_standw[1,]$lng, lat = geom_validity_ws_standw[1,]$lat, 
          zoom = 9) %>% 
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = watersurfaces_standw %>%
            st_transform(crs = 4326),
            popup = paste("polygon_id_ws:", watersurfaces_standw$polygon_id_ws)) %>%
  addCircleMarkers(data = geom_validity_ws_standw, 
                   lng  = ~lng, lat = ~lat, 
                   color = "red",
                   label = paste(geom_validity_ws_standw$polygon_id_ws,
                                 geom_validity_ws_standw$problem))
```

```{r correct geometry, out.width='800px', out.height='500px'}

watersurfaces_standw <- st_make_valid(watersurfaces_standw)

st_is_valid(watersurfaces_standw, reason = TRUE) %>% 
  .[. != "Valid Geometry"]

# let's compare this map with the previous map: no difference at first sight
leaflet() %>%
  setView(lng = geom_validity_ws_standw[1,]$lng, lat = geom_validity_ws_standw[1,]$lat, 
          zoom = 9) %>% 
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = watersurfaces_standw %>%
            st_transform(crs = 4326),
            popup = paste("polygon_id_ws:", watersurfaces_standw$polygon_id_ws)) %>%
  addCircleMarkers(data = geom_validity_ws_standw, 
                   lng  = ~lng, lat = ~lat, 
                   color = "black",
                   label = paste(geom_validity_ws_standw$polygon_id_ws,
                                 geom_validity_ws_standw$problem))
```

#### Create map 

Now we can proceed to the intersect and add the needed attributes.

```{r}
intersection_ws_standw <- watersurfaces_standw %>%
    st_intersection(habitatmap_standw_polygons) %>%
    select(polygon_id_ws, polygon_id, description_orig)

watersurfaces_standw_types <- intersection_ws_standw %>%
    st_drop_geometry() %>%
    left_join(habitatmap_standw_types, by = "polygon_id") %>%
    rename(polygon_id_habitatmap = polygon_id) %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig, code_orig, certain, type) %>%
    group_by(polygon_id_ws, code_orig, certain, type) %>%
    summarise(polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
           description_orig =  str_c(description_orig, collapse = "+")) %>%
    ungroup() %>%
    group_by(polygon_id_ws) %>%
    mutate(polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
           description_orig =  str_c(description_orig, collapse = "+")) %>%
    ungroup()

description <- watersurfaces_standw_types %>%
    distinct(polygon_id_ws, polygon_id_habitatmap, description_orig)

watersurfaces_standw_polygons <- watersurfaces_standw %>%
    # filter(polygon_id_ws %in% watersurfaces_standw_types$polygon_id_ws) %>%
    left_join(description, by = "polygon_id_ws") %>%
    mutate(polygon_type = "watersurface map") %>%
    select(polygon_id_ws, polygon_id_habitatmap, description_orig, polygon_type)

check_unique_polygon <- nrow(watersurfaces_standw_polygons) == n_distinct(watersurfaces_standw_polygons$geom)
check_unique_polygon
    
watersurfaces_standw_types <- watersurfaces_standw_types %>%
    select(-description_orig)


```


### Standing water habitat that does not intersect with the water surface map

We also select all habitat map polygons that do not intersect with the water surface map (see map below). 

```{r}

habitatmap_standw_polygons_no_ws <- habitatmap_standw_polygons %>%
    filter(!polygon_id %in% intersection_ws_standw$polygon_id) %>%
    rename(polygon_id_habitatmap = polygon_id) %>%
    mutate(polygon_id_ws = NA,
           polygon_type = "habitatmap")

habitatmap_standw_types_no_ws <- habitatmap_standw_types %>%
    filter(!polygon_id %in% intersection_ws_standw$polygon_id) %>%
    mutate(polygon_id_habitatmap = polygon_id,
           polygon_id_ws = NA) %>%
    select(-phab)

```



```{r, out.width='800px', out.height='500px'}

leaflet(habitatmap_standw_polygons_no_ws %>%
            st_transform(crs = 4326)) %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(popup = ~htmltools::htmlEscape(description_orig)) %>% 
  addLayersControl(
    baseGroups = c("OSM (default)", "Esri.WorldImagery"),
    options = layersControlOptions(collapsed = FALSE)) 

```




### Combine habitatmap polygons and water surface polygons

```{r}

habmap_standw_types <- watersurfaces_standw_types %>%
    bind_rows(habitatmap_standw_types_no_ws) %>%
    mutate(polygon_id = ifelse(is.na(polygon_id_ws),
                               as.character(polygon_id_habitatmap),
                               as.character(polygon_id_ws))) %>%
    select(polygon_id, everything())

habmap_standw_polygons <- watersurfaces_standw_polygons %>%
    rbind(habitatmap_standw_polygons_no_ws) %>%
    mutate(polygon_id = ifelse(is.na(polygon_id_ws),
                               as.character(polygon_id_habitatmap),
                               as.character(polygon_id_ws))) %>%
    select(polygon_id, polygon_id_ws, polygon_id_habitatmap, everything())
    
```

## Combine 31xx and 2190_a map

```{r}
watersurfaces_hab_types <- habmap_standw_types %>%
    bind_rows(watersurfaces_2190_a_types) %>%
    select(polygon_id, code_orig, certain, type)

watersurfaces_hab_polygons <- habmap_standw_polygons %>%
    rbind(watersurfaces_2190_a_polygons) %>%
    group_by(polygon_id, polygon_id_ws) %>%
    summarise(polygon_id_habitatmap = str_c(polygon_id_habitatmap, collapse = "+"),
              description_orig = str_c(description_orig, collapse = "+")
              ) %>%
    ungroup() %>%
    arrange(polygon_id_ws)

check_unique_id <- nrow(watersurfaces_hab_polygons) == n_distinct(watersurfaces_hab_polygons$polygon_id)
check_unique_id

check_unique_polygon <- nrow(watersurfaces_hab_polygons) == n_distinct(watersurfaces_hab_polygons$geom)
check_unique_polygon
```


## Write results into a geopackage

```{r}

if(!dir.exists(file.path(path, "20_processed/watersurfaces_hab"))){
    
    dir.create(file.path(path, "20_processed/watersurfaces_hab"), recursive = TRUE)
}

filepath <- file.path(path,"20_processed/watersurfaces_hab/watersurfaces_hab.gpkg")

```


```{r}

st_write(watersurfaces_hab_polygons, 
         filepath, 
         layer = "watersurfaces_hab_polygons", 
         driver = "GPKG", 
         delete_dsn = TRUE)

st_write(watersurfaces_hab_types, 
         filepath, 
         layer = "watersurfaces_hab_types", 
         driver = "GPKG", 
         append = TRUE)

```
