# Check results

## The new version v7

Based on habitatmap_stdized_2025_v1 and watersurfaces_2024

Checksums:

```{r}
c("xxh64", "md5", "sha256") %>%
  map_dfr(~ tibble(
    algorithm = .,
    checksum = checksum(filepath, .)
  )) %>%
  kable()
```

Read file:

```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath, layer = "watersurfaces_hab_polygons"))
```

```{r paged.print=FALSE, warning=FALSE}
(types <- read_sf(filepath, layer = "watersurfaces_hab_types"))
```


- Do all records in the watersurfaces_hab_types table have a corresponding spatial feature in the watersurfaces_hab_polygons layer?

```{r}
check_polygon_id_types <- types %>%
  anti_join(pol, by = "polygon_id")
nrow(check_polygon_id_types) == 0
```

- Do all spatial features in the watersurfaces_hab_polygons layer have a corresponding record in the watersurfaces_hab_types table?

```{r}
check_polygon_id_polygons <- pol %>%
  st_drop_geometry() %>%
  anti_join(types, by = "polygon_id")
nrow(check_polygon_id_polygons) == 0
```

- Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object? 
  
```{r} 
types %>%
  distinct(polygon_id) %>%
  nrow() == nrow(pol)
``` 

- Is the CRS conform EPSG:31370?

```{r} 
st_crs(pol) == st_crs(31370)
```

- Are the geometries valid?

```{r}
validities <- st_is_valid(pol)
sum(!validities | is.na(validities)) == 0
```

- Are concatenated values contained in the polygons element composed of unique strings? This refers to <https://github.com/inbo/n2khab-preprocessing/issues/75>.

```{r}
pol %>% 
  st_drop_geometry() %>% 
  filter(if_any(-c(polygon_id, polygon_id_ws), \(x) str_detect(x, "\\+"))) %>% 
  pivot_longer(
    cols = -c(polygon_id, polygon_id_ws),
    names_to = "variable",
    values_to = "string"
  ) %>% 
  filter(!is.na(string)) %>% 
  mutate(
    unique_combs = map_lgl(string, function(x) {
      pieces <- str_match(x, "(.+)(?:\\s?\\+\\s?)?(.+)")[, -1]
      n_distinct(pieces) == length(pieces)
    })
  ) %>% 
  .$unique_combs %>% 
  all()
```

- Are the rows of the watersurfaces_hab_types table unique by polygon_id and type column?

```{r}
types %>% 
  distinct(polygon_id, type) %>% 
  nrow() == nrow(types)
```

## Compare with the previous version 

We load the previous version of watersurfaces_hab (version 6 based on watersurfaces_2024 and habitatmap_stdized_2023_v1)

```{r paged.print=FALSE, warning=FALSE}
filepath_previous <- file.path(path, "20_processed/_versions/watersurfaces_hab/watersurfaces_hab_v6/watersurfaces_hab.gpkg")

hashes <-
  tibble(filepath_previous) %>%
  mutate(
    filename = basename(filepath_previous),
    md5 = map(filepath_previous, function(x) {
      x %>%
        md5sum() %>%
        str_c(collapse = "")
    }) %>% as.character()
  ) %>%
  mutate(
    md5_ref = c("4f589b43c6d3d38e06ee3ea81e3d2058"),
    match = md5 == md5_ref
  ) %>%
  select(
    filename,
    md5,
    md5_ref,
    match
  )

if (!identical(hashes$md5, hashes$md5_ref)) {
  stop("The source map is NOT the previous version ! Please check the datasource. ")
}

(pol_previous <- read_sf(filepath_previous, layer = "watersurfaces_hab_polygons"))
(types_previous <- read_sf(filepath_previous, layer = "watersurfaces_hab_types"))
```


Apparently the previous version uses MULTIPOLYGONs while the current uses POLYGONs:

```{r}
st_geometry_type(pol$geom) %>% unique() %>% as.character()
st_geometry_type(pol_previous$geom) %>% unique() %>% as.character()
```

So we'd better convert them all to the same type before comparing:

```{r}
pol <- pol %>% st_cast("MULTIPOLYGON")
```

Are there differences between previous and current version and where are they located?

In the table below we check the differences between both versions.

```{r}
# polygons with polygon_id in current but not in previous version
check_polygon_id_current_previous <- pol %>%
  anti_join(
    pol_previous %>%
      st_drop_geometry(),
    by = c("polygon_id_habitatmap", "polygon_id_ws")
  ) %>%
  left_join(
    pol_previous %>%
      mutate(geom_text_previous = st_as_text(geom)) %>%
      st_drop_geometry() %>%
      select(polygon_id, description_orig_previous = description_orig, geom_text_previous),
    by = c("polygon_id")
  ) %>%
  mutate(
    new_polygon_id = !(polygon_id %in% pol_previous$polygon_id),
    new_polygon_id_ws = !(polygon_id_ws %in% pol_previous$polygon_id_ws),
    new_polygon_id_habitatmap = !(polygon_id_habitatmap %in% pol_previous$polygon_id_habitatmap),
    description_orig_update = description_orig != description_orig_previous,
    geom_text_current = st_as_text(geom),
    geom_update = geom_text_current != geom_text_previous
  )

check_polygon_id_current_previous %>%
  st_drop_geometry() %>%
  count(
    new_polygon_id,
    new_polygon_id_ws,
    new_polygon_id_habitatmap,
    geom_update,
    description_orig_update,
    name = "n_records"
  ) %>% 
  kable() %>%
  kable_styling()
```

The geometries haven't changed but there are differences in the descriptions 
(which is perfectly plausible with a new version of the habitatmap).

Let us take a look at the modified descriptions:

```{r}
check_polygon_id_current_previous %>%
  st_drop_geometry() %>% 
  filter(description_orig_update) %>% 
  select(
    polygon_id,
    polygon_id_ws,
    description_orig_previous,
    description_orig_current = description_orig
  ) %>% 
  kable() %>%
  kable_styling() %>% 
  scroll_box(height = "600px")
```

There is a major effect of fixing duplicates and of the fact that the separator '+' is now surrounded by spaces.

Number of `polygon_id` values unique to the previous version:

```{r}
check_polygon_id_previous_current <- pol_previous %>%
  anti_join(
    pol %>%
      st_drop_geometry(),
    join_by(polygon_id)
  ) %>%
  mutate(ws_removed = !(polygon_id_ws %in% pol$polygon_id_ws))
nrow(check_polygon_id_previous_current)
```

Subset of the above where the corresponding `polygon_id_ws` is not at all represented in the current version.
This is the number of actually dropped watersurfaces:

```{r}
sum(check_polygon_id_previous_current$ws_removed)
```

Below we show:

+ new polygons from the watersurfaces layer that are included in the current version of `watersurfaces_hab` (blue polygons)
+ polygons from the watersurfaces layer that are removed in the current version of `watersurfaces_hab` compared to the previous version (black polygons)


```{r}
ws_new <- check_polygon_id_current_previous %>%
  filter(new_polygon_id_ws) %>%
  st_transform(crs = 4326)

ws_removed <- check_polygon_id_previous_current %>%
  filter(ws_removed) %>%
  st_transform(crs = 4326)

leaflet() %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(
    data = ws_new,
    group = "in current and not previous (blue)",
    popup = paste(
      "polygon_id_habitatmap:", ws_new$polygon_id_habitatmap, "<br>",
      "polygon_id_ws:", ws_new$polygon_id_ws
    )
  ) %>%
  addPolygons(
    data = ws_removed,
    color = "black", group = "in previous and not current (black)",
    popup = paste(
      "polygon_id_habitatmap:", ws_removed$polygon_id_habitatmap, "<br>",
      "polygon_id_ws:", ws_removed$polygon_id_ws
    )
  ) %>%
  addLayersControl(
    overlayGroups = c("in current and not previous (blue)", "in previous and not current (black)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

