# Check results

## The new version v4

Based on habitatmap_stdized_2020_v1 and watersurfaces_v1.1

Checksums:

```{r}
filepath %>% 
  n2khab::md5sum() %>% 
  `names<-`("md5sum")
filepath %>% 
  n2khab::sha256sum() %>% 
  `names<-`("sha256sum")
filepath %>% 
  n2khab::xxh64sum() %>% 
  `names<-`("xxh64sum")
```

Read file:

```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath, 
                layer = "watersurfaces_hab_polygons"))
```

```{r paged.print=FALSE, warning=FALSE}
(types <- read_sf(filepath, 
                layer = "watersurfaces_hab_types"))
```


- Do all records in the watersurfaces_hab_types table have a corresponding spatial feature in the watersurfaces_hab_polygons layer?

```{r}
check_polygon_id_types <- types %>%
  anti_join(pol, by = "polygon_id")
nrow(check_polygon_id_types) == 0
```

- Do all spatial features in the watersurfaces_hab_polygons layer have a corresponding record in the watersurfaces_hab_types table?

```{r}
check_polygon_id_polygons <- pol %>%
  st_drop_geometry() %>%
  anti_join(types, by = "polygon_id")
nrow(check_polygon_id_polygons) == 0
```

- Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object? 
  
```{r} 
 types %>%  
   distinct(polygon_id) %>%  
   nrow == nrow(pol) 
``` 

- Is the CRS conform EPSG:31370?

```{r} 
 st_crs(pol) == st_crs(31370) 
```

## Compare with the previous version 

We load the previous version of watersurfaces_hab (version 3 based on watersurfaces_v1 and habitatmap_stdized_2018_v2)

```{r paged.print=FALSE, warning=FALSE}
# adapt this part of the script to load version v3 (expected md5 = fcf021e4bfca827dfc13466295d65174
# as referenced in https://docs.google.com/spreadsheets/d/1E8ERlfYwP3OjluL8d7_4rR1W34ka4LRCE35JTxf3WMI/edit#gid=2100595853)
filepath_v3 <- file.path(path,"20_processed/watersurfaces_hab/versies/v3/watersurfaces_hab.gpkg")

hashes <-
    tibble(filepath_v3) %>%
    mutate(filename = basename(filepath_v3),
           md5 = map(filepath_v3, function(x) {
                           x %>% md5sum() %>% str_c(collapse = '')
                         }) %>% as.character) %>% 
           mutate(md5_ref = c("fcf021e4bfca827dfc13466295d65174"),
           match = md5 == md5_ref) %>%
    select(filename,
           md5,
           md5_ref,
           match)

if (!all.equal(hashes$md5, hashes$md5_ref)) {
    stop(cat("The source map is NOT v3 ! Please check the datasource. "))
}
  
(pol_v3 <- read_sf(filepath_v3, 
                layer = "watersurfaces_hab_polygons"))
(types_v3 <- read_sf(filepath_v3, 
                layer = "watersurfaces_hab_types"))
```

- Are there differences between version v3 and version v4 and where are they located?

There are many changes between versions, and unfortunately it is difficult
to check them without showing the source maps for both versions.
A detailed check is not easy to realize, so we merely show a map of the polygons 
with a difference in polygon_id for watersurface and/or habitatmap_stdized.

```{r}
# polygons with polygon_id in v4 butnot in version v3
check_polygon_id_v4_v3 <- pol %>%
  anti_join(pol_v3 %>%
              st_drop_geometry(), 
            by = c("polygon_id_habitatmap", "polygon_id_ws"))
 nrow(check_polygon_id_v4_v3)
   
# we expected new polygons with polygon_id like "xxx_v2020" so we neglect 
# them when we count
 nrow(check_polygon_id_v4_v3 %>% 
   filter(!str_detect(string = polygon_id_habitatmap, pattern = "v2020")))

```

```{r}
check_polygon_id_v3_v4 <- pol_v3 %>%
  anti_join(pol %>%
              st_drop_geometry(), 
            by = c("polygon_id_habitatmap", "polygon_id_ws"))

nrow(check_polygon_id_v3_v4) 
```

```{r, out.width='800px', out.height='500px'}
leaflet() %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = check_polygon_id_v4_v3 %>%
            st_transform(crs = 4326), 
            group = "in v4 and not v3 (blue)",
            popup = paste("polygon_id_habitatmap:", check_polygon_id_v4_v3$polygon_id_habitatmap, "<br>",
                           "polygon_id_ws:", check_polygon_id_v4_v3$polygon_id_ws)) %>%
  addPolygons(data = check_polygon_id_v3_v4 %>%
            st_transform(crs = 4326), 
            color = "black", group = "in v3 and not v4 (black)",
            popup = paste("polygon_id_habitatmap:", check_polygon_id_v3_v4$polygon_id_habitatmap, "<br>",
                           "polygon_id_ws:", check_polygon_id_v3_v4$polygon_id_ws)) %>% 
  addLayersControl(
    overlayGroups = c("in v4 and not v3 (blue)", "in v3 and not v4 (black)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

