# Check results

## The new version v6

Based on habitatmap_stdized_2023_v1 and watersurfaces_2024

Checksums:

```{r}
c("xxh64", "md5", "sha256") %>% 
  map_dfr(~tibble(algorithm = ., 
                  checksum = checksum(filepath, .))) %>% 
  kable
```

Read file:

```{r paged.print=FALSE, warning=FALSE}
(pol <- read_sf(filepath, 
                layer = "watersurfaces_hab_polygons"))
```

```{r paged.print=FALSE, warning=FALSE}
(types <- read_sf(filepath, 
                layer = "watersurfaces_hab_types"))
```


- Do all records in the watersurfaces_hab_types table have a corresponding spatial feature in the watersurfaces_hab_polygons layer?

```{r}
check_polygon_id_types <- types %>%
  anti_join(pol, by = "polygon_id")
nrow(check_polygon_id_types) == 0
```

- Do all spatial features in the watersurfaces_hab_polygons layer have a corresponding record in the watersurfaces_hab_types table?

```{r}
check_polygon_id_polygons <- pol %>%
  st_drop_geometry() %>%
  anti_join(types, by = "polygon_id")
nrow(check_polygon_id_polygons) == 0
```

- Is the number of unique IDs in the dataframe the same as the number of polygons in the sf object? 
  
```{r} 
 types %>%  
   distinct(polygon_id) %>%  
   nrow == nrow(pol) 
``` 

- Is the CRS conform EPSG:31370?

```{r} 
 st_crs(pol) == st_crs(31370) 
```

- Are the geometries valid?

```{r}
validities <- st_is_valid(pol)
sum(!validities | is.na(validities)) == 0
```

## Compare with the previous version 

We load the previous version of watersurfaces_hab (version 5 based on watersurfaces_v1.2 and habitatmap_stdized_2023_v1)

```{r paged.print=FALSE, warning=FALSE}

filepath_v5 <- file.path(path,"20_processed/_versions/watersurfaces_hab/watersurfaces_hab_v5/watersurfaces_hab.gpkg")

hashes <-
    tibble(filepath_v5) %>%
    mutate(filename = basename(filepath_v5),
           md5 = map(filepath_v5, function(x) {
                           x %>% md5sum() %>% str_c(collapse = '')
                         }) %>% as.character) %>% 
           mutate(md5_ref = c("e7d9930938f5111de33de6ecaec31a66"),
           match = md5 == md5_ref) %>%
    select(filename,
           md5,
           md5_ref,
           match)

if (!all.equal(hashes$md5, hashes$md5_ref)) {
    stop(cat("The source map is NOT v5 ! Please check the datasource. "))
}
  
(pol_v5 <- read_sf(filepath_v5, 
                layer = "watersurfaces_hab_polygons"))
(types_v5 <- read_sf(filepath_v5, 
                layer = "watersurfaces_hab_types"))
```

- Are there differences between version v5 and version v6 and where are they located?

In the table below we check the differences between both versions.
Note that for a large number of records only polygon_habitatmap_id changes, but the geometry and the type description remain the same.

```{r}
# polygons with polygon_id in v6 but not in version v5
check_polygon_id_v6_v5 <- pol %>%
  anti_join(pol_v5 %>%
              st_drop_geometry(), 
            by = c("polygon_id_habitatmap", "polygon_id_ws")) %>%
  left_join(pol_v5 %>%
              mutate(geom_text_v5 = st_as_text(geom)) %>%
              st_drop_geometry() %>%
              select(polygon_id, description_orig_v5 = description_orig, geom_text_v5), 
            by = c("polygon_id")) %>%
  mutate(new_polygon_id = !(polygon_id %in% pol_v5$polygon_id),
         new_polygon_id_ws = !(polygon_id_ws %in% pol_v5$polygon_id_ws),
         new_polygon_id_habitatmap = !(polygon_id_habitatmap %in% pol_v5$polygon_id_habitatmap),
         description_orig_update = description_orig != description_orig_v5,
         geom_text_v6 = st_as_text(geom),  
         geom_update = geom_text_v6 != geom_text_v5)

check_polygon_id_v6_v5 %>%
  st_drop_geometry() %>%
  group_by(new_polygon_id, new_polygon_id_ws, new_polygon_id_habitatmap, geom_update, description_orig_update) %>%
  summarise(n_records = n()) %>%
  ungroup() %>%
  kable() %>%
  kable_styling()

```

We check some of the polygons for which the geometry has changed. 

In this case there are only 2 polygons with modified geometry
Changes are minor for Stappersven, and bigger extent for Houtsaegerduinen.

```{r}
check_geom <- check_polygon_id_v6_v5 %>%
  filter(geom_update & !is.na(geom_update)) %>%
  slice_head(n = 5) %>%
  st_transform(4326)

check_geom_v5 <- pol_v5 %>%
  filter(polygon_id %in% check_geom$polygon_id) %>%
  st_transform(4326)

check_geom %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(group = "v6") %>%
  addPolygons(data = check_geom_v5, color = "red", group = "v5") %>%
   addLayersControl(
   overlayGroups = c("v6 (blue)", "v5 (red)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```




```{r}
check_polygon_id_v5_v6 <- pol_v5 %>%
  anti_join(pol %>%
              st_drop_geometry(), 
            by = c("polygon_id_habitatmap", "polygon_id_ws")) %>%
  mutate(ws_removed = !(polygon_id_ws %in% pol$polygon_id_ws))

nrow(check_polygon_id_v5_v6) 
```

Here we show:

+ new polygons from the watersurfaces layer that are included in `watersurfaces_hab_v6` (blue polygons)
+ polygons from the watersurfaces layer that are removed in `watersurfaces_hab_v6` compared to `watersurfaces_hab_v5` (black polygons)


```{r}

ws_new <- check_polygon_id_v6_v5 %>%
  filter(new_polygon_id_ws) %>%
  st_transform(crs = 4326)

ws_removed <- check_polygon_id_v5_v6 %>%
  filter(ws_removed) %>%
  st_transform(crs = 4326)

leaflet() %>%
  addTiles(group = "OSM (default)") %>%
  addPolygons(data = ws_new, 
            group = "in v6 and not v5 (blue)",
            popup = paste("polygon_id_habitatmap:", ws_new$polygon_id_habitatmap, "<br>",
                           "polygon_id_ws:", ws_new$polygon_id_ws)) %>%
  addPolygons(data = ws_removed, 
            color = "black", group = "in v5 and not v6 (black)",
            popup = paste("polygon_id_habitatmap:", ws_removed$polygon_id_habitatmap, "<br>",
                           "polygon_id_ws:", ws_removed$polygon_id_ws)) %>% 
  addLayersControl(
    overlayGroups = c("in v6 and not v5 (blue)", "in v5 and not v6 (black)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

